### 2017-02-09 TeX/paper.tex
Add files via upload

\documentclass[headsepline=true, abstracton]{scrartcl}
%\usepackage[ngerman]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{verbatim}
\usepackage[a4paper,text={140mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
\usepackage{aliascnt}
\usepackage{lmodern}
\usepackage{mdwlist}
\usepackage{listings}
\usepackage[activate]{pdfcprot}
\usepackage{graphicx}
\usepackage{slashed}
\usepackage{mathrsfs}
\usepackage{geometry}
\usepackage{float}
\usepackage{oldgerm}
\usepackage{setspace}
\usepackage{dsfont}
%\usepackage{mathtools}
\usepackage[all]{xy}
\usepackage{cite}
\usepackage{url}
\pagestyle{headings}
\newcommand{\myclearpage}{\clearpage}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newenvironment{gelaber}{}{}
\newenvironment{preamble}{}{}
\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
\newtheorem{mydef}{Definition}
\newtheorem{bem}{Bemerkung}
\usepackage{tikz}
\newcommand\circlearound[1]{%
  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}

\begin{document}

\setcounter{section}{-1}

\renewcommand{\refname}{Bibliography}

\onehalfspacing
\setlength{\headsep}{15mm}

\thispagestyle{plain}

 \section*{Introduction}

 Coming...

 \section*{The Exponential Random Graph Model}
The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks. This means neither the weighting nor the temporal change of ties is considered in the model.
In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}). In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm] %This chapter is mainly based on Harris \cite{Harris.}, Hunter et al. \cite{Hunter.2008} and Hunter \cite{Hunter.}. \\[0.3cm]
The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. Therefore, for an already small number of actors the cardinalty of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable. As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\  
With the definition of $\mathcal{G}(N)$ we define 
$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
\begin{equation}
\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$. The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).

\section*{Estimation}
As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology. Therefore, the question arises how one can estimate the parameter vector $\theta$?
A first idea could be the following: One can assume that the \textit{dyads} are independent of each other, which means that the random variables $Y_{ij}$ inside the random matrix $Y$ are independent of each other. 
In this case, one can show that
%
\begin{equation*}
\text{logit}(\mathbb{P}_{\theta}(Y_{ij}=1))= \theta^T \cdot (\Delta G)_{ij}
%\label{ILogit}
\end{equation*}
This corresponds with the \textit{logistic regression} approach, where the observations of the dependent variables are simply edge values of the observed adjacency matrix, and the observations of the covariate values are given as the scores of every single change statistic. Therefore, the estimation of $\theta$ can then be obtained as usual using straight forward maximum-likelihood estimation. The resulting likelihood function is of the following form:
\begin{equation}
\text{lik}(\theta)= \mathbb{P}_{\theta}(Y=G)= \prod_{i,j} \dfrac{ \exp \left(\theta^T \Delta(G))_{ij} \right)}{1+\exp \left(\theta^T \Delta(G))_{ij} \right)}
\label{NaiveLik}
\end{equation}
The problem with this simple estimation procedure is that the assumed hypothesis of the independence of the dyads turns out to be erroneous in most cases (see van Duijn et all ?citevanDuijn. This is a systematic problem: The presence of network data is inextricably connected with the presence of \textit{relational data}, which by definition should not be assumed to be independent of each other. If this dependency structure is deliberately ignored and equation (\ref{NaiveLik}) is used to estimate $\theta$, it results in a \textit{maximum pseudo-likelihood estimation} (MPLE). This technique tends to underestimate the standard error. However, Desmarais and Cranmer \cite{Desmarais.2012} show that the pseudo-likelihood provides a consistent approximation of the maximum likelihood, meaning that the MPLE converges to the MLE as the size of the network increases.\\[0.3cm]
There are several techniques to circumvent estimators, which underestimate the standard error of $\theta$. In the following, we will introduce a technique based on \textit{Markov Chain Monte Carlo (MCMC)} and maximum-likelihood methods.\\[0.3cm]
The more rigorous technique is to estimate the parameters directly with the log-likelihood function derived from (\ref{ERGM}), which has the following form:
%
\begin{equation}
\text{loglik}(\theta)=\theta^T \cdot \Gamma(G)-\log(c(\theta))
\label{loglik}
\end{equation}
%
where $G$ is the observed network. For the vector of network statistics, one can assume without loss of generality
%
\begin{equation}
\Gamma(G)=0 \label{assumption}
\end{equation}
%
This means that centering the vector of network statistics does not affect the probability function of the network variable $Y$. Therefore, in context of the likelihood function (\ref{loglik}) the vector of statistics can always be assumed to be centered around the observed network.\\
Due to assumption (\ref{assumption}), one gets from (\ref{loglik}) the simplified log-likelihood function
%
\begin{equation}
\text{loglik}(\theta)= -\log(c(\theta))
\label{vloglik}
\end{equation}
%
The problem resulting from estimating the parameters with (\ref{loglik}) is that the term
%
$$c(\theta):= \sum_{G^* \in \mathcal{G}(N_V)} \exp(\theta^T \cdot \Gamma(G^*))$$ 
%
which sums up the weighted network statistics of all possible networks of $N$ nodes, has to be evaluated. Even for networks with small numbers of nodes this presents an enormous computational obstacle, and the necessary calculations for larger networks can not currently be completed in any reasonable timeframe.
As a result, for sufficiently large networks it is not possible to estimate the parameters directly with the likelihood function.
\\[0.3cm]
An expedient for this limitation is based on the following consideration: Fix any vector of parameters $\theta_0 \in \Theta$ from the underlying parameter range $\Theta$ and compute for $\theta \in \Theta$ the expected value. Then, one can show that
%
\begin{equation*}
\mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]=\frac{c(\theta)}{c(\theta_0)}
\end{equation*}
%
This equation offers the following possibility: If one draws $L$ random networks $G_1, \dots ,G_L$ out of a distribution $\mathbb{P}_{\theta_0}$ appropriately, one gets with the \textit{law of big numbers} the following relation:
%
\begin{equation}
\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] = \frac{c(\theta)}{c(\theta_0)}
\label{konver}
\end{equation}
%
For a big enough number, $L$, of random networks, the following approximation is reasonable:
%
\begin{equation}
\frac{c(\theta)}{c(\theta_0)} \approx \frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
\label{approx}
\end{equation}
%
One can now use equation (\ref{approx}) to determine an approximation of the log-likelihood function (\ref{vloglik}):
%
\begin{eqnarray*}
\text{loglik}(\theta)-\text{loglik}(\theta_0)&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\end{eqnarray*}
%
By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
%
\begin{equation}
s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\label{score}
\end{equation}
%
This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm}. As a result, the approximate maximum-likelihood estimator for the parameters can be computed.\\[0.4cm]
As pleasant as this may sound, the immediate question arises: How can a sufficient number of suitable drawings $G_1,...,G_L$ be taken from the distribution $\mathbb{P}_{\theta_0}$? \\
For this purpose, the \textit{Markov Chain Monte Carlo (MCMC)} methods can be used. This approach does not deliberately ignore the dependency structures inside the network. Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.
When it comes to the interpretation of the coefficients one can show that
\begin{equation}
\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 | Y_{ij}^c=G_{ij}^c)}
= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
\label{Inter2}
\end{equation}
where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}. $G_{ij}^+$ emerges from $G$, while assuming $g_{ij}=1$ and $G_{ij}^-$ emerges from $G$, while assuming $g_{ij}=0$. This means that the $k$th component of $(\Delta G)_{ij}$ captures the difference between the networks $G_{ij}^+$ and $G_{ij}^-$ on the $k$th integrated statistic in the model.\\[0.3cm]
Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$. This means that for $\theta_k$, $k \in \{1, \ldots , q\}$ the conditional odds of occurence increase if $\theta_k > 0 $, decrease if $\theta_k < 0 $ and stay the same if $\theta_k = 0 $. 
Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.

\section*{Efficiency of MPLE/MCMCMLE}
Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression analysis and does not exert elaborate MCMC methods. Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}). This means that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE should be considered as an option to computationally intensive MCMC methods.\\
On the other hand a major reason for MCMCMLE methods is, as shown by van Duijn et al. \cite{vanDuijn.2009}, that the MCMCMLE is more efficient than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE. However, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. Furthermore, the number of simulated networks needed in order for the MCMCMLE to surpass the MPLE increases as the size of the network increases, making it difficult to evaluate whether the chosen number of simulated networks is large enough. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the MCMCMLE approach.

\textcolor{blue}{We need at least two different networks of different size in order to show the points above.}\\
In order to proof these disadvantages of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students. The original data comes from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
\begin{itemize}
\item explain simulation study
\end{itemize}
Based on these results we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 
$$RMSE = \sqrt{\sum_{i=1}^{n}(\theta - \hat{\theta})^2}$$
meaning that the smaller the RMSE, the more accurate the estimator. Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE, we take the log of the ratio of the MCMCMLE and the MPLE. As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance.

\begin{itemize}
\item RMSE plot + explanation + discussion
\end{itemize}

\section*{Bootstrapped MPLE}
\begin{itemize}

\item Describe method 
\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
\item still consistent
\item better efficiency than MPLE and similar results as MCMCMLE
\item explain coverage probability and inlcude coverage plots
\item can easily be parallelized/ using multiple cores -> increases speed
\item this method is also able to detect degeneracy

\end{itemize}

 \section*{Results/Discussion}
\begin{itemize}

 \item  Bootstrap MPLE is an improvement of the MPLE and performs just as good as the MCMCMLE. \\
\item Bootstrap MPLE has the advantage of being fast + multiple cores\\ 
\item Bootstrap MPLE is consistent\\
\item Conclusion: For very large networks the Bootstrap MPLE approach is more reasonable

\end{itemize}

 \section*{To determine}

 \begin{enumerate}
 \item where to include Boltzmann machine application
\end{enumerate}

\newpage

\bibliography{bib} 
\bibliographystyle{plain}

\end{document}

### 2017-02-16 TeX/paper.tex
added an abstract
added the abstract

@@
 \thispagestyle{plain}
+
+\begin{abstract}
+With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data.
+ERGM parameter estimation requires the approximation of an intractable normalizing constant.
+Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE).
+MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant.
+However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000.
+When the network is large, one option for estimation is maximum pseudolikelihood (MPLE).
+MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks.
+The standard MPLE is simple and fast, but generally underestimates standard errors.
+We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals.
+Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel.
+We compare the two different approaches by applying them on U.S.
+Supreme Court Citation Network Data.
+\end{abstract}
 

### 2017-02-20 small_network_fauxmesahigh/RMSE_Mesa_Magnolia.R
Add files via upload

@@ -1,0 +1,702 @@
+# calculating RMSE for faux mesa high and faux magnolia high
+
+# first Faux mesa high
+
+library(statnet)
+
+data("faux.mesa.high")
+
+plot(faux.mesa.high)
+
+set.seed(123)
+model=ergm(faux.mesa.high~edges+nodefactor("Sex")+gwesp(0,fixed=TRUE))
+summary(model)
+#get true coefficients
+coefficient <- summary(model)$coef[,1]
+
+# function to calculate RMSE for MPLE, MCMCMLE with samplesize= 1500, 800, 400, 200, 100, 50, 20
+main.function.faux.mesa.RMSE <- function(iters,  coefs=coefficient){
+  # size = number of nodes in network
+  
+  set.seed(999)
+  
+  l=length(coefs) # number of coefficients
+  
+  #creating empty matrices to store results
+  coef.matrix.mple <- matrix(0,l,iters)
+  se.matrix.mple <- matrix(0,l,iters)
+  
+  coef.matrix.1500 <- matrix(0,l,iters)
+  se.matrix.1500 <- matrix(0,l,iters)
+  
+  coef.matrix.800 <- matrix(0,l,iters)
+  se.matrix.800 <- matrix(0,l,iters)
+  
+  coef.matrix.400 <- matrix(0,l,iters)
+  se.matrix.400 <- matrix(0,l,iters)
+  
+  coef.matrix.200 <- matrix(0,l,iters)
+  se.matrix.200 <- matrix(0,l,iters)
+  
+  coef.matrix.100 <- matrix(0,l,iters)
+  se.matrix.100 <- matrix(0,l,iters)
+  
+  coef.matrix.50 <- matrix(0,l,iters)
+  se.matrix.50 <- matrix(0,l,iters)
+  
+  coef.matrix.25 <- matrix(0,l,iters)
+  se.matrix.25 <- matrix(0,l,iters)
+  
+  
+  # start loop
+  for(i in 1:iters){
+    print(i)
+    
+    # simulate new network 
+    g.sim <- simulate(faux.mesa.high ~ edges+nodefactor("Sex")+gwesp(0,fixed=TRUE), nsim=1, coef=coefs #true coefficients
+                      , control=control.simulate(MCMC.burnin=1000000,MCMC.interval=1000))
+    
+    # MPLE
[…]

### 2017-02-20 TeX/paper.tex
Update paper.tex

@@
 
+\title{\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap}
+\maketitle
+
 \begin{abstract}
-With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data.
+\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data.
 ERGM parameter estimation requires the approximation of an intractable normalizing constant.
@@
 Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy.
-Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression analysis and does not exert elaborate MCMC methods.
+Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
 Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}).
-This means that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE should be considered as an option to computationally intensive MCMC methods.\\
-On the other hand a major reason for MCMCMLE methods is, as shown by van Duijn et al.
-\cite{vanDuijn.2009}, that the MCMCMLE is more efficient than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE.
-However, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
-Furthermore, the number of simulated networks needed in order for the MCMCMLE to surpass the MPLE increases as the size of the network increases, making it difficult to evaluate whether the chosen number of simulated networks is large enough.
-In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the MCMCMLE approach.
+This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
+However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al.
+\cite{vanDuijn.2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE.
+Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
+It is even more remarkable that the number of simulated networks needed for the MCMCMLE, in order to surpass the MPLE increases as the size of the network increases.
+This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE.
+In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\
 
-\textcolor{blue}{We need at least two different networks of different size in order to show the points above.}\\
-In order to proof these disadvantages of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students.
-The original data comes from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
-\begin{itemize}
-\item explain simulation study
-\end{itemize}
-Based on these results we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance.
+\noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
+The data for both networks originates from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
+For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$.
+For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution.
+We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMCMLE of these simulated networks.
+For every single simulated network the MCMCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
+Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance.
 Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as
-$$RMSE = \sqrt{\sum_{i=1}^{n}(\theta - \hat{\theta})^2}$$
-meaning that the smaller the RMSE, the more accurate the estimator.
+$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$
+implying that the smaller the RMSE, the more accurate is the estimator.
 Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases.
 On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required.
-In order to compare the RMSE, we take the log of the ratio of the MCMCMLE and the MPLE.
+In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMCMLE to the MPLE.
 As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance.
+Figure \ref{rmse} visualizes the results of the simulation study.
+The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.
 
-\begin{itemize}
-\item RMSE plot + explanation + discussion
-\end{itemize}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=.5]{RMSE_paper}
+\caption{The log of the ratio of the RMSE for the MCMCMLE to the MPLE for different samplesizes and two different networks, Faux Mesa High and Faux Magnolia High}
+\label{rmse}
+\end{center}
+\end{figure}
+
+\noindent The plots support the fact that larger networks require a larger samplesize of simulated networks in order for the MCMCMLE to outperform the MPLE.
+While the fairly small Faux Mesa High network only requires a samplesize of about 50 networks the larger Faux Magnolia High network already requires a samplesize of at least 1500 networks for the MCMCMLE to surpass the MPLE.
+These results propose especially for very large networks (e.g.
+social media data) the question, of how large the samplesize has to be set in order to justify the approximately exact, but comutationally expensive MCMCMLE method.
+The identification of the required samplesize is not only becoming more difficult as the size of the network increases, but an increase of the samplesize will also extends the mere computation time.
 

### 2017-02-25 TeX/paper.tex
chapter bootstrapped MPLE

@@
 
- Coming...
+ Coming soon...
 
@@
 \includegraphics[scale=.5]{RMSE_paper}
-\caption{The log of the ratio of the RMSE for the MCMCMLE to the MPLE for different samplesizes and two different networks, Faux Mesa High and Faux Magnolia High}
+\caption{The log of the ratio of the RMSE for the MCMCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
 \label{rmse}
@@
 
-\noindent The plots support the fact that larger networks require a larger samplesize of simulated networks in order for the MCMCMLE to outperform the MPLE.
-While the fairly small Faux Mesa High network only requires a samplesize of about 50 networks the larger Faux Magnolia High network already requires a samplesize of at least 1500 networks for the MCMCMLE to surpass the MPLE.
+\noindent The plots support the fact that larger networks require a larger sample size of simulated networks in order for the MCMCMLE to outperform the MPLE.
+While the fairly small Faux Mesa High network only requires a sample size of about 50 networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMCMLE to surpass the MPLE.
 These results propose especially for very large networks (e.g.
-social media data) the question, of how large the samplesize has to be set in order to justify the approximately exact, but comutationally expensive MCMCMLE method.
-The identification of the required samplesize is not only becoming more difficult as the size of the network increases, but an increase of the samplesize will also extends the mere computation time.
+social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMCMLE method.
+The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extends the mere computation time.
 
 \section*{Bootstrapped MPLE}
+As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
+Moreover, it is able to outperform the MCMCMLE if sample size is not chosen large enough.
+The main reason, why the MCMCLE is still widely preferred is, that in contrast to the MPLE, it does not underestimate the variance of its estimates.
+By the definition of the ERGM it is obvious that this model is an exponential family of distributions where $\theta$ is the natural parameter and $\Gamma (Y)$ takes the part of the sufficient statistic.
+This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
+The problem with the MPLE is that the inverse of the negative hessian matrix calculated by the pseudolikelihood function will underestimate the variance of the MPLE \cite{Duijnetall2009}, resulting in an underestimate of the width of the confidence intervals.
+Duijn et all (cite) show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%.
+In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use $[-H]^{-1}$ as an estimate for the covariance matrix.
+\\
+Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
+Desmarais and Cranmer \cite{Desmarais.2012} also show that this bootstrapped MPLE is a multivariate \textit{M}-estimator, a special class of robust estimators, proving that this algorithm consistently estimates the confidence intervals of the MPLE.
+In order to calculate the bootstrapped MPLE, a sample of $n$ networks are being drawn with replacement from the distribution of $Y$.
+Then, we evaluate the MPLE on every sampled network and take the $2.5$th and $97.5$th percentile of the MPLE estimates to obtain 95\% bootstrap confidence intervals.
+Obviously, we obtain bootstrap confidence intervals with a different significance level $\alpha$ by simply picking the corresponding percentiles.
+
 \begin{itemize}

### 2017-03-13 large_network_scnetwork54/Coverage/combine_workspace_bootCI.R
Add files via upload

@@ -1,0 +1,38 @@
+# loading different workspaces
+
+load("supreme54.RData")
+
+load("coverage_parallel_scnetwork1.RData")
+boot1<- boot
+
+load("coverage_parallel_scnetwork2.RData")
+boot2<- boot
+
+load("coverage_parallel_scnetwork3.RData")
+boot3<- boot
+
+load("coverage_parallel_scnetwork4.RData")
+boot4<- boot
+
+rm(boot)
+
+# combine results
+boot<-cbind(boot1, boot2, boot4)
+
+# bootstrap.CI: if coef(mple) is in BootCI, save 1 otherwise 0
+# create empty matrix
+bootstrap.CI <- matrix(0,7,2)
+colnames(bootstrap.CI)<- c("upper", "lower")
+rownames(bootstrap.CI)<- c("edges", "nodecov.salience", "nodecov.year.center","absdiff.year.center", "absdiff.year.center.square", "dsp0", "esp0")
+
+# calculate the bootstrap intervals
+bootstrap.CI[1,]<-quantile(boot[1,], probs = c(0.975, 0.025)) # edges
+bootstrap.CI[2,]<-quantile(boot[2,], probs = c(0.975, 0.025)) # salience
+bootstrap.CI[3,]<-quantile(boot[3,], probs = c(0.975, 0.025)) # nodecov.year
+bootstrap.CI[4,]<-quantile(boot[4,], probs = c(0.975, 0.025)) # absdiff.year
+bootstrap.CI[5,]<-quantile(boot[5,], probs = c(0.975, 0.025)) # absdiff.year.square
+bootstrap.CI[6,]<-quantile(boot[6,], probs = c(0.975, 0.025)) # dsp0
+bootstrap.CI[7,]<-quantile(boot[7,], probs = c(0.975, 0.025)) # esp0
+
+mple
+bootstrap.CI

### 2017-03-13 TeX/paper.tex
started intro paragraph

@@
 
- Coming soon...
+The field of network science faces a double-edge sword when it comes to computational limitations to innovation.
+First, the networks under study are growing larger with each year.
+{\bf [CHRISTIAN, WOULD YOU CHECK OUT N NODES IN {\em SOCIAL NETWORKS}?]}.
+Second, analytical methods are growing more sophisticated, increasingly involving estimation and optimization, going beyond descriptive calculations {\bf [Could also classify articles based on whether a statistical model is used {\em SOCIAL NETWORKS}?]}.
+In order to avoid a limiting horizon in which state-of-the-art methods cannot be used with state-of-the-art ``big'' data, network methodologists need to consider all available options in easing the computational burden of tools for network analysis.
 

### 2017-03-14 TeX/paper.tex
coverage probability

--- 2017-02-25 TeX/paper.tex
@@
 This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE.
-In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\
-
+In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\[0.3cm]
 \noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
-The data for both networks originates from Resnick et al \cite{Resnick.1997}.\\[0.3cm]
+The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
 For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$.
@@
 As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
-Moreover, it is able to outperform the MCMCMLE if sample size is not chosen large enough.
-The main reason, why the MCMCLE is still widely preferred is, that in contrast to the MPLE, it does not underestimate the variance of its estimates.
-By the definition of the ERGM it is obvious that this model is an exponential family of distributions where $\theta$ is the natural parameter and $\Gamma (Y)$ takes the part of the sufficient statistic.
+Moreover, the MPLE is able to outperform the MCMCMLE if the sample size is not chosen large enough.
+The main reason why the MCMCLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}.
+By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic.
 This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
-The problem with the MPLE is that the inverse of the negative hessian matrix calculated by the pseudolikelihood function will underestimate the variance of the MPLE \cite{Duijnetall2009}, resulting in an underestimate of the width of the confidence intervals.
-Duijn et all (cite) show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%.
-In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use $[-H]^{-1}$ as an estimate for the covariance matrix.
-\\
-Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
-Desmarais and Cranmer \cite{Desmarais.2012} also show that this bootstrapped MPLE is a multivariate \textit{M}-estimator, a special class of robust estimators, proving that this algorithm consistently estimates the confidence intervals of the MPLE.
-In order to calculate the bootstrapped MPLE, a sample of $n$ networks are being drawn with replacement from the distribution of $Y$.
-Then, we evaluate the MPLE on every sampled network and take the $2.5$th and $97.5$th percentile of the MPLE estimates to obtain 95\% bootstrap confidence intervals.
-Obviously, we obtain bootstrap confidence intervals with a different significance level $\alpha$ by simply picking the corresponding percentiles.
+The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals.
+Duijn et al.
+show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%.
+In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix.
+\\[0.3cm]
+Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
+Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
+We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
+\\[0.3cm]
+For the simulation study, we determine the MPLE for the model and treat these estimates as the networks 'true' parameter values.
+We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$.
+For each of the $1000$ networks we calculate the confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
+In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values.
+For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the MPLE estimates to obtain 95\% bootstrap confidence intervals.
+Similar as for the MCMCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
+Figures ref1 and ref2 visualize the coverage probabilities for each of the three methods for both networks.
 
@@
 
-\item Describe method 
+\item include plot
+\item Describe plots
 \item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
-\item still consistent
-\item better efficiency than MPLE and similar results as MCMCMLE
-\item explain coverage probability and inlcude coverage plots
 \item can easily be parallelized/ using multiple cores -> increases speed
@@
 \end{itemize}
+
+\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
+Furthermore, this method also adopts the advantage of the MPLE of being consistent.
+However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit especially if the size of the examined network is extremely large.
+
+\section*{Supreme Court Citation Network Data}
+
+\cite{FowlerJeon2008}, \cite{Fowleretal2007}
 

### 2017-03-14 small_network_fauxmesahigh/Coverage_Probability_fauxmesahigh.R

@@ -200,8 +200,8 @@ bootSE.faux <- function( erg.obj, simulation, iters){
 
 # simulation MPLE
 set.seed(5555)
-sim=100
-iter=250
+sim=500
+iter=30
 faux.results.mple <- bootSE.faux(mple, sim,iter)
 
 # coef=Inf appears
@@ -234,8 +234,8 @@ counts[2,2] <- mean(faux.results.mple[[5]][2,]) # sex mcmcmle
 counts[2,3] <- mean(faux.results.mple[[5]][3,]) # gwesp mcmcmle
 
 # Plot
-barplot(counts, beside=TRUE, main="Coverage Probability 90%", ylim=c(0,1),legend = rownames(counts),args.legend = list(x="bottomleft"))
-abline(h=0.9,col="grey", lty=2)
+barplot(counts, beside=TRUE, main="Coverage Probability 95%", ylim=c(0,1),legend = rownames(counts),args.legend = list(x="bottomleft"))
+abline(h=0.95,col="grey", lty=2)
 
 # Histogram of the density of simulated networks
 hist(faux.results.mple[[6]], main="Density of Simulated Networks", xlab="Density")

### 2017-03-14 TeX/paper.tex
simulation of networks

--- 2017-03-14 TeX/paper.tex
@@
 Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.
-When it comes to the interpretation of the coefficients one can show that
-\begin{equation}
-\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 | Y_{ij}^c=G_{ij}^c)}
-= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
-\label{Inter2}
-\end{equation}
-where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
-The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}.
+%When it comes to the interpretation of the coefficients one can show that
+%\begin{equation}
+%\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 %| Y_{ij}^c=G_{ij}^c)}
+%= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
+%\label{Inter2}
+%\end{equation}
+%where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
+%The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}.
 $G_{ij}^+$ emerges from $G$, while assuming $g_{ij}=1$ and $G_{ij}^-$ emerges from $G$, while assuming $g_{ij}=0$.
 This means that the $k$th component of $(\Delta G)_{ij}$ captures the difference between the networks $G_{ij}^+$ and $G_{ij}^-$ on the $k$th integrated statistic in the model.\\[0.3cm]
-Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
-If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$.
+%Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
+%If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$.
 This means that for $\theta_k$, $k \in \{1, \ldots , q\}$ the conditional odds of occurence increase if $\theta_k > 0 $, decrease if $\theta_k < 0 $ and stay the same if $\theta_k = 0 $.
-Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
-Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.
+%Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
+%Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.
+
+No matter which kind of MCMC algorithm is used, the basic idea is the following:
+\\One constructs a \textit{Markov chain} $(Y_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{A}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$.
+One can show that every single realization (or \textit{trajectory})
+%$$(A_t)_{t \in \mathbb{N}}:= (Y_t(\omega))_{t \in \mathbb{N}}$$
+of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}).
+As a result, sub-sequences of $(A_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
+Snijders \cite{Snijders.2002b} provides a Metropolis Hastings algorithm on how to simulate networks:
+Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
+\begin{enumerate}
+\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
+\item Compute the value
+$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
+\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
+\begin{itemize}
+\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
+\item $Z=1$, define $G^{(k+1)}$ via
+$$g_{pq}^{(k+1)}=\begin{cases}
+1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
+g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
+\end{cases}$$
+\end{itemize}
+\item Start at step 1 with $G^{(k+1)}$.
+\end{enumerate}
+The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$.
+Since the orignial matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
+the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
 
@@
 Furthermore, this method also adopts the advantage of the MPLE of being consistent.
-However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit especially if the size of the examined network is extremely large.
+However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit, especially if the size of the examined network is extremely large.
 

### 2017-03-19 small_network_fauxmesahigh/Coverage_Probability_fauxmesahigh.R
Add files via upload

@@ -115,7 +115,7 @@ bootSE.faux <- function( erg.obj, simulation, iters){
         logistic.coverage[m,j]=1
       }}
     
-    
+    # parametric bootstrap
     for ( i in 1:simulation){  # loop for numbers of simulations in each iteration
       
       net<- simulate.ergm(sim.mple, nsim=1)
@@ -200,9 +200,10 @@ bootSE.faux <- function( erg.obj, simulation, iters){
 
 # simulation MPLE
 set.seed(5555)
-sim=500
-iter=30
+sim=100
+iter=500
 faux.results.mple <- bootSE.faux(mple, sim,iter)
+coverage.mesa<- faux.results.mple
 
 # coef=Inf appears
 faux.results.mple[[1]][!is.finite(faux.results.mple[[1]])] <- NA
@@ -234,7 +235,9 @@ counts[2,2] <- mean(faux.results.mple[[5]][2,]) # sex mcmcmle
 counts[2,3] <- mean(faux.results.mple[[5]][3,]) # gwesp mcmcmle
 
 # Plot
-barplot(counts, beside=TRUE, main="Coverage Probability 95%", ylim=c(0,1),legend = rownames(counts),args.legend = list(x="bottomleft"))
+barplot(counts, beside=TRUE, main="Coverage Probability Faux Mesa High", ylim=c(0,1),
+        legend = rownames(counts),args.legend = list(x="bottomleft", cex=1.5), cex.axis=1.5,
+        cex.names = 2, cex.main=2)
 abline(h=0.95,col="grey", lty=2)
 
 # Histogram of the density of simulated networks

### 2017-03-19 TeX/paper.tex
scnetwork and coverage probability

--- 2017-03-14 TeX/paper.tex
@@
 This means neither the weighting nor the temporal change of ties is considered in the model.
-In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}).
+In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al.
+\cite{robins.pattison.2007}).
 In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm] %This chapter is mainly based on Harris \cite{Harris.}, Hunter et al.
@@
 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=.4]{Coverage_faux_mesa_high}
+\caption{The Coverage Probability results of the Faux Mesa High network for bootstrapped MPLE, MCMCMLE and logistic regression }
+\label{rmse}
+\end{center}
+\end{figure}
+
 \begin{itemize}
 
-\item include plot
+\item include magnolia plot
 \item Describe plots
@@
 
-\cite{FowlerJeon2008}, \cite{Fowleretal2007}
+In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMCMLE we apply both approaches to the U.S.
+Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al.
+\cite{Fowleretal2007}.
+The citation network we use consists of 9223 majority opinions written by the U.S.
+Supreme Court from 1954 to 2002.
+In this undirected network an edge occurs between two verticies (i.e.
+between two majority opinions) if one majority opinion cites the other.
+The endogenous statistics we include into this model are the number of edges, the edge-wise 0-shared partner (esp(0)) statistic and the dyad-wise 0-shared partner statistic (dsp(0)) (Snijders et al.
+\cite{SnijdersTomA.B..2006}, Hunter and Handcock \cite{Hunter.2006}).
+The esp(k) statistic counts the number of pairs which are directly linked and share exactly k partners.
+The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked.
+({\it add explanation for adding esp(0) and dsp(0)}).
+The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases.
+Beside these two nodal attributes we furthermore include the difference between the years of publication as an edge attribute.
+Since we do expect a non-linear effect fot this edge attribute, we include this variable as a second-degree polynomial.
 

### 2017-03-19 small_network_fauxmesahigh/Coverage_Probability_fauxmesahigh.R

@@ -201,7 +201,7 @@ bootSE.faux <- function( erg.obj, simulation, iters){
 # simulation MPLE
 set.seed(5555)
 sim=500
-iter=30
+iter=1000
 faux.results.mple <- bootSE.faux(mple, sim,iter)
 
 # coef=Inf appears

### 2017-03-25 TeX/paper.tex
Coverage_probability_plots

--- 2017-03-19 TeX/paper.tex
@@
 However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al.
-\cite{vanDuijn.2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE.
+\cite{vanDuijnetal2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE.
 Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
@@
 The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals.
-Duijn et al.
+van Duijn et al.
 show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%.
@@
 \\[0.3cm]
-Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=.45]{Coverage_Mesa_Magnolia}
+\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMCMLE and logistic regression }
+\label{coverage}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[scale=.40]{Bias_plots}
+\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
+\label{bias}
+\end{center}
+\end{figure}
+\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
 Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
-We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
+
+\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
 \\[0.3cm]
@@
 We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$.
-For each of the $1000$ networks we calculate the confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
+For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
 In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values.
-For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the MPLE estimates to obtain 95\% bootstrap confidence intervals.
+For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals.
 Similar as for the MCMCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
-Figures ref1 and ref2 visualize the coverage probabilities for each of the three methods for both networks.
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=.4]{Coverage_faux_mesa_high}
-\caption{The Coverage Probability results of the Faux Mesa High network for bootstrapped MPLE, MCMCMLE and logistic regression }
-\label{rmse}
-\end{center}
-\end{figure}
-
-\begin{itemize}
-
-\item include magnolia plot
-\item Describe plots
-\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
-\item can easily be parallelized/ using multiple cores -> increases speed
-\item this method is also able to detect degeneracy
-
-\end{itemize}
-
+\noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
+The dashed line is set at $0.95$ and represents the optimal value.
+It is in evidence that the bootstrapped MPLE performed equally well as the MCMCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases.
+Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable.
+Similar to the results of van Duijn et al \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
+Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
 \noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
@@
 
+\begin{itemize}
+
+\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
+\item can easily be parallelized/ using multiple cores -> increases speed
+\item this method is also able to detect degeneracy
+\item include table with boot MPLE results and MCMCMLE results
+\end{itemize}
+
+ \section*{Other chapters}
+
+ \begin{enumerate}
+ \item Fields that made fruitful use of the MPLE for networks, e.g. Boltzmann machine
+ \item Examples where a logit approach was used instead of an ERGM, because the network was too big and it was too time consuming using the MCMCMLE.
+We can then try to apply the bootstrapped MPLE approach.
+\end{enumerate}
+
  \section*{Results/Discussion}
@@
 
- \section*{To determine}
-
- \begin{enumerate}
- \item where to include Boltzmann machine application
-\end{enumerate}
-
 \newpage

### 2017-03-25 medium_network_fauxmagnoliahigh/Coverage_plot.R
Add files via upload

@@ -1,0 +1,132 @@
+library(statnet)
+# plot coverage probability plots next to each other
+
+par(mfrow=c(1,2))
+
+# Mesa High
+load("Coverage_mesahigh2.RData")
+
+# coef=Inf appears
+faux.results.mple[[1]][!is.finite(faux.results.mple[[1]])] <- NA
+faux.results.mple[[2]][!is.finite(faux.results.mple[[2]])] <- NA
+faux.results.mple[[3]][!is.finite(faux.results.mple[[3]])] <- NA
+faux.results.mple[[4]][!is.finite(faux.results.mple[[4]])] <- NA
+faux.results.mple[[5]][!is.finite(faux.results.mple[[5]])] <- NA
+
+
+## Coverage probability plot
+
+# create empty matrix for barplot values
+counts<- matrix(0,3,3)
+colnames(counts)<- c("Edges", "Sex", "Gwesp")
+rownames(counts)<- c("Bootstrap", "MCMCMLE", "Logistic")
+
+# fill matrix with corresponding values
+counts[1,1] <- mean(faux.results.mple[[2]][1,]) # edges bootstrap
+counts[1,2] <- mean(faux.results.mple[[2]][2,]) # sex bootstrap
+counts[1,3] <- mean(faux.results.mple[[2]][3,]) # gwesp bootstrap
+
+counts[3,1] <- mean(faux.results.mple[[3]][1,]) # edges logistic
+counts[3,2] <- mean(faux.results.mple[[3]][2,]) # sex logistic
+counts[3,3] <- mean(faux.results.mple[[3]][3,]) # gwesp logistic
+
+counts[2,1] <- mean(faux.results.mple[[5]][1,]) # edges mcmcmle
+counts[2,2] <- mean(faux.results.mple[[5]][2,]) # sex mcmcmle
+counts[2,3] <- mean(faux.results.mple[[5]][3,]) # gwesp mcmcmle
+
+# Plot
+barplot(counts, beside=TRUE, main="Coverage Probability Faux Mesa High", ylim=c(0,1),
+        legend = rownames(counts),args.legend = list(x="bottomleft", cex=1.3), cex.axis=1.5,
+        cex.names = 2, cex.main=1.5)
+abline(h=0.95,col="grey", lty=2, lwd=2)
+
+
+# Magnolia High
+load("Coverage_magnoliahigh2.RData")
+
+
+# coef=Inf appears
[…]

### 2017-03-31 TeX/paper.tex
scnetwork statistics

--- 2017-03-25 TeX/paper.tex
@@
 This approach does not deliberately ignore the dependency structures inside the network.
-Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.
+Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.
 %When it comes to the interpretation of the coefficients one can show that
@@
 
-\section*{Efficiency of MPLE/MCMCMLE}
-Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy.
+\section*{Efficiency of MPLE/MCMLE}
+Even though the MCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy.
 Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
-Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}).
+As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
 This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
-However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al.
-\cite{vanDuijnetal2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE.
-Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
-It is even more remarkable that the number of simulated networks needed for the MCMCMLE, in order to surpass the MPLE increases as the size of the network increases.
-This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE.
-In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\[0.3cm]
-\noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
+However, another major advantage of the MCMLE method is, as shown by van Duijn et al.
+\cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
+Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
+It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases.
+This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMLE to outperform the MPLE.
+In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
+\noindent In order to demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
 The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
-For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$.
+For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
 For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution.
-We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMCMLE of these simulated networks.
-For every single simulated network the MCMCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
+We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
+For every single simulated network the MCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
 Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance.
@@
 implying that the smaller the RMSE, the more accurate is the estimator.
-Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases.
+Since the MCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases.
 On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required.
-In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMCMLE to the MPLE.
-As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance.
+In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE.
+As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.
 Figure \ref{rmse} visualizes the results of the simulation study.
@@
 \includegraphics[scale=.5]{RMSE_paper}
-\caption{The log of the ratio of the RMSE for the MCMCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
+\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
 \label{rmse}
@@
 
-\noindent The plots support the fact that larger networks require a larger sample size of simulated networks in order for the MCMCMLE to outperform the MPLE.
-While the fairly small Faux Mesa High network only requires a sample size of about 50 networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMCMLE to surpass the MPLE.
+\noindent The plots support the fact that larger networks require a larger sample size of simulated networks in order for the MCMLE to outperform the MPLE.
+While the fairly small Faux Mesa High network only requires a sample size of about 50 networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE.
 These results propose especially for very large networks (e.g.
-social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMCMLE method.
+social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method.
 The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extends the mere computation time.
@@
 As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
-Moreover, the MPLE is able to outperform the MCMCMLE if the sample size is not chosen large enough.
+Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough.
 The main reason why the MCMCLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}.
@@
 \includegraphics[scale=.45]{Coverage_Mesa_Magnolia}
-\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMCMLE and logistic regression }
+\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
 \label{coverage}
@@
 We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$.
-For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
+For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
 In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values.
 For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals.
-Similar as for the MCMCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
+Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
 
@@
 The dashed line is set at $0.95$ and represents the optimal value.
-It is in evidence that the bootstrapped MPLE performed equally well as the MCMCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases.
+It is in evidence that the bootstrapped MPLE performed equally well as the MCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases.
 Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable.
@@
 
-In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMCMLE we apply both approaches to the U.S.
+In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the U.S.
 Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al.
@@
 The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked.
+Mathematically written, esp(k) and dsp(k) are defined as
+$$\Gamma_{esp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm}g_{ij} \right)\right]$$
+$$\Gamma_{dsp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm} \right)\right]$$
+where $m\neq i$, $m \neq j$ and $\mathds{1}_k(x)$ is an indicator that is defined as $1$, if $k=x$ and $0$, if $k\neq x$.
 ({\it add explanation for adding esp(0) and dsp(0)}). 
 The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases.
+In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
+The first row of matrix $C$ consists of the first actors attribute, repeated $N$ times.
+The second row of matrix $C$, consists of the second actors atribute, repeated $N$ times, and so on.
+Then, the statistics for a nodal covariate is defined as
+$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
 Beside these two nodal attributes we furthermore include the difference between the years of publication as an edge attribute.
-Since we do expect a non-linear effect fot this edge attribute, we include this variable as a second-degree polynomial.
+An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$.
+In this case entry $E_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$.
+Then, the corresponding statistic can be calculated the same way as nodal covariates.
+Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared componentwise.\\[0.3cm]
+We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
+The MCMLE required a sample size of ??
+networks to converge.
+The bootstrapped MPLE was estimated by using 500 simulated networks.
+As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks.
+For better comparison, be chose the same Burn-In (add number) and the same number of iterations (add number) until a new network is sampled.
+The results can be found in table ??.
+\\[0.3cm]
+Both methods were run on a single core....
 
@@
 
-\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
+\item describe advantage over MPLE (better coverage) and MCMLE (faster)
 \item can easily be parallelized/ using multiple cores -> increases speed
 \item this method is also able to detect degeneracy
-\item include table with boot MPLE results and MCMCMLE results
+\item include table with boot MPLE results and MCMLE results
 \end{itemize}
@@
  \item Fields that made fruitful use of the MPLE for networks, e.g. Boltzmann machine
- \item Examples where a logit approach was used instead of an ERGM, because the network was too big and it was too time consuming using the MCMCMLE.
+ \item Examples where a logit approach was used instead of an ERGM, because the network was too big and it was too time consuming using the MCMLE.
 We can then try to apply the bootstrapped MPLE approach.
@@
 
- \item  Bootstrap MPLE is an improvement of the MPLE and performs just as good as the MCMCMLE. \\
+ \item  Bootstrap MPLE is an improvement of the MPLE and performs just as good as the MCMLE. \\
 \item Bootstrap MPLE has the advantage of being fast + multiple cores\\ 

### 2017-04-02 TeX/paper.tex
degeneracy

--- 2017-03-31 TeX/paper.tex
@@
 For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution.
+For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.5$.\\
 We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 \begin{center}
-\includegraphics[scale=.45]{Coverage_Mesa_Magnolia}
+\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
 \caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
@@
 \begin{center}
-\includegraphics[scale=.40]{Bias_plots}
+\includegraphics[scale=.40]{bias_plots_05}
 \caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
@@
 As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks.
-For better comparison, be chose the same Burn-In (add number) and the same number of iterations (add number) until a new network is sampled.
+For better comparison, we chose the same Burn-In (add number) and the same number of iterations (add number) for sampling networks.
 The results can be found in table ??.
 \\[0.3cm]
-Both methods were run on a single core....
+$$\text{insert table with boot MPLE and MCMLE results}$$
+Both methods were run on a single core....\textcolor{blue}{(=> compare running time)}.
+The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem seperating the computation into several tasks.
+In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.\\[0.3cm]
+One of the major disadvantages of MPLE over MCMLE is that it can not assess {\it degeneracy}.
+When fitting an ERGM one usually has to deal with the problem of unreliable approximative likelihood estimates for the model's parameters.
+The reason why degeneray occurs is that the stochastic process gernerated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003}).
+The bootstrapped MPLE, however, allows assessing degenerate models as well.
+In order to verify whether a model is degenerate or not, one can take a look at density and trace plots as visualized in figure ??.
+The trace plots on the left side depict the the attained values via MCMC simulated networks for every single statistic included into the model, centered around the statistic values of the observed network.
+The plots on the right side visualize the empirical density function of the respective statistic, based on the simulated networks (Hunter and Handcock \cite{Hunter.2006}).
+For a non-degenerated model the empirical density function should be symmetrical around zerofor every included centered statistic, since the expected value of a centered statistic should be zero (compare equation (\ref{assumption})).
+Otherwise, the values of the simulated networks systematically differ from the corresponding statistics in the observed network, making it unreasonable to assume that the simulated networks originate from the same distribution as the observed network.
+Furthermore, the trajecories in the trace plot should not indicate a dependence structure.
+This would be a signal that the constructed stochastic process violates the Markov properties.
+$$\text{insert density plots and trace plots}$$
 

### 2017-04-07 TeX/paper.tex
introduction

@@
 \cite{robins.pattison.2007}).
-In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm] %This chapter is mainly based on Harris \cite{Harris.}, Hunter et al.
-\cite{Hunter.2008} and Hunter \cite{Hunter.}.
-\\[0.3cm]
+In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm]
 The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$.
@@
 This approach does not deliberately ignore the dependency structures inside the network.
-Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.
-%When it comes to the interpretation of the coefficients one can show that
-%\begin{equation}
-%\frac{\mathbb{P}_{\theta}(Y_{ij}=1 | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}=0 %| Y_{ij}^c=G_{ij}^c)}
-%= \exp(\theta_1 (\Delta_1 G)_{ij}) \cdot ... \cdot \exp( \theta_q (\Delta_q G)_{ij})
-%\label{Inter2}
-%\end{equation}
-%where the condition $Y_{ij}^c=G_{ij}^c$ is short for: $Y_{pq}=g_{pq}$ for all $(p,q)\in \{1,\dots,N\}^2$ with $(p,q) \neq (i,j)$ and refers to the rest of the network.
-%The expression $(\Delta G)_{ij}:=\Gamma(G_{ij}^+) - \Gamma(G_{ij}^-)$ is called the \textit{change statistic}.
-$G_{ij}^+$ emerges from $G$, while assuming $g_{ij}=1$ and $G_{ij}^-$ emerges from $G$, while assuming $g_{ij}=0$.
-This means that the $k$th component of $(\Delta G)_{ij}$ captures the difference between the networks $G_{ij}^+$ and $G_{ij}^-$ on the $k$th integrated statistic in the model.\\[0.3cm]
-%Equation (\ref{Inter2}) now enables a \textit{ceteris-paribus analysis} of the parameters in the model:\\
-%If the $k$th change statistic $(\Delta_k G)_{ij}$ increases one unit to $(\Delta_k G)_{ij}+1$, while all the other change statistics remain unchanged, the odds of occurrence of edge $(i,j)$, conditional on the rest of the network, is multiplied by the factor $\exp(\theta_k)$.
-This means that for $\theta_k$, $k \in \{1, \ldots , q\}$ the conditional odds of occurence increase if $\theta_k > 0 $, decrease if $\theta_k < 0 $ and stay the same if $\theta_k = 0 $.
-%Therefore, the interpretation of the parameter happens very similar to logistic regression analysis. 
-%Notice that for the sake of simplicity we did not condition on exogenous network statistics in this chapter.
+Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
 
@@
 One can show that every single realization (or \textit{trajectory})
-%$$(A_t)_{t \in \mathbb{N}}:= (Y_t(\omega))_{t \in \mathbb{N}}$$
 of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}).

### 2017-04-16 medium_network_fauxmagnoliahigh/Coverage_probability_magnoliahigh.R
Add files via upload

@@ -1,0 +1,373 @@
+library(statnet)
+
+# Code for the Coverage Probability Barplots. The Bootstrap CI's always seem to cover the 'true' coefficients
+
+
+##################################
+# Faux_mesa_high, 205 nodes
+##################################
+#qw<-ergm(faux.magnolia.high ~ edges +nodematch("Sex")+esp(0))
+#summary(qw)
+
+# get data
+data(faux.magnolia.high)
+
+mple <- try(ergm(faux.magnolia.high ~ edges +nodematch("Sex")+gwesp(0.5,fixed=TRUE), estimate = "MPLE" )) #
+summary(mple)
+
+#main function
+bootSE.faux <- function( erg.obj, simulation, iters){
+  
+  # create empty matrices for logistic coverage
+  logistic.coverage<- matrix(0,3, iters)
+  rownames(logistic.coverage)<- c("edges", "sex", "gwesp")
+  
+  # create empty matrices for bootstrap coverage
+  bootstrap.coverage<- matrix(0,3,iters)
+  rownames(bootstrap.coverage)<- c("edges", "sex", "gwesp")
+  
+  # create empty matrices for estimated MPLE
+  mple.sim<- matrix(0,3,simulation*iters)
+  rownames(mple.sim)<- c("edges", "sex", "gwesp")
+  
+  # create empty matrices for estimated MCMCMLE
+  mcmcmle.sim<- matrix(0,3,iters)
+  rownames(mcmcmle.sim)<- c("edges", "sex", "gwesp")
+  
+  # create empty matrices for logistic coverage
+  mcmcmle.coverage<- matrix(0,3, iters)
+  rownames(mcmcmle.coverage)<- c("edges", "sex", "gwesp")
+  
+  # create vector for density
+  density <- c()
+  
+  net.mple <- simulate.ergm(mple, nsim=iters, 
+             control=control.simulate.ergm(MCMC.burnin=5000000, MCMC.interval=3000000)) 
+  
+  for(j in 1:iters){ # loop for number of iterations
+    
+    print(j)
+    
+    #create empty matrices for bootstrap estimates
+    coef <- matrix(0, 3, simulation)
+    rownames(coef) <- c("edges", "sex", "gwesp")
+    Std.err <- matrix(0,3,simulation)
[…]

### 2017-04-16 small_network_fauxmesahigh/Coverage_Probability_fauxmesahigh.R
Add files via upload

@@ -11,7 +11,7 @@ library(statnet)
 # get data
 data(faux.mesa.high)
 
-mple <- try(ergm(faux.mesa.high ~ edges +nodematch("Sex")+gwesp(0,fixed=TRUE), estimate = "MPLE" )) #
+mple <- try(ergm(faux.mesa.high ~ edges +nodematch("Sex")+gwesp(0.25,fixed=TRUE), estimate = "MPLE" )) #
 summary(mple)
 
 #main function
@@ -54,13 +54,13 @@ bootSE.faux <- function( erg.obj, simulation, iters){
     net.mple <- simulate.ergm(mple, nsim=1)#, control=control.simulate.ergm(MCMC.burnin=5000000, MCMC.interval=100) ) # simulate network
     
     #estimate coefficients
-    sim.mple <- ergm(net.mple~edges+ nodematch("Sex")+gwesp(0,fixed=TRUE), estimate= "MPLE")
+    sim.mple <- ergm(net.mple~edges+ nodematch("Sex")+gwesp(0.25,fixed=TRUE), estimate= "MPLE")
     
     #############
     # MCMCMLE
     ############
     
-    bootest.mcmcmle <- try(ergm(net.mple~edges+ nodematch("Sex")+gwesp(0,fixed=TRUE))) #estimate coefficients
+    bootest.mcmcmle <- try(ergm(net.mple~edges+ nodematch("Sex")+gwesp(0.25,fixed=TRUE))) #estimate coefficients
     
     # save values
     if(summary(bootest.mcmcmle)[2]=="try-error"){ # in case the ERGM degenerates => safe NA
@@ -127,7 +127,7 @@ bootSE.faux <- function( erg.obj, simulation, iters){
       density[(j-1)*simulation+i]<- network.density(net)
       
       #estimate coefficients
-      bootest <- try(ergm(net~edges+ nodematch("Sex")+gwesp(0,fixed=TRUE), estimate= "MPLE"))
+      bootest <- try(ergm(net~edges+ nodematch("Sex")+gwesp(0.25,fixed=TRUE), estimate= "MPLE"))
       
       # save values
       if(summary(bootest)[2]=="try-error"){ # in case the ERGM degenerates => safe NA
@@ -200,9 +200,8 @@ bootSE.faux <- function( erg.obj, simulation, iters){
 
 # simulation MPLE
 set.seed(5555)
-
-sim=500
-iter=1000
+sim=100
[…]

### 2017-04-16 small_network_fauxmesahigh/Coverage_probability_magnoliahigh.R
Delete Coverage_probability_magnoliahigh.R

@@ -1,373 +1,0 @@
-library(statnet)
-
-# Code for the Coverage Probability Barplots. The Bootstrap CI's always seem to cover the 'true' coefficients
-
-
-##################################
-# Faux_mesa_high, 205 nodes
-##################################
-#qw<-ergm(faux.magnolia.high ~ edges +nodematch("Sex")+esp(0))
-#summary(qw)
-
-# get data
-data(faux.magnolia.high)
-
-mple <- try(ergm(faux.magnolia.high ~ edges +nodematch("Sex")+gwesp(0.5,fixed=TRUE), estimate = "MPLE" )) #
-summary(mple)
-
-#main function
-bootSE.faux <- function( erg.obj, simulation, iters){
-  
-  # create empty matrices for logistic coverage
-  logistic.coverage<- matrix(0,3, iters)
-  rownames(logistic.coverage)<- c("edges", "sex", "gwesp")
-  
-  # create empty matrices for bootstrap coverage
-  bootstrap.coverage<- matrix(0,3,iters)
-  rownames(bootstrap.coverage)<- c("edges", "sex", "gwesp")
-  
-  # create empty matrices for estimated MPLE
-  mple.sim<- matrix(0,3,simulation*iters)
-  rownames(mple.sim)<- c("edges", "sex", "gwesp")
-  
-  # create empty matrices for estimated MCMCMLE
-  mcmcmle.sim<- matrix(0,3,iters)
-  rownames(mcmcmle.sim)<- c("edges", "sex", "gwesp")
-  
-  # create empty matrices for logistic coverage
-  mcmcmle.coverage<- matrix(0,3, iters)
-  rownames(mcmcmle.coverage)<- c("edges", "sex", "gwesp")
-  
-  # create vector for density
-  density <- c()
-  
-  net.mple <- simulate.ergm(mple, nsim=iters, 
-             control=control.simulate.ergm(MCMC.burnin=5000000, MCMC.interval=3000000)) 
-  
-  for(j in 1:iters){ # loop for number of iterations
-    
-    print(j)
-    
-    #create empty matrices for bootstrap estimates
-    coef <- matrix(0, 3, simulation)
-    rownames(coef) <- c("edges", "sex", "gwesp")
-    Std.err <- matrix(0,3,simulation)
[…]

### 2017-04-16 TeX/paper.tex
relative running time

@@
 For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.5$.\\
+For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.25$.\\
 We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 $$\text{insert table with boot MPLE and MCMLE results}$$
-Both methods were run on a single core....\textcolor{blue}{(=> compare running time)}.
 The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem seperating the computation into several tasks.
-In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.\\[0.3cm]
+In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
+Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Supreme Court Citation (9223 nodes) for an increasing number of computing cores.
+For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate 1 million networks using $10,000$ MCMC steps in order to approximate the likelihood.
+The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit.
+We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x
+$$\text{boostrapped MPLE time} = \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}$$     
+and define the relative computing time as
+$$\text{relative computing time}= \cfrac{\text{bootstrapped MPLE time}}{\text{MCMLE time}}$$
+This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
+\begin{figure}
+\begin{center}
+\includegraphics[scale=.45]{rel_time}
+\caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
+\label{comptime}
+\end{center}
+\end{figure}
+Figure \ref{comptime} demonstrates that the small highschool friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
+Even more interesting is that the even larger Supreme Court Citation network only requires two cores for the bootstrapped MPLE to outperform the MCMLE.\\[0.3cm]
 One of the major disadvantages of MPLE over MCMLE is that it can not assess {\it degeneracy}.

### 2017-04-22 TeX/paper.tex
computation time

@@
 \end{equation}
-The problem with this simple estimation procedure is that the assumed hypothesis of the independence of the dyads turns out to be erroneous in most cases (see van Duijn et all ?citevanDuijn.
+The problem with this simple estimation procedure is that the assumed hypothesis of the independence of the dyads turns out to be erroneous in many cases (see van Duijn et al.
+\cite{vanDuijnetal2009}).
 This is a systematic problem: The presence of network data is inextricably connected with the presence of \textit{relational data}, which by definition should not be assumed to be independent of each other.
@@
 Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough.
-The main reason why the MCMCLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}.
+The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}.
 By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic.
@@
 Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable.
-Similar to the results of van Duijn et al \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
+Similar to the results of van Duijn et al.
+\cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
 Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
@@
 In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
-The first row of matrix $C$ consists of the first actors attribute, repeated $N$ times.
-The second row of matrix $C$, consists of the second actors atribute, repeated $N$ times, and so on.
+The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
+The second row of matrix $C$, consists of the second actor's atribute, repeated $N$ times, and so on.
 Then, the statistics for a nodal covariate is defined as
 $$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-Beside these two nodal attributes we furthermore include the difference between the years of publication as an edge attribute.
+Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute.
 An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$.
-In this case entry $E_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$.
+In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$.
 Then, the corresponding statistic can be calculated the same way as nodal covariates.
@@
 We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
-The MCMLE required a sample size of ??
-networks to converge.
+The MCMLE requires a sample size of at least $100000$ networks to converge.
 The bootstrapped MPLE was estimated by using 500 simulated networks.
 As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks.
-For better comparison, we chose the same Burn-In (add number) and the same number of iterations (add number) for sampling networks.
+For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks.
 The results can be found in table ??.
@@
 In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
-Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Supreme Court Citation (9223 nodes) for an increasing number of computing cores.
-For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate 1 million networks using $10,000$ MCMC steps in order to approximate the likelihood.
+Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Supreme Court Citation (9223 nodes) for an increasing number of computing cores.
+For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate 1 million networks using $10,000$ MCMC steps in order to approximate the likelihood appropriately.
 The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit.
-We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x
+We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:
 $$\text{boostrapped MPLE time} = \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}$$     
-and define the relative computing time as
+Based on this, we define the relative computing time as
 $$\text{relative computing time}= \cfrac{\text{bootstrapped MPLE time}}{\text{MCMLE time}}$$
@@
 \end{figure}
-Figure \ref{comptime} demonstrates that the small highschool friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
+Figure \ref{comptime} demonstrates that the two highschool friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
 Even more interesting is that the even larger Supreme Court Citation network only requires two cores for the bootstrapped MPLE to outperform the MCMLE.\\[0.3cm]
@@
 When fitting an ERGM one usually has to deal with the problem of unreliable approximative likelihood estimates for the model's parameters.
-The reason why degeneray occurs is that the stochastic process gernerated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003}).
+The reason why degeneray occurs is that the stochastic process gernerated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003} and Rinaldo et al.
+\cite{Rinaldo2009}).
 The bootstrapped MPLE, however, allows assessing degenerate models as well.
@@
 The plots on the right side visualize the empirical density function of the respective statistic, based on the simulated networks (Hunter and Handcock \cite{Hunter.2006}).
-For a non-degenerated model the empirical density function should be symmetrical around zerofor every included centered statistic, since the expected value of a centered statistic should be zero (compare equation (\ref{assumption})).
+For a non-degenerated model the empirical density function should be symmetrical around zero for every included centered statistic, since this corresponds with the expected value of a centered statistic (compare equation (\ref{assumption})).
 Otherwise, the values of the simulated networks systematically differ from the corresponding statistics in the observed network, making it unreasonable to assume that the simulated networks originate from the same distribution as the observed network.

### 2017-04-24 TeX/paper.tex

@@
 \usepackage{mdwlist}
+\usepackage{multirow}
+\usepackage[table,xcdraw]{xcolor}
 \usepackage{listings}
@@
 For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks.
-The results can be found in table ??.
-\\[0.3cm]
+The results can be found in table \ref{sc_results}.\\[0.3cm]
+
+\begin{table}[]
+\centering
+\begin{tabular}{|>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|}
+\hline
+\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{l|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   \\ \cline{2-5} 
+\multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error \\ \hline \hline
+Edges                                      &                                  &                                   & -5.657                           & 0.028                             \\ \hline
+Nodecov.Salience                           &                                  &                                   & 0.592                            & 0.012                             \\ \hline
+Nodecov.Year                               &                                  &                                   & 0.253                            & 0.012                             \\ \hline
+Absdiff.Year                               &                                  &                                   & -2.470                           & 0.160                             \\ \hline
+Absdiff.Year.Squared                       &                                  &                                   & 0.221                            & 0.066                             \\ \hline
+DSP(0)                                     &                                  &                                   & 0.0005                           & 0.0002                            \\ \hline
+ESP(0)                                     &                                  &                                   & -3.129                           & 0.012                             \\ \hline
+\end{tabular}
+\caption{Estimation results for the Supreme Court Network using the MCMLE and the bootstrapped MPLE}
+\label{sc_results}
+\end{table}
 $$\text{insert table with boot MPLE and MCMLE results}$$

### 2017-05-16 TeX/paper.tex

--- 2017-04-22 TeX/paper.tex
@@
 For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks.
-The results can be found in table ??.
-\\[0.3cm]
-$$\text{insert table with boot MPLE and MCMLE results}$$
-The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem seperating the computation into several tasks.
+The results can be found in table \ref{sc_results}.\\[0.3cm]
+\begin{table}[]
+\begin{tabular}{|
+>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|l|l|}
+\hline
+\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}Logistic Regression}     \\ \cline{2-7}
+\multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error \\ \hline
+Edges                                      & -5.657                           & 0.028                             &                                  &                                   & -4.568                           & 0.038                             \\ \hline
+Nodecov.Salience                           & 0.592                            & 0.012                             &                                  &                                   & 1.482                            & 0.009                             \\ \hline
+Nodecov.Year                               & 0.253                            & 0.012                             &                                  &                                   & 0.208                            & 0.015                             \\ \hline
+Absdiff.Year                               & -2.470                           & 0.160                             &                                  &                                   & -3.121                           & 0.194                             \\ \hline
+Absdiff.Year.Squared                       & 0.221                            & 0.066                             &                                  &                                   & 0.436                            & 0.080                             \\ \hline
+DSP(0)                                     & 0.0005                           & 0.0002                            &                                  &                                   & 0.021                            & 0.0002                            \\ \hline
+ESP(0)                                     & -3.129                           & 0.012                             &                                  &                                   & -2.162                           & 0.004                             \\ \hline
+\end{tabular}
+\caption{Estimation results for the Supreme Court Network using MCMLE, bootstrapped MPLE and logistic regression}
+\label{sc_results}
+\end{table}
+\noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem seperating the computation into several tasks.
 In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.

### 2017-05-16 TeX/paper.tex
consistency
consistency as described in shalize and rinaldo is different to
consistency in lindsay (1988)

--- 2017-04-24 TeX/paper.tex
@@
 \cite{robins.pattison.2007}).
-In the following, we will focus on directed networks, however, undirected networks can be introduced very similar.\\[0.3cm]
+In the following, we will focus on directed networks; however, undirected networks can be introduced very similar.\\[0.3cm]
 The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$.
@@
 Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements.
-Therefore, for an already small number of actors the cardinalty of $\mathcal{G}(N)$ turns out to be an astronomically large number.
+Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
 For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable.
@@
 This means that centering the vector of network statistics does not affect the probability function of the network variable $Y$.
-Therefore, in context of the likelihood function (\ref{loglik}) the vector of statistics can always be assumed to be centered around the observed network.\\
+Therefore, in context of the likelihood function (\ref{loglik}) the vector of statistics can always be assumed to be centered on the observed network.\\
 Due to assumption (\ref{assumption}), one gets from (\ref{loglik}) the simplified log-likelihood function
@@
 which sums up the weighted network statistics of all possible networks of $N$ nodes, has to be evaluated.
-Even for networks with small numbers of nodes this presents an enormous computational obstacle, and the necessary calculations for larger networks can not currently be completed in any reasonable timeframe.
+Even for networks with small numbers of nodes this presents an enormous computational obstacle, and the necessary calculations for larger networks cannot currently be completed in any reasonable timeframe.
 As a result, for sufficiently large networks it is not possible to estimate the parameters directly with the likelihood function.
@@
 This approach does not deliberately ignore the dependency structures inside the network.
-Furthermore, the Marcov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
+Furthermore, the Markov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
 
@@
 The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$.
-Since the orignial matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
+Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
 the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
@@
 Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
-As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
+As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
 This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
-However, another major advantage of the MCMLE method is, as shown by van Duijn et al.
+At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework.
+They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.
+The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is seen by Lindsay \cite{Lindsay1988}, i.e.
+instead of considering sub-networks, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks increases.
+This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
+\\
+Another major advantage of the MCMLE method is, as shown by van Duijn et al.
 \cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
@@
 \\[0.3cm]
-For the simulation study, we determine the MPLE for the model and treat these estimates as the networks 'true' parameter values.
+For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values.
 We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$.
@@
 Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
-\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
+\noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
 Furthermore, this method also adopts the advantage of the MPLE of being consistent.
-However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit, especially if the size of the examined network is extremely large.
+However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provides a computational benefit, especially if the size of the examined network is extremely large.
 
@@
 Supreme Court from 1954 to 2002.
-In this undirected network an edge occurs between two verticies (i.e.
+In this undirected network an edge occurs between two vertices (i.e.
 between two majority opinions) if one majority opinion cites the other.
@@
 The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
-The second row of matrix $C$, consists of the second actor's atribute, repeated $N$ times, and so on.
+The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on.
 Then, the statistics for a nodal covariate is defined as
@@
 Then, the corresponding statistic can be calculated the same way as nodal covariates.
-Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared componentwise.\\[0.3cm]
+Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared component wise.\\[0.3cm]
 We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
@@
 \begin{table}[]
-\centering
-\begin{tabular}{|>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|}
+\hspace*{-1.5cm}
+\begin{tabular}{|
+>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|l|l|}
 \hline
-\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{l|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   \\ \cline{2-5} 
+\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}Logistic Regression}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}     \\ \cline{2-7}
 \multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
 Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
-Error \\ \hline \hline
-Edges                                      &                                  &                                   & -5.657                           & 0.028                             \\ \hline
-Nodecov.Salience                           &                                  &                                   & 0.592                            & 0.012                             \\ \hline
-Nodecov.Year                               &                                  &                                   & 0.253                            & 0.012                             \\ \hline
-Absdiff.Year                               &                                  &                                   & -2.470                           & 0.160                             \\ \hline
-Absdiff.Year.Squared                       &                                  &                                   & 0.221                            & 0.066                             \\ \hline
-DSP(0)                                     &                                  &                                   & 0.0005                           & 0.0002                            \\ \hline
-ESP(0)                                     &                                  &                                   & -3.129                           & 0.012                             \\ \hline
+Error & \cellcolor[HTML]{EFEFEF}Lower Bound & \cellcolor[HTML]{EFEFEF}Upper Bound \\ \hline
+Edges                                      & -5.657                           & 0.028                             & -4.568                       & 0.038                                    & -5.160 & -4.530  \\ \hline
+Nodecov.Salience                           & 0.592                            & 0.012                             &   1.482                               & 0.009                                   & 1.413 & 1.890  \\ \hline
+Nodecov.Year                               & 0.253                            & 0.012                             &     0.208                             & 0.015                                   & 0.182 & 0.504  \\ \hline
+Absdiff.Year                               & -2.470                           & 0.160                             &  -3.121                                & 0.194                             & -3.308 & -2.434  \\ \hline
+Absdiff.Year.Squared                       & 0.221                            & 0.066                             &   0.436                               & 0.080                                   & -0.028 & 0.517  \\ \hline
+DSP(0)                                     & 0.0005                           & 0.0002                            &    0.021                              & 0.0002                                   & 0.019 & 0.031  \\ \hline
+ESP(0)                                     & -3.129                           & 0.012                             &    -2.162                              & 0.004                                   & -2.179 & -1.511  \\ \hline
 \end{tabular}
-\caption{Estimation results for the Supreme Court Network using the MCMLE and the bootstrapped MPLE}
+\caption{Estimation results for the Supreme Court Network using MCMLE, bootstrapped MPLE and logistic regression}
 \label{sc_results}
+\hspace*{-1cm}
 \end{table}
-$$\text{insert table with boot MPLE and MCMLE results}$$
-The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem seperating the computation into several tasks.
+
+\noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks.
 In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
@@
 \end{figure}
-Figure \ref{comptime} demonstrates that the two highschool friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
+Figure \ref{comptime} demonstrates that the two high school friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
 Even more interesting is that the even larger Supreme Court Citation network only requires two cores for the bootstrapped MPLE to outperform the MCMLE.\\[0.3cm]
-One of the major disadvantages of MPLE over MCMLE is that it can not assess  {\it degeneracy}.
-When fitting an ERGM one usually has to deal with the problem of unreliable approximative likelihood estimates for the model's parameters.
-The reason why degeneray occurs is that the stochastic process gernerated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003} and Rinaldo et al.
+One of the major disadvantages of MPLE over MCMLE is that it cannot assess {\it degeneracy}.
+When fitting an ERGM one usually has to deal with the problem of unreliable approximatively likelihood estimates for the model's parameters.
+The reason why degeneracy occurs is that the stochastic process generated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003} and Rinaldo et al.
 \cite{Rinaldo2009}).
@@
 In order to verify whether a model is degenerate or not, one can take a look at density and trace plots as visualized in figure ??.
-The trace plots on the left side depict the the attained values via MCMC simulated networks for every single statistic included into the model, centered around the statistic values of the observed network.
+The trace plots on the left side depict the the attained values via MCMC simulated networks for every single statistic included into the model, centered on the statistic values of the observed network.
 The plots on the right side visualize the empirical density function of the respective statistic, based on the simulated networks (Hunter and Handcock \cite{Hunter.2006}).
@@
 Otherwise, the values of the simulated networks systematically differ from the corresponding statistics in the observed network, making it unreasonable to assume that the simulated networks originate from the same distribution as the observed network.
-Furthermore, the trajecories in the trace plot should not indicate a dependence structure.
+Furthermore, the trajectories in the trace plot should not indicate a dependence structure.
 This would be a signal that the constructed stochastic process violates the Markov properties.

### 2017-05-23 TeX/paper.tex

@@
 As a result, sub-sequences of $(A_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
-Snijders \cite{Snijders.2002b} provides a Metropolis Hastings algorithm on how to simulate networks:
+Snijders \cite{Snijders.2002} provides a Metropolis Hastings algorithm on how to simulate networks:
 Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
@@
 instead of considering sub-networks, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks increases.
-This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
+\\%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
 \\
-Another major advantage of the MCMLE method is, as shown by van Duijn et al.
-\cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
-Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
+\textcolor{red}{Another major advantage of the MCMLE method is, as shown by van Duijn et al.
+\cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.-Rewrite} Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
 It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases.
@@
 For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.25$.\\
+For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(see Hunter \cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.25$.\\
 We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.

### 2017-05-26 large_network_scnetwork54/supreme_court_code_readin_5_26.R
Add files via upload

@@ -1,0 +1,59 @@
+### Edgelist gives case ids of citing and cited cases
+edgelist <- read.table("allcites.txt",header=F)
+## Need to add characters to ids so that they are not used as integers by network functions
+edgelist[,1] <- paste("c",edgelist[,1],sep="")
+edgelist[,2] <- paste("c",edgelist[,2],sep="")
+
+### Read in node attributes
+node_attributes <- read.csv("judicial.csv",stringsAsFactors=F)
+node_attributes[,1] <- paste("c",node_attributes$caseid,sep="")
+
+### Add additional attributes from the SCDB
+## See data documentation, http://scdb.wustl.edu/_brickFiles/2016_01/SCDB_2016_01_codebook.pdf
+SCDB <- read.csv("SCDB_2016_01_caseCentered_Citation.csv",stringsAsFactors=F)
+SCDBinFowler <- match(SCDB$usCite,node_attributes$usid)
+SCDB <- SCDB[!is.na(SCDBinFowler),]
+SCDBinFowler <- match(SCDB$usCite,node_attributes$usid)
+node_attributes$issueArea <- NA
+node_attributes$issueArea[SCDBinFowler] <- SCDB$issueArea
+node_attributes$lawType <- NA
+node_attributes$lawType[SCDBinFowler] <- SCDB$lawType
+
+### Create a netowrk object
+## The following sequence assures that edges are matched to the correct nodes
+library(network)
+## this is a directed acyclic graph, so it makes most senese to analyze it as undirected (i.e., since there can be no cycling)
+scnetwork1 <- network.initialize(nrow(node_attributes),dir=F)
+## set name to case ids, which will make it easy to add edges
+network.vertex.names(scnetwork) <- node_attributes[,1]
+
+# convert edgelist to a matrix
+edgelist <- as.matrix(edgelist)
+
+# match node indices to numeric place
+edge_row <- match(edgelist[,1],network.vertex.names(scnetwork))
+edge_col <- match(edgelist[,2],network.vertex.names(scnetwork))
+
+# estimate timing
+test_time <- system.time( for(i in 2:200){ scnetwork[cbind(edge_row[(i-1):i], edge_col[(i-1):i])] <- 1} )
+
+# convert to hours
+estimated_hours <- test_time/199*length(edge_row)/60/60
+
+# view estimated timing
+print(estimated_hours)
+
+# run all rows to store edges
[…]

### 2017-05-26 large_network_scnetwork54/supreme_court_code_readin.R
Delete supreme_court_code_readin.R

@@ -1,59 +1,0 @@
-### Edgelist gives case ids of citing and cited cases
-edgelist <- read.table("allcites.txt",header=F)
-## Need to add characters to ids so that they are not used as integers by network functions
-edgelist[,1] <- paste("c",edgelist[,1],sep="")
-edgelist[,2] <- paste("c",edgelist[,2],sep="")
-
-### Read in node attributes
-node_attributes <- read.csv("judicial.csv",stringsAsFactors=F)
-node_attributes[,1] <- paste("c",node_attributes$caseid,sep="")
-
-### Add additional attributes from the SCDB
-## See data documentation, http://scdb.wustl.edu/_brickFiles/2016_01/SCDB_2016_01_codebook.pdf
-SCDB <- read.csv("SCDB_2016_01_caseCentered_Citation.csv",stringsAsFactors=F)
-SCDBinFowler <- match(SCDB$usCite,node_attributes$usid)
-SCDB <- SCDB[!is.na(SCDBinFowler),]
-SCDBinFowler <- match(SCDB$usCite,node_attributes$usid)
-node_attributes$issueArea <- NA
-node_attributes$issueArea[SCDBinFowler] <- SCDB$issueArea
-node_attributes$lawType <- NA
-node_attributes$lawType[SCDBinFowler] <- SCDB$lawType
-
-### Create a netowrk object
-## The following sequence assures that edges are matched to the correct nodes
-library(network)
-## this is a directed acyclic graph, so it makes most senese to analyze it as undirected (i.e., since there can be no cycling)
-scnetwork1 <- network.initialize(nrow(node_attributes),dir=F)
-## set name to case ids, which will make it easy to add edges
-network.vertex.names(scnetwork) <- node_attributes[,1]
-
-# convert edgelist to a matrix
-edgelist <- as.matrix(edgelist)
-
-# match node indices to numeric place
-edge_row <- match(edgelist[,1],network.vertex.names(scnetwork))
-edge_col <- match(edgelist[,2],network.vertex.names(scnetwork))
-
-# estimate timing
-test_time <- system.time( for(i in 2:200){ scnetwork[cbind(edge_row[(i-1):i], edge_col[(i-1):i])] <- 1} )
-
-# convert to hours
-estimated_hours <- test_time/199*length(edge_row)/60/60
-
-# view estimated timing
-print(estimated_hours)
-
-# run all rows to store edges
[…]

### 2017-06-24 TeX/paper.tex
cosponsorship and altkstar

@@
 
-\section*{Supreme Court Citation Network Data}
+\section*{Cosponsorship Network Data}
 
-In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the U.S.
-Supreme Court Citation Network developed by Fowler and Jeon \cite{FowlerJeon2008} and Fowler et al.
-\cite{Fowleretal2007}.
-The citation network we use consists of 9223 majority opinions written by the U.S.
-Supreme Court from 1954 to 2002.
-In this undirected network an edge occurs between two vertices (i.e.
-between two majority opinions) if one majority opinion cites the other.
-The endogenous statistics we include into this model are the number of edges, the edge-wise 0-shared partner (esp(0)) statistic and the dyad-wise 0-shared partner statistic (dsp(0)) (Snijders et al.
-\cite{SnijdersTomA.B..2006}, Hunter and Handcock \cite{Hunter.2006}).
-The esp(k) statistic counts the number of pairs which are directly linked and share exactly k partners.
-The dsp(k) statistic is quite similar to esp(k), but does not require the two vertices to be directly linked.
-Mathematically written, esp(k) and dsp(k) are defined as
-$$\Gamma_{esp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm}g_{ij} \right)\right]$$
-$$\Gamma_{dsp(k)}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N} \left[\mathds{1}_{k} \left(\sum_{m=1}^{N} g_{im}g_{jm} \right)\right]$$
-where $m\neq i$, $m \neq j$ and $\mathds{1}_k(x)$ is an indicator that is defined as $1$, if $k=x$ and $0$, if $k\neq x$.
-({\it add explanation for adding esp(0) and dsp(0)}). 
+In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
+Senate and U.S.
+House of Representatives for the 93rd to 108th Congresses developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
+The cosponsorship network consists of 2635 nodes and contains information of over 280,000 pieces of legislation proposed in the U.S.
+House and Senate for the time period 1973 - 2004.
+In this network each node indicates on cosponsor and an edge indicates the link of each cosponsor of a piece of legislation to its sponsor.
+The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al.
+\cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock (2006) \cite{Hunter.2006}.
+The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic.
+In this model we fix the weight paramter $\lambda=0.4975$.
+Snijders et al.
+\cite{SnijdersTomA.B..2006} introduced an approach involving $k$-star statistics $S_1(A), \dots , S_{N-1}(A)$, where $S_k(A)$ denotes the number of $k$-stars in the network, $k \in \{1, \dots , N-1\}$.
+For simplicity, let us define
+$$S_k(A):=\Gamma_{star(k)}(A) $$
+\textit{define kstar}.
+Note that in every network $S_1(A)=\Gamma_{edges}(A)$, i.e., $S_1(A)$ is equal to the number of edges in the network.
+On this basis, Snijders introduces the \textit{alternating k-star statistics}
+\begin{equation*}
+\mathfrak{S}(A,\lambda):= \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(A)=    S_2(A)-\cfrac{S_3(A)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(A)}{\lambda^{N-3}}
+\end{equation*} 
+Models with this statistic and a fixed decay parameter turn out to be standard ERGMs and Hunter and Handcock \cite{Hunter.2006} succeeded in proving that one can also rewrite alternating k-stars as a function of a network's degree distribution
+\begin{equation}
+\mathfrak{S}(A,\lambda)= \lambda\Biggr(\lambda\sum_{j=1}^{N-1}\Biggr( 1-\biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(A) + 2S_1(A)\Biggl)
+\label{alto}
+\end{equation}
+where $D_j(A):= \Gamma_{deg(j)}(A)$ is the number of nodes with and degree of $j$. 
+In the next step, we define the \textit{geometrically weighted degree} statistic as the first summand of (\ref{alto})
+\begin{equation}
+\Gamma_{gwd}(A,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(A)
+\end{equation}
+At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
+\textbf{continue here}
 The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases.

### 2017-06-25 TeX/paper.tex
cosponsorship description

@@
 House and Senate for the time period 1973 - 2004.
-In this network each node indicates on cosponsor and an edge indicates the link of each cosponsor of a piece of legislation to its sponsor.
+In this network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
 The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al.
@@
 \textbf{continue here}
-The two nodal attributes we consider for this network are the year the majority opinion was drafted as well as an attribute that indicates whether a case appears on the Oxford list of salient cases.
+The nodal attribute that we implement for this network 
  In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.

### 2017-06-25 TeX/paper.tex
Add files via upload

@@
 At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
-\textbf{continue here}
-The nodal attribute that we implement for this network 
+As nodal attribute we implement the partisanship of the bill sponsor.
 In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
@@
 We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
-The MCMLE requires a sample size of at least $100000$ networks to converge.
+The MCMLE requires a sample size of at least $1000$ networks to converge.
 The bootstrapped MPLE was estimated by using 500 simulated networks.
 As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks.
-For better comparison, we chose the same Burn-In ($163840$ MH-steps) and the same number of iterations ($10240$ MH-steps) for sampling networks.
+For better comparison, we chose the same Burn-In ($300.000$ MH-steps) and the same number of iterations ($30.000$ MH-steps) for sampling networks.
 The results can be found in table \ref{sc_results}.\\[0.3cm]
@@
 \begin{table}[]
-\hspace*{-1.5cm}
+%\hspace*{-1.5cm}
 \begin{tabular}{|
@@
 Edges                                      & -5.657                           & 0.028                             & -4.568                       & 0.038                                    & -5.160 & -4.530  \\ \hline
-Nodecov.Salience                           & 0.592                            & 0.012                             &   1.482                               & 0.009                                   & 1.413 & 1.890  \\ \hline
-Nodecov.Year                               & 0.253                            & 0.012                             &     0.208                             & 0.015                                   & 0.182 & 0.504  \\ \hline
-Absdiff.Year                               & -2.470                           & 0.160                             &  -3.121                                & 0.194                             & -3.308 & -2.434  \\ \hline
-Absdiff.Year.Squared                       & 0.221                            & 0.066                             &   0.436                               & 0.080                                   & -0.028 & 0.517  \\ \hline
-DSP(0)                                     & 0.0005                           & 0.0002                            &    0.021                              & 0.0002                                   & 0.019 & 0.031  \\ \hline
-ESP(0)                                     & -3.129                           & 0.012                             &    -2.162                              & 0.004                                   & -2.179 & -1.511  \\ \hline
+Spnsor Party                          & 0.592                            & 0.012                             &   1.482                               & 0.009                                   & 1.413 & 1.890  \\ \hline
+Alternating k-star                              & 0.253                            & 0.012                             &     0.208                             & 0.015                                   & 0.182 & 0.504  \\ \hline
 \end{tabular}
-\caption{Estimation results for the Supreme Court Network using MCMLE, bootstrapped MPLE and logistic regression}
+\caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
 \label{sc_results}
-\hspace*{-1cm}
+%\hspace*{-1cm}
 \end{table}
-
 \noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks.

### 2017-06-26 Cosponsorship/MPLE_additional_dataExperimental.R
Add files via upload

@@ -1,4 +1,3 @@
-<<<<<<< HEAD
 #load(file="supreme_additional_data.RData")
 
 library(ergm)
@@ -79,7 +78,7 @@ for(i in 1:3){
     abline(h=net.stats[i])
 }
 
-### Code from here on needs to be revised ###
+
 
 # get mode and save observed stats
 edges.obs<- net.stats[1]
@@ -153,159 +152,16 @@ for(i in 1:5){
 
 median(time.mcmle)
 
-=======
-#load(file="supreme_additional_data.RData")
 
-library(ergm)
+## calculate the MPLE Bootstrap confidence intervals
 
-sponsorshipMatrix <- (as.matrix(read.csv("./cosponsorship2010/house_matrices/108_housematrix.txt",stringsAsFactors=F,header=F))==1)
+# simulate 500 networks from mple
 
-bills <- read.csv("./cosponsorship2010/house_bills.txt",stringsAsFactors=F,header=F)
+sim.mple <- simulate(mple, nsim=500)#, control=control.simulate(MCMC.burnin=3000000,MCMC.interval=30000))
 
-bills <- bills[which(bills[,1]==108),]
-
-
-sponsorParty <- rep(NA,ncol(sponsorshipMatrix))
-
-members <- read.csv("./cosponsorship2010/house_members/108_house.txt",stringsAsFactors=F,header=F)
-
-memberData <- read.csv("SH.csv",stringsAsFactors=F)
-memberData <- memberData[memberData$congress==108,]
-
-for(i in 1:length(sponsorParty)){
-    idSponsori <- as.numeric(gsub(" ","",members[which(sponsorshipMatrix[,i]),3]))
-    if(length(idSponsori)>0){
-        sponsorParty[i] <- memberData$party[match(idSponsori,memberData$ids)]
-    }
-    
+for ( i in 1:500){
+  
+  bootest<- ergm(sim.mple[[i]]~edges+ +nodematch("sponsorParty")+altkstar(0.4975,fixed=T), eval.loglik = FALSE, estimate="MPLE" ) 
+  
+  
 }
-
-cosponsorshipMatrix <- (as.matrix(read.csv("./cosponsorship2010/house_matrices/108_housematrix.txt",stringsAsFactors=F,header=F))==2)
-
-
-allSponsorMat <- sponsorshipMatrix + cosponsorshipMatrix
-
[…]

### 2017-06-26 TeX/paper.tex
Add files via upload

@@
 \section*{Efficiency of MPLE/MCMLE}
-Even though the MCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy.
+Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy.
 Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
@@
 They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.
-The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is seen by Lindsay \cite{Lindsay1988}, i.e.
-instead of considering sub-networks, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks increases.
+The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e.
+instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.
 \\%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
 \\
-\textcolor{red}{Another major advantage of the MCMLE method is, as shown by van Duijn et al.
-\cite{vanDuijnetal2009}, that the MCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.-Rewrite} Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
+Another major advantage of the MCMLE method is, as shown by van Duijn et al.
+\cite{vanDuijnetal2009}, that the MCMLE is substantially more efficient than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
+Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
 It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases.
-This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMLE to outperform the MPLE.
+This means that for very large networks it becomes difficult to determine the number of simulated networks required in order for the MCMLE to outperform the MPLE.
 In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
-\noindent In order to demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
-The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
+\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
+The data for both networks originates from Resnick et al.
+\cite{Resnicketal1997}.\\[0.3cm]
 For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(see Hunter \cite{Hunter.2006}) distribution were we fix the decay parameter $\lambda$ at $0.25$.\\
-We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
-For every single simulated network the MCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
+Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.\\
+We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
+For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10.000$ simulated networks used in the likelihood approximation.
 Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance.
 Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as
-$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$
+$$RMSE = \sqrt{\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$$
 implying that the smaller the RMSE, the more accurate is the estimator.
@@
 \begin{center}
-\includegraphics[scale=.5]{RMSE_paper}
+\includegraphics[scale=.5]{RMSE_10k}
 \caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
@@
 
-\noindent The plots support the fact that larger networks require a larger sample size of simulated networks in order for the MCMLE to outperform the MPLE.
-While the fairly small Faux Mesa High network only requires a sample size of about 50 networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE.
+\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
+While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE.
 These results propose especially for very large networks (e.g.
 social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method.
-The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extends the mere computation time.
+The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extend the mere computation time.
 
@@
 Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough.
-The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}.
+The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al.
+\cite{vanDuijnetal2009}).
 By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic.
-This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
+For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
 The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals.
 van Duijn et al.
-show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%.
+show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%.
 In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix.
 \\[0.3cm]
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
-\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
-\label{coverage}
-\end{center}
-\end{figure}
+\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
+Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
 
-\begin{figure}[H]
-\begin{center}
-\includegraphics[scale=.40]{bias_plots_05}
-\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
-\label{bias}
-\end{center}
-\end{figure}
-\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
-Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
-
-\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
+\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
 \\[0.3cm]
@@
 
+\begin{figure}[]
+\begin{center}
+\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
+\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
+\label{coverage}
+\end{center}
+\end{figure}
+
+\begin{figure}[]
+\begin{center}
+\includegraphics[scale=.40]{bias_plots_05}
+\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
+\label{bias}
+\end{center}
+\end{figure}
+
 \noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
 The dashed line is set at $0.95$ and represents the optimal value.
-It is in evidence that the bootstrapped MPLE performed equally well as the MCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases.
+It is evident that the bootstrapped MPLE performed equally well as the MCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases.
 Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable.
@@
 \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
-Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
+Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates.
+We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.\\
 \noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
@@
 
-In order to demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
+To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
 Senate and U.S.
@@
 $$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute.
+%Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute.
 An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$.
@@
 For better comparison, we chose the same Burn-In ($300.000$ MH-steps) and the same number of iterations ($30.000$ MH-steps) for sampling networks.
-The results can be found in table \ref{sc_results}.\\[0.3cm]
+The results can be found in table \ref{co_results}.\\
+It is interesting to see that the standard error calculated by the logistic regression approach is constantly smaller than the standard error of the MCMLE, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
+Furthermore, we can observe that the coefficients for the number of edges and the nodal attribute are very similar, but the estimates for the alternating k-star statistic do differ \textcolor{blue}{reason?}.
+The bootstrapped MPLE bounds, however, include each estimate for both methods.
+\\[0.3cm]
 
 \begin{table}[]
-%\hspace*{-1.5cm}
+\hspace*{-1.5cm}
 \begin{tabular}{|
@@
 Error & \cellcolor[HTML]{EFEFEF}Lower Bound & \cellcolor[HTML]{EFEFEF}Upper Bound \\ \hline
-Edges                                      & -5.657                           & 0.028                             & -4.568                       & 0.038                                    & -5.160 & -4.530  \\ \hline
-Spnsor Party                          & 0.592                            & 0.012                             &   1.482                               & 0.009                                   & 1.413 & 1.890  \\ \hline
-Alternating k-star                              & 0.253                            & 0.012                             &     0.208                             & 0.015                                   & 0.182 & 0.504  \\ \hline
+Edges                                      & -5.884 & 0.065                             & -5.869                       & 0.015                                    & -6.007 & -5.751  \\ \hline
+Sponsor Party                          & 1.440                            & 0.015                             &   1.440                               & 0.015                                   & 1.411 & 1.467  \\ \hline
+Alternating k-star                              & 0.125                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
 \end{tabular}
 \caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
-\label{sc_results}
-%\hspace*{-1cm}
+\label{co_results}
+\hspace*{-1cm}
 \end{table}
@@
 In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
-Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Supreme Court Citation (9223 nodes) for an increasing number of computing cores.
-For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate 1 million networks using $10,000$ MCMC steps in order to approximate the likelihood appropriately.
+Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Cosponsorship (2635 nodes) for an increasing number of computing cores.
+For the small network we simulate $2000$ networks using a MCMC interval of $2000$ steps, for the medium network we simulate $8000$ networks using a MCMC interval of $5000$ steps and for the large network we simulate $10000$ networks using $30,000$ MCMC steps in order to approximate the likelihood appropriately.
 The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit.
@@
 \end{figure}
-Figure \ref{comptime} demonstrates that the two high school friendship networks only require three cores for the bootstrapped MPLE to outperform the MCMLE.
-Even more interesting is that the even larger Supreme Court Citation network only requires two cores for the bootstrapped MPLE to outperform the MCMLE.\\[0.3cm]
+Figure \ref{comptime} demonstrates that all three networks only require two cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
+This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\
+Of course the actual computing time for a network always depends on the statistics that are included in the network, but in general larger networks require longer computation times, since a larger MCMC sample size is required and more MC steps are necessary to simulate a new network that does not overly depend on the previous sample.
+This makes the bootstrapped MPLE a beneficial alternative especially for very large networks.
+\\[0.3cm]
 One of the major disadvantages of MPLE over MCMLE is that it cannot assess {\it degeneracy}.

### 2017-06-26 Cosponsorship/Relative Time plot.R
Add files via upload

@@ -1,13 +1,41 @@
-<<<<<<< HEAD
 # comparison plot MCMLE vs MPLE
 
 # time in seconds
 
-# mesa high
+library(statnet)
 
-time.mple<- 0.05
-time.mcmle<- 12.59
-time.sim<- 1.573 # simulation time
+#############################
+# mesa high
+data("faux.mesa.high")
+
+time.mple.mesa<- c()
+set.seed(1)
+for(i in 1:100){
+  b<- system.time(ergm(faux.mesa.high~edges+nodefactor("Sex")+gwesp(0.25,fixed=TRUE), estimate="MPLE"))
+  time.mple.mesa[i]<- b[3]                  
+}
+
+median(time.mple.mesa)
+
+
+# MCMLE runtime
+
+time.mcmle.mesa<- c() # for storing the time
+
+set.seed(1)
+for(i in 1:20){
+  print(i)
+  b<- system.time(ergm(faux.mesa.high~edges+nodefactor("Sex")+gwesp(0.25,fixed=TRUE), 
+                       eval.loglik = FALSE, control = control.ergm(MCMC.interval = 2000,MCMC.samplesize=1000)))
+  time.mcmle.mesa[i]<- b[3]                  
+}
+
+median(time.mcmle.mesa)
+
+
+time.mple<- median(time.mple.mesa)
+time.mcmle<- median(time.mcmle.mesa)
+time.sim<- median(time.mcmle.mesa)/5 #takes 2 iterations, MCMC samplesize is quadrupled in last iteration
 
 # calculate bootMPLE time for 500 iterations and different number of cores (1, 5, 10, 25, 50, 100, 250, 500)
 
@@ -25,11 +53,38 @@ boot.mple.mesa[8]<- 500*time.mple/500 + time.sim
 
 rel.time.mesa<- boot.mple.mesa/time.mcmle
 
+########################################
 # magnolia high
+data("faux.magnolia.high")
 
-time.mple<- 0.51
-time.mcmle <- 119.07
-time.sim<- 2.49 # simulation time
+time.mple.magno<- c()
+set.seed(1)
+for(i in 1:100){
+  b<- system.time(ergm(faux.magnolia.high~edges+nodefactor("Sex")+gwesp(0.25,fixed=TRUE), estimate="MPLE"))
+  time.mple.magno[i]<- b[3]                  
+}
+
+median(time.mple.magno)
[…]

### 2017-06-26 Cosponsorship/MPLE_additional_dataExperimental.R
Add files via upload

@@ -157,11 +157,23 @@ median(time.mcmle)
 
 # simulate 500 networks from mple
 
-sim.mple <- simulate(mple, nsim=500)#, control=control.simulate(MCMC.burnin=3000000,MCMC.interval=30000))
+coefs <- matrix(0, 3, 10)
 
-for ( i in 1:500){
-  
-  bootest<- ergm(sim.mple[[i]]~edges+ +nodematch("sponsorParty")+altkstar(0.4975,fixed=T), eval.loglik = FALSE, estimate="MPLE" ) 
-  
-  
+set.seed(1111)
+sim.mple <- simulate(mple, nsim=10)#, control=control.simulate(MCMC.burnin=3000000,MCMC.interval=30000))
+
+for ( i in 1:10){
+  print(i)
+  bootest<- ergm(sim.mple[[i]]~edges +nodematch("sponsorParty")+altkstar(0.4975,fixed=T), eval.loglik = FALSE, estimate="MPLE" ) 
+  coefs[,i]<-coef(bootest)
 }
+
+# create empty matrix to store bootstrap.CI
+bootstrap.CI <- matrix(0,3,2)
+colnames(bootstrap.CI)<- c("upper", "lower")
+rownames(bootstrap.CI)<- c("edges", "sponsor", "altkstar")
+
+# calculate the bootstrap intervals
+bootstrap.CI[1,]<-quantile(coefs[1,], probs = c(0.975, 0.025)) # edges
+bootstrap.CI[2,]<-quantile(coefs[2,], probs = c(0.975, 0.025)) # sponsor
+bootstrap.CI[3,]<-quantile(coefs[3,], probs = c(0.975, 0.025)) # altkstar

### 2017-06-27 TeX/paper.tex
Add files via upload

@@
 Sponsor Party                          & 1.440                            & 0.015                             &   1.440                               & 0.015                                   & 1.411 & 1.467  \\ \hline
-Alternating k-star                              & 0.125                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
+Alternating k-star                              & 0.124                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
 \end{tabular}
@@
 The bootstrapped MPLE, however, allows assessing degenerate models as well.
-In order to verify whether a model is degenerate or not, one can take a look at density and trace plots as visualized in figure ??.
+In order to verify whether a model is degenerate or not, one can take a look at density and trace plots as visualized in figure \ref{diagnostics}.
 The trace plots on the left side depict the the attained values via MCMC simulated networks for every single statistic included into the model, centered on the statistic values of the observed network.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-$$\text{insert density plots and trace plots}$$
-
-\begin{itemize}
-
-\item describe advantage over MPLE (better coverage) and MCMLE (faster)
-\item can easily be parallelized/ using multiple cores -> increases speed
-\item this method is also able to detect degeneracy
-\item include table with boot MPLE results and MCMLE results
-\end{itemize}
+\begin{figure}
+\begin{center}
+\includegraphics[scale=.6]{MPLE_histogram_traceplot_cosponsorship}
+\caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.
+The thick line in both, the traceplots and the histograms, represents the network statistics of the observed network.
+}
+\label{diagnostics}
+\end{center}
+\end{figure}
 

### 2017-06-27 Cosponsorship/MPLE_additional_dataExperimental.R
Add files via upload

@@ -55,6 +55,8 @@ summary(mple)
 # simulate 500 networks
 set.seed(1234)
 sim.mple<- simulate.ergm(mple, nsim=500,  statsonly = TRUE, control=control.simulate.ergm(MCMC.interval=200000))
+
+
 # get observed statistics by simulating 25 times with MCMC.interval=MCMC.brunin=1. Then, take the mode
 
 d<- dim(sim.mple)[2] # save number of columns
@@ -64,6 +66,7 @@ set.seed(1)
 net.stats <- numeric(d)
 A <- matrix(0,25,d)
 for(i in 1:25){
+  print(i)
     set.seed(i*1000)
     simulated.network = simulate.ergm(mple, nsim=1, control=control.simulate.ergm(MCMC.burnin=1, MCMC.interval=1), statsonly = TRUE)
     A[i,]<- simulated.network
@@ -94,25 +97,25 @@ altkstar.obs
 pdf(file="MPLEDegeneracyCheck.pdf",width=4,height=8)
 par(mfrow=c(3,2))
 
-plot(sim.mple[,1], ylab="Value", type="l", main="Edges")
-abline(h=net.stats[1])
+plot(sim.mple[,1], ylab="Value", xlab="Sample" , type="l", main="Edges", cex.main=2, cex.lab=1.8, cex.axis=1.4)
+abline(h=net.stats[1], col='red', lwd=2)
 
-hist(sim.mple[,1], main="Edges", xlab="Value") 
-abline(v=edges.obs, col='red')
+hist(sim.mple[,1], main="Edges", xlab="Value", cex.main=2, cex.lab=1.8, cex.axis=1.4) 
+abline(v=edges.obs, col='red', lwd=3)
 
-plot(sim.mple[,2], ylab="Value", type="l", main="Sponsor Party")
-abline(h=net.stats[2])
+plot(sim.mple[,2], ylab="Value", xlab="Sample", type="l", main="Sponsor Party", cex.main=2, cex.lab=1.8, cex.axis=1.4)
+abline(h=net.stats[2], col='red', lwd=2)
 
-hist(sim.mple[,2], main="Sponsor Party", xlab="Value") 
-abline(v=sponsorParty.obs, col='red')
+hist(sim.mple[,2], main="Sponsor Party", xlab="Value", cex.main=2, cex.lab=1.8, cex.axis=1.4) 
+abline(v=sponsorParty.obs, col='red', lwd=3)
 
[…]

### 2017-07-05 TeX/paper.tex
Add files via upload

@@
 House of Representatives for the 93rd to 108th Congresses developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
-The cosponsorship network consists of 2635 nodes and contains information of over 280,000 pieces of legislation proposed in the U.S.
+The cosponsorship network consists of 2635 nodes (in this case senators) and contains information of over 280,000 pieces of legislation proposed in the U.S.
 House and Senate for the time period 1973 - 2004.
 In this network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
+This means that an edge emerges between senators $i$ and $j$ if senator $i$ cosponsors legislation sponsored by senator $j$.
 The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al.
@@
  \section*{Results/Discussion}
-\begin{itemize}
-
- \item  Bootstrap MPLE is an improvement of the MPLE and performs just as good as the MCMLE. \\
-\item Bootstrap MPLE has the advantage of being fast + multiple cores\\ 
-\item Bootstrap MPLE is consistent\\
-\item Conclusion: For very large networks the Bootstrap MPLE approach is more reasonable
-
-\end{itemize}
+In the past years the ERGM grew in popularity in many different fields and with this popularity also the size of the considered networks rose.
+As the size of the network increases, it becomes unclear how many simulated networks are necessary for the conventional MCMLE method in order to provide reliable results.
+Furthermore, as an increasing number of simulated networks is necessary for the MCMLE, the computing time rapidly goes sky high.
+For this reason it is essential to develop different methods that provide faster estimation than the MCMLE, but still lead to reliable results.\\
+In this paper we introduced the bootstrapped MPLE as an alternative estimation method for ERGMs and compared the performance to the commonly applied MCMLE.
+Based on a simulation study we first pointed out that the larger the size of a network is the larger the MCMC sample size has to be chosen in order for the MCMLE to outperform the fast and simple MPLE.
+However, the big disadvantage of the MPLE is that even though it is an approximately exact estimator, it underestimates the standard error.
+For this reason, we propose a parametric bootstrap method to attune the standard error.
+On the basis of another simulation study on two different networks we demonstrate that the bootstrapped MPLE covers the true coefficients just as well as the MCMLE, while the simple MPLE performs clearly poorer.
+This means that the bootstrapped MPLE combines the advantages of both methods, the MPLE and the MCMLE, because it is still simple, fast and provides approximately exact results, but also does not underestimate the standard error.
+We conclude that the bootstrapped MPLE should be regarded as a propitious alternative to the MCMLE.
+It also has the advantage of being embarrassingly parallel, which leads to a rapid speed-up of the calculation if multiple computing cores are used.
 

### 2017-07-17 TeX/conference_paper.tex
Big Data Conference paper
I have written the paper using the latex template that is required for the Big Data Conference

--- 2017-07-05 TeX/paper.tex
@@
-\documentclass[headsepline=true, abstracton]{scrartcl}
-%\usepackage[ngerman]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
+%% bare_conf.tex
+%% V1.3
+%% 2007/01/11
+%% by Michael Shell
+%% See:
+%% http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
+%% and
+%% http://www.ieee.org/
+
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE! 
+%% User assumes all risk.
+%% In no event shall IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%
+%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
+%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
+%%*************************************************************************
+
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. IEEE's font choices can trigger bugs that do  ***
+% *** not appear when using other class files.                            ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
+% Note that the a4paper option is mainly intended so that authors in
+% countries using A4 can easily print to A4 and see how their papers will
+% look in print - the typesetting of the document will not typically be
+% affected with changes in paper size (but the bottom and side margins will).
+% Use the testflow package mentioned above to verify correct handling of
+% both paper sizes by the user's LaTeX system.
+%
+% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
+% should be used if it is desired that the figures are to be displayed in
+% draft mode.
+%
+\documentclass[10pt, conference, compsocconf]{IEEEtran}
+% Add the compsocconf option for Computer Society conferences.
+%
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[conference]{../sty/IEEEtran}
+
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
+
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
+
+% *** CITATION PACKAGES ***
+%
+\usepackage{cite}
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
+% not currently provide for hyperlinked citations.
+% The latest version can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
+% The documentation is contained in the cite.sty file itself.
+
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+   \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation can
+% be obtained at: 
+% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
+% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
+
+% *** MATH PACKAGES ***
+%
+\usepackage[cmex10]{amsmath}
 \usepackage{amssymb}
-\usepackage{amsmath}
 \usepackage{amsthm}
-\usepackage{enumerate}
-\usepackage{verbatim}
-\usepackage[a4paper,text={140mm,215mm},centering,headsep=10mm,footskip=10mm]{geometry}
-\usepackage[urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black,bookmarks]{hyperref}
-\usepackage{aliascnt}
-\usepackage{lmodern}
-\usepackage{mdwlist}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics. If using
+% it, be sure to load this package with the cmex10 option to ensure that
+% only type 1 fonts will utilized at all point sizes. Without this option,
+% it is possible that some math symbols, particularly those within
+% footnotes, will be rendered in bitmap form which will result in a
+% document that can not be IEEE Xplore compliant!
+%
+% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
+% There is also a support site at:
+% http://algorithms.berlios.de/index.html
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
+
+% *** ALIGNMENT PACKAGES ***
+%
+\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/required/tools/
+
+%\usepackage{mdwmath}
+%\usepackage{mdwtab}
+% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
+% especially mdwmath.sty and mdwtab.sty which are used to format equations
+% and tables, respectively. The MDWtools set is already installed on most
+% LaTeX systems. The lastest version and documentation is available at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+%\usepackage{eqparbox}
+% Also of notable interest is Scott Pakin's eqparbox package for creating
+% (automatically sized) equal width boxes - aka "natural width parboxes".
+% Available at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
+
+% *** SUBFIGURE PACKAGES ***
+%\usepackage[tight,footnotesize]{subfigure}
+% subfigure.sty was written by Steven Douglas Cochran. This package makes it
+% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
+% work, it is a good idea to load it with the tight package option to reduce
+% the amount of white space around the subfigures. subfigure.sty is already
+% installed on most LaTeX systems. The latest version and documentation can
+% be obtained at:
+% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
+% subfigure.sty has been superceeded by subfig.sty.
+
+%\usepackage[caption=false]{caption}
+%\usepackage[font=footnotesize]{subfig}
+% subfig.sty, also written by Steven Douglas Cochran, is the modern
+% replacement for subfigure.sty. However, subfig.sty requires and
+% automatically loads Axel Sommerfeldt's caption.sty which will override
+% IEEEtran.cls handling of captions and this will result in nonIEEE style
+% figure/table captions. To prevent this problem, be sure and preload
+% caption.sty with its "caption=false" package option. This is will preserve
+% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
+% (recommended due to many improvements over 1.2) of subfig.sty supports
+% the caption=false option directly:
+%\usepackage[caption=false,font=footnotesize]{subfig}
+%
+% The latest version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
+% The latest version and documentation of caption.sty can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
+
+% *** FLOAT PACKAGES ***
+%
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure. The latest version and documentation can be found at:
+% http://www.ctan.org/tex-archive/macros/latex/base/
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
+% Documentation is contained in the stfloats.sty comments as well as in the
+% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
+% does not allow \baselineskip to stretch. Authors submitting work to the
+% IEEE should note that IEEE rarely uses double column equations and
+% that authors should try to avoid such use. Do not be tempted to use the
+% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
+% not format its papers in such ways.
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
+% Read the url.sty source comments for usage information. Basically,
+% \url{my_url_here}.
+
+\usepackage[table,xcdraw]{xcolor}
 \usepackage{multirow}
-\usepackage[table,xcdraw]{xcolor}
-\usepackage{listings}
-\usepackage[activate]{pdfcprot}
-\usepackage{graphicx}
-\usepackage{slashed}
-\usepackage{mathrsfs}
-\usepackage{geometry}
-\usepackage{float}
-\usepackage{oldgerm}
-\usepackage{setspace}
-\usepackage{dsfont}
-%\usepackage{mathtools}
-\usepackage[all]{xy}
-\usepackage{cite}
-\usepackage{url}
-\pagestyle{headings}
-\newcommand{\myclearpage}{\clearpage}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\newenvironment{gelaber}{}{}
-\newenvironment{preamble}{}{}
-\newcommand{\tostar}{\overset{*}{\lower0.5em\hbox{$\smash{\scriptscriptstyle\rightharpoonup}$}}}
-\newtheorem{mydef}{Definition}
-\newtheorem{bem}{Bemerkung}
-\usepackage{tikz}
-\newcommand\circlearound[1]{%
-  \tikz[baseline]\node[draw,shape=circle,anchor=base] {#1} ;}
+\usepackage{multicol}
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
+
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
 
 \begin{document}
+%
+% paper title
+% can use linebreaks \\ within to get better formatting as desired
+\title{Exponential Random Graph Models with Big Networks:\\ Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}
 
-\setcounter{section}{-1}
+% author names and affiliations
+% use a multiple column layout for up to two different
+% affiliations
 
-\renewcommand{\refname}{Bibliography}
+\author{\IEEEauthorblockN{Christian S. Schmid}
+\IEEEauthorblockA{Department of Statistics\\
+The Pennsylvania State University\\
+State College, USA\\
+Email: schmid@psu.edu}
+\and
+\IEEEauthorblockN{Bruce A. Desmarais}
+\IEEEauthorblockA{Department of Political Science\\
+The Pennsylvania State University\\
+State College, USA\\
+Email:  bdesmarais@psu.edu}
+}
 
-\onehalfspacing
-\setlength{\headsep}{15mm}
+% conference papers do not typically use \thanks and this command
+% is locked out in conference mode. If really needed, such as for
+% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
+% after \documentclass
 
-\thispagestyle{plain}
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
+% 
+%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
+%Homer Simpson\IEEEauthorrefmark{2},
+%James Kirk\IEEEauthorrefmark{3}, 
+%Montgomery Scott\IEEEauthorrefmark{3} and
+%Eldon Tyrell\IEEEauthorrefmark{4}}
+%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
+%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
 
-\title{\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap}
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
+
+% make the title area
 \maketitle
@@
 We compare the two different approaches by applying them on U.S.
-Supreme Court Citation Network Data.
+Cosponsorship Network Data.
+
 \end{abstract}
 
- \section*{Introduction}
+\begin{IEEEkeywords}
+network, ERGM, parametric bootstrap, maximum pseudolikelihood
 
+\end{IEEEkeywords}
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
+
+\section{Introduction}
 The field of network science faces a double-edge sword when it comes to computational limitations to innovation.
@@
 
- \section*{The Exponential Random Graph Model}
-The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks.
+\section{The Exponential Random Graph Model}
+\noindent The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks.
 This means neither the weighting nor the temporal change of ties is considered in the model.
@@
 
-\section*{Estimation}
+\section{Estimation}
 As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology. Therefore, the question arises how one can estimate the parameter vector $\theta$?
@@
 %
-\begin{equation}
-\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
-~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] = \frac{c(\theta)}{c(\theta_0)}
-\label{konver}
+\begin{equation}\label{konver}
+\begin{split}
+\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right) &
+~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~\\ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]& = \frac{c(\theta)}{c(\theta_0)}
+\end{split}
 \end{equation}
@@
 \begin{eqnarray*}
-\text{loglik}(\theta)-\text{loglik}(\theta_0)&=&- \log(c(\theta))+\log(c(\theta_0))\\
+&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                              &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
@@
 
-\section*{Efficiency of MPLE/MCMLE}
+\section{Efficiency of MPLE and MCMLE}
 Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy.
@@
 The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=.5]{RMSE_10k}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3.5in]{RMSE_10k}
 \caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
 \label{rmse}
-\end{center}
 \end{figure}
-
 \noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
@@
 
-\section*{Bootstrapped MPLE}
-As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
+\section{Bootstrapped MPLE}
+\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
 Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough.
@@
 Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
-
+\\[0.3cm]
 \noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
@@
 Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
-
-\begin{figure}[]
-\begin{center}
-\includegraphics[scale=.5]{Coverage_Mesa_Magnolia_height05}
-\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression }
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
+\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
 \label{coverage}
-\end{center}
 \end{figure}
 
-\begin{figure}[]
-\begin{center}
-\includegraphics[scale=.40]{bias_plots_05}
-\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3.5in]{bias_plots_05}
+\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right)}
 \label{bias}
-\end{center}
 \end{figure}
-
 \noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
@@
 
-\section*{Cosponsorship Network Data}
-
+\section{Cosponsorship Network Data}
+\begin{table*}[!t]
+%% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\label{co_results}
+\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+\begin{tabular}{|
+>{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|l|l|}
+\hline
+\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}Logistic Regression}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}     \\ \cline{2-7}
+\multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
+Error & \cellcolor[HTML]{EFEFEF}Lower Bound & \cellcolor[HTML]{EFEFEF}Upper Bound \\ \hline
+Edges                                      & -5.884 & 0.065                             & -5.869                       & 0.015                                    & -6.007 & -5.751  \\ \hline
+Sponsor Party                          & 1.440                            & 0.015                             &   1.440                               & 0.015                                   & 1.411 & 1.467  \\ \hline
+Alternating k-star                              & 0.124                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
+\end{tabular}
+\caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
+\end{table*}
 To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
@@
 \begin{equation*}
-\mathfrak{S}(A,\lambda):= \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(A)=    S_2(A)-\cfrac{S_3(A)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(A)}{\lambda^{N-3}}
+\begin{split}
+\mathfrak{S}(A,\lambda):= & \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(A)\\= & ~   S_2(A)-\cfrac{S_3(A)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(A)}{\lambda^{N-3}}
+\end{split}
 \end{equation*} 
@@
 At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
+
 As nodal attribute we implement the partisanship of the bill sponsor.
@@
 $$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-%Beside these two nodal attributes, we furthermore include the difference between the years of publication as an edge attribute.
-An edge attribute can be written as a matrix $E$ of the same dimensions as the observed adjacency matrix $G$.
-In this case entry $e_{ij}$ is filled with the absolute difference between the year of case $i$ and case $j$.
-Then, the corresponding statistic can be calculated the same way as nodal covariates.
-Since we do expect a non-linear effect for this edge attribute, we include this variable as a second-degree polynomial by creating a matrix $E^2$, where the matrix gets squared component wise.\\[0.3cm]
 We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
@@
 It is interesting to see that the standard error calculated by the logistic regression approach is constantly smaller than the standard error of the MCMLE, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
-Furthermore, we can observe that the coefficients for the number of edges and the nodal attribute are very similar, but the estimates for the alternating k-star statistic do differ \textcolor{blue}{reason?}.
+Furthermore, we can observe that the coefficients for the number of edges and the nodal attribute are very similar, but the estimates for the alternating k-star statistic do differ \textit{(REASON FOR altkstar?)}.
 The bootstrapped MPLE bounds, however, include each estimate for both methods.
-\\[0.3cm]
+\\
 
-\begin{table}[]
-\hspace*{-1.5cm}
-\begin{tabular}{|
->{\columncolor[HTML]{EFEFEF}}l |l|l|l|l|l|l|}
-\hline
-\cellcolor[HTML]{EFEFEF}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}MCMLE}                   & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}Logistic Regression}       & \multicolumn{2}{c|}{\cellcolor[HTML]{EFEFEF}bootstrapped MPLE}     \\ \cline{2-7}
-\multirow{-2}{*}{\cellcolor[HTML]{EFEFEF}} & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
-Error & \cellcolor[HTML]{EFEFEF}Estimate & \cellcolor[HTML]{EFEFEF}St.
-Error & \cellcolor[HTML]{EFEFEF}Lower Bound & \cellcolor[HTML]{EFEFEF}Upper Bound \\ \hline
-Edges                                      & -5.884 & 0.065                             & -5.869                       & 0.015                                    & -6.007 & -5.751  \\ \hline
-Sponsor Party                          & 1.440                            & 0.015                             &   1.440                               & 0.015                                   & 1.411 & 1.467  \\ \hline
-Alternating k-star                              & 0.124                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
-\end{tabular}
-\caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
-\label{co_results}
-\hspace*{-1cm}
-\end{table}
 \noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks.
@@
 We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:
-$$\text{boostrapped MPLE time} = \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}$$     
+\begin{equation*}
+\begin{split}
+\text{boostrapped MPLE time} = ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ \\ \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}
+\end{split}
+\end{equation*}     
 Based on this, we define the relative computing time as
@@
 This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.45]{rel_time}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3in]{rel_time}
 \caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
 \label{comptime}
-\end{center}
 \end{figure}
+
 Figure \ref{comptime} demonstrates that all three networks only require two cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-\begin{figure}
-\begin{center}
-\includegraphics[scale=.6]{MPLE_histogram_traceplot_cosponsorship}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3in]{MPLE_histogram_traceplot_cosponsorship}
 \caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.
@@
 \label{diagnostics}
-\end{center}
 \end{figure}
 
- \section*{Other chapters}
-
- \begin{enumerate}
- \item Fields that made fruitful use of the MPLE for networks, e.g. Boltzmann machine
- \item Examples where a logit approach was used instead of an ERGM, because the network was too big and it was too time consuming using the MCMLE.
-We can then try to apply the bootstrapped MPLE approach.
-\end{enumerate}
-
- \section*{Results/Discussion}
-In the past years the ERGM grew in popularity in many different fields and with this popularity also the size of the considered networks rose.
+\section{Conclusion}
+\noindent In the past years the ERGM grew in popularity in many different fields and with this popularity also the size of the considered networks rose.
 As the size of the network increases, it becomes unclear how many simulated networks are necessary for the conventional MCMLE method in order to provide reliable results.
@@
 
-\newpage
+% An example of a floating figure using the graphicx package.
+% Note that \label must occur AFTER (or within) \caption.
+% For figures, \caption should occur after the \includegraphics.
+% Note that IEEEtran v1.7 and later has special internal code that
+% is designed to preserve the operation of \label within \caption
+% even when the captionsoff option is in effect. However, because
+% of issues like this, it may be the safest practice to put all your
+% \label just after \caption rather than within \caption{}.
+%
+% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
+% option should be used if it is desired that the figures are to be
+% displayed while in draft mode.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation Results}
+%\label{fig_sim}
+%\end{figure}
+
+% Note that IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
+
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command, the
+% \label for the overall figure must come after \caption.
+% \hfil must be used as a separator to get equal spacing.
+% The subfigure.sty package works much the same way, except \subfigure is
+% used instead of \subfloat.
+%
+%\begin{figure*}[!t]
+%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+%\label{fig_second_case}}}
+%\caption{Simulation results}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
+
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
+%\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that IEEE does not put floats in the very first column - or typically
+% anywhere on the first page for that matter. Also, in-text middle ("here")
+% positioning is not used. Most IEEE journals/conferences use top floats
+% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
+% command of the stfloats package.
+
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
+
+% references section
+
+% can use a bibliography generated by BibTeX as a .bbl file
+% BibTeX documentation can be easily obtained at:
+% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
+% The IEEEtran BibTeX style support page is at:
+% http://www.michaelshell.org/tex/ieeetran/bibtex/
+%\bibliographystyle{IEEEtran}
+% argument is your BibTeX string definitions and bibliography database(s)
+%\bibliography{IEEEabrv,../bib/paper}
+%
+% <OR> manually copy in the resultant .bbl file
+% set second argument of \begin to the number of references
+% (used to reserve space for the reference number labels box)
+%\begin{thebibliography}{1}
+
+%\bibitem{IEEEhowto:kopka}
+%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.%\hskip 1em plus
+ % 0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, %1999.
+
+%\end{thebibliography}
 
@@
 
+% that's all folks
 \end{document}

### 2017-07-18 TeX/conference_paper.tex
Big data conference template
Paper in Big data conference template. I also added a more precise explanation of included network statistics

@@
 \usepackage{amsthm}
+\usepackage{dsfont}
 % A popular package from the American Mathematical Society that provides
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Exponential Random Graph Models with Big Networks:\\ Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}
+\title{Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}
 
@@
 \section{Introduction}
-The field of network science faces a double-edge sword when it comes to computational limitations to innovation.
+\noindent The field of network science faces a double-edge sword when it comes to computational limitations to innovation.
 First, the networks under study are growing larger with each year.
@@
 \cite{robins.pattison.2007}).
-In the following, we will focus on directed networks; however, undirected networks can be introduced very similar.\\[0.3cm]
+In the following, we will introduce the ERGM for directed networks; however, undirected networks can be introduced very similar.\\[0.3cm]
 The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$.
@@
 For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable.
-As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods.\\
+As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods as we will intriduce in the next chapter.\\
 With the definition of $\mathcal{G}(N)$ we define 
 $$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
-as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N_V)$ we define the ERGM as
+as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N)$ we define the ERGM as
 \begin{equation}
@@
 \section{Estimation}
-As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology. Therefore, the question arises how one can estimate the parameter vector $\theta$?
+\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology. Therefore, the question arises how one can estimate the parameter vector $\theta$?
 A first idea could be the following: One can assume that the \textit{dyads} are independent of each other, which means that the random variables $Y_{ij}$ inside the random matrix $Y$ are independent of each other.
@@
 This technique tends to underestimate the standard error.
-However, Desmarais and Cranmer \cite{Desmarais.2012} show that the pseudo-likelihood provides a consistent approximation of the maximum likelihood, meaning that the MPLE converges to the MLE as the size of the network increases.\\[0.3cm]
+However, Desmarais and Cranmer \cite{Desmarais.2012} argue that the pseudo-likelihood provides a consistent approximation of the maximum likelihood, meaning that the MPLE converges to the MLE as the size of the network increases.\\[0.3cm]
 There are several techniques to circumvent estimators, which underestimate the standard error of $\theta$.
@@
 %
-This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm}.
-As a result, the approximate maximum-likelihood estimator for the parameters can be computed.\\[0.4cm]
+This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
+As a result, the approximate maximum-likelihood estimator for the parameters can be computed.\\[0.3cm]
 As pleasant as this may sound, the immediate question arises: How can a sufficient number of suitable drawings $G_1,...,G_L$ be taken from the distribution $\mathbb{P}_{\theta_0}$? \\
@@
 Furthermore, the Markov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
-
 No matter which kind of MCMC algorithm is used, the basic idea is the following:
-\\One constructs a \textit{Markov chain} $(Y_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{A}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$.
+\\One constructs a \textit{Markov chain} $(G_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{G}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$.
 One can show that every single realization (or \textit{trajectory})
 of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}).
-As a result, sub-sequences of $(A_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
+As a result, sub-sequences of $(G_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
 Snijders \cite{Snijders.2002} provides a Metropolis Hastings algorithm on how to simulate networks:
@@
 \end{enumerate}
-The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$.
+\vspace{0.3cm}
+\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$.
 Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
@@
 \section{Efficiency of MPLE and MCMLE}
-Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy.
+\noindent Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy.
 Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
 As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
-This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
+This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\[0.3cm]
 At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework.
@@
 instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.
-\\%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
+\\[0.3cm]%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
 \\
-Another major advantage of the MCMLE method is, as shown by van Duijn et al.
+%A major advantage of the MCMLE method is, as shown by van Duijn et al.
 \cite{vanDuijnetal2009}, that the MCMLE is substantially more efficient than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
-Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
+Nevertheless, as shown by
+Next, we are going to compare the efficiency of the MCMLE compared to the MPLE.
+Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
 It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases.
@@
 For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.\\
+Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.
+The number of edges in these unidrected networks are defined as
+$$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
+In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
+The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
+The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on.
+Then, the statistics for a nodal covariate is defined as
+$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
+A precise description of the gwesp statistic will be given in the chapter VI.\\[0.3cm]
 We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE.
-As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.
+As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\[0.3cm]
 Figure \ref{rmse} visualizes the results of the simulation study.
@@
 \centering
-\includegraphics[width=3.5in]{RMSE_10k}
+\includegraphics[width=3.6in]{RMSE_10k}
 \caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
@@
 \noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
-\\[0.3cm]
 For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values.
@@
 \centering
-\includegraphics[width=3.5in]{bias_plots_05}
+\includegraphics[width=3.5in, height=1.75in]{bias_plots_05}
 \caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right)}
@@
 Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates.
-We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.\\
+We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.
+It is especially worthwile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.\\
 \noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
@@
 \end{table*}
-To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
+\noindent To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
 Senate and U.S.
@@
 House and Senate for the time period 1973 - 2004.
-In this network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
+In this undirected network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
 This means that an edge emerges between senators $i$ and $j$ if senator $i$ cosponsors legislation sponsored by senator $j$.
@@
 Snijders et al.
-\cite{SnijdersTomA.B..2006} introduced an approach involving $k$-star statistics $S_1(A), \dots , S_{N-1}(A)$, where $S_k(A)$ denotes the number of $k$-stars in the network, $k \in \{1, \dots , N-1\}$.
+\cite{SnijdersTomA.B..2006} introduced an approach involving $k$-star statistics $S_1(G), \dots , S_{N-1}(G)$, where $S_k(G)$ denotes the number of $k$-stars in the network, $k \in \{1, \dots , N-1\}$.
 For simplicity, let us define
-$$S_k(A):=\Gamma_{star(k)}(A) $$
-\textit{define kstar}.
-Note that in every network $S_1(A)=\Gamma_{edges}(A)$, i.e., $S_1(A)$ is equal to the number of edges in the network.
+$$S_k(G):=\Gamma_{star(k)}(G) $$
+where
+$$\Gamma_{star(k)}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i,j,k}^N g_{ij}g_{ik}$$
+Note that in every network $S_1(G)=\Gamma_{edges}(G)$, i.e., $S_1(G)$ is equal to the number of edges in the network.
 On this basis, Snijders introduces the \textit{alternating k-star statistics}
@@
 \begin{split}
-\mathfrak{S}(A,\lambda):= & \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(A)\\= & ~   S_2(A)-\cfrac{S_3(A)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(A)}{\lambda^{N-3}}
+\mathfrak{S}(G,\lambda):= & \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(G)\\= & ~   S_2(G)-\cfrac{S_3(G)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(G)}{\lambda^{N-3}}
 \end{split}
@@
 \begin{equation}
-\mathfrak{S}(A,\lambda)= \lambda\Biggr(\lambda\sum_{j=1}^{N-1}\Biggr( 1-\biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(A) + 2S_1(A)\Biggl)
+\mathfrak{S}(G,\lambda)= \lambda\Biggr(\lambda\sum_{j=1}^{N-1}\Biggr( 1-\biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G) + 2S_1(G)\Biggl)
 \label{alto}
 \end{equation}
-where $D_j(A):= \Gamma_{deg(j)}(A)$ is the number of nodes with and degree of $j$. 
-In the next step, we define the \textit{geometrically weighted degree} statistic as the first summand of (\ref{alto})
+where $D_j(G):= \Gamma_{deg(j)}(G)$ is the number of nodes with a degree of $j$. 
+In the next step, we define the \textit{geometrically weighted degree} (gwd) statistic as the first summand of (\ref{alto})
 \begin{equation}
-\Gamma_{gwd}(A,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(A)
+\Gamma_{gwd}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G)
 \end{equation}
 At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
-
-As nodal attribute we implement the partisanship of the bill sponsor.
-In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
-The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
-The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on.
-Then, the statistics for a nodal covariate is defined as
-$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
+In a very similar way we can also define gwesp, as applied in chapter IV, as
+$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
+where 
+$$
+\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr)
+$$
+This means this statistic counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors.\\
+\noindent As nodal attribute we implement the partisanship of the bill sponsor.\\
 We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
@@
 This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
-
+\\[0.3cm]
 \begin{figure}[!t]
 \centering
-\includegraphics[width=3in]{rel_time}
+\includegraphics[width=3.5in, height=2.7in]{rel_time}
 \caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
@@
 \end{figure}
-
 Figure \ref{comptime} demonstrates that all three networks only require two cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
@@
 The plots on the right side visualize the empirical density function of the respective statistic, based on the simulated networks (Hunter and Handcock \cite{Hunter.2006}).
-For a non-degenerated model the empirical density function should be symmetrical around zero for every included centered statistic, since this corresponds with the expected value of a centered statistic (compare equation (\ref{assumption})).
+For a non-degenerated model the empirical density function should be approximately symmetrical around zero for every included centered statistic, since this corresponds with the expected value of a centered statistic (compare equation (\ref{assumption})).
 Otherwise, the values of the simulated networks systematically differ from the corresponding statistics in the observed network, making it unreasonable to assume that the simulated networks originate from the same distribution as the observed network.
@@
 \centering
-\includegraphics[width=3in]{MPLE_histogram_traceplot_cosponsorship}
+\includegraphics[width=3.5in, height=2.7in]{MPLE_histogram_traceplot_cosponsorship}
 \caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.

### 2017-07-24 Cosponsorship/Relative Time plot.R
Relative time code and new plot
newest version of the code for the relative time plot

@@ -26,32 +26,37 @@ set.seed(1)
 for(i in 1:20){
   print(i)
   b<- system.time(ergm(faux.mesa.high~edges+nodefactor("Sex")+gwesp(0.25,fixed=TRUE), 
-                       eval.loglik = FALSE, control = control.ergm(MCMC.interval = 2000,MCMC.samplesize=1000)))
+                       eval.loglik = FALSE, control = control.ergm(MCMC.interval = 2000,MCMC.samplesize=2000)))
   time.mcmle.mesa[i]<- b[3]                  
 }
 
 median(time.mcmle.mesa)
 
-
-time.mple<- median(time.mple.mesa)
-time.mcmle<- median(time.mcmle.mesa)
-time.sim<- median(time.mcmle.mesa)/5 #takes 2 iterations, MCMC samplesize is quadrupled in last iteration
+time.mple.mesa.med<- median(time.mple.mesa)
+time.mcmle.mesa.med<- median(time.mcmle.mesa)
+time.sim.mesa.med<- median(time.mcmle.mesa)/5 #takes 2 iterations, MCMC samplesize is quadrupled in last iteration
 
 # calculate bootMPLE time for 500 iterations and different number of cores (1, 5, 10, 25, 50, 100, 250, 500)
 
 boot.mple.mesa<-c()
-boot.mple.mesa[1]<- 500*time.mple/1 + time.sim
-boot.mple.mesa[2]<- 500*time.mple/5 + time.sim
-boot.mple.mesa[3]<- 500*time.mple/10 + time.sim
-boot.mple.mesa[4]<- 500*time.mple/25 + time.sim
-boot.mple.mesa[5]<- 500*time.mple/50 + time.sim
-boot.mple.mesa[6]<- 500*time.mple/100 + time.sim
-boot.mple.mesa[7]<- 500*time.mple/250 + time.sim
-boot.mple.mesa[8]<- 500*time.mple/500 + time.sim
+boot.mple.mesa[1]<- 500*time.mple.mesa.med/1 + time.sim.mesa.med
+boot.mple.mesa[2]<- 500*time.mple.mesa.med/2 + time.sim.mesa.med
+boot.mple.mesa[3]<- 500*time.mple.mesa.med/3 + time.sim.mesa.med
+boot.mple.mesa[4]<- 500*time.mple.mesa.med/4 + time.sim.mesa.med
+boot.mple.mesa[5]<- 500*time.mple.mesa.med/5 + time.sim.mesa.med
[…]

### 2017-07-25 TeX/conference_paper.tex
Relative Time ratio
Added a sentence about the relative time ratio: The ratio of boot MPLE to MCMLE goes to 0.2, meaning that the boot MPLE can be about 5 times faster

@@
 \end{figure}
-Figure \ref{comptime} demonstrates that all three networks only require two cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
+Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
+If more computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at about $0.2$, meaning that the computing time can be quintupled using the bootstrapped MPLE.
 This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\

### 2017-07-26 TeX/conference_paper.tex
revising cosponsorship section

@@
 \end{table*}
-\noindent To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on Cosponsorships in the U.S.
-Senate and U.S.
-House of Representatives for the 93rd to 108th Congresses developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
-The cosponsorship network consists of 2635 nodes (in this case senators) and contains information of over 280,000 pieces of legislation proposed in the U.S.
-House and Senate for the time period 1973 - 2004.
+\noindent To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on cosponsorship of bills in the U.S.
+House of Representatives for the 108th Congress (2003--2004), developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
+The cosponsorship network consists of 2635 nodes (i.e., senators) and contains information of over [CHRISTIAN, HOW MANY BILLS IN THE 108th?] pieces of legislation considered by the Senate during the 108th Congress.
 In this undirected network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
-This means that an edge emerges between senators $i$ and $j$ if senator $i$ cosponsors legislation sponsored by senator $j$.
-The endogenous statistics we include into this model are the number of edges and the alternating k-star statistic as it was introduced by Snijders et al.
+Specifically, we include an edge between senators $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ sponsored each bill is greater than a random uniform draw.
+This results in an undirected network with [CHRISTIAN, HOW MANY?] edges.
+
+We build an ERGM specification that extends the work of  Zhang et al.
+(2008) \cite{zhang2008community} in exploring the structure of cosponsorship ties.
+They find that congressional cosponsorship is primarily characterized by intra-party ties---among Republicans and among Democrats, but few cross-party ties.
+We test for this party-based clustering (i.e., homophily) in our ERGM.
+This is done through a term that accounts for the party of senators.
+The party homophily term is defined as $$ h(G,X) = \sum_{i < j} G_{ij}X_{ij},$$ where $G$ is the network adjacency matrix, and X is an indicator matrix that assumes the value 0 if $i$ and $j$ are from different political parties and 1 if they are from the same party.
+$h(G,X)$ measures the number of intra-party ties in the network.
+A positive parameter value for this statistic indicates that ties tend to be formed between nodes with the same political party.
+
+We extend the homophily-based model to account for a network dynamic that is commonly found in the study of networks---that of popularity or preferential attachment \cite{barabasi1999emergence}.
+Preferential attachment is the tendency for new ties to be formed with nodes who already have many ties (i.e., popular nodes).
+The alternating k-star statistic was introduced by Snijders et al.
 \cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock (2006) \cite{Hunter.2006}.
+A positive parameter estimate associated with the alternating k-star statistic indicates that tie formation follows a form of preferential attachment \cite{SnijdersTomA.B..2006}.
 The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic.

### 2017-07-26 TeX/conference_paper.tex
editing abstract and application

@@
 MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant.
-However, as the size of the network increases, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 10,000.
-When the network is large, one option for estimation is maximum pseudolikelihood (MPLE).
-MPLE is considerably less accurate than MLE in small networks, but exhibits comparable mean squared error in large networks.
+However, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 1,000 nodes (i.e., one million potential ties) or greater.
+When the network is large, one option is maximum pseudolikelihood estimation (MPLE).
 The standard MPLE is simple and fast, but generally underestimates standard errors.
 We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals.
-Furthermore, bootstrapped MPLE has the advantage of being embarrassingly parallel.
-We compare the two different approaches by applying them on U.S.
-Cosponsorship Network Data.
+We find that bootstrapped MPLE can be run in 1/10th the time of MCMLE.
+We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.
 
@@
 \end{table*}
-\noindent To demonstrate the computational advantage of the bootstrapped MPLE to the commonly used MCMLE we apply both approaches to the data on cosponsorship of bills in the U.S.
+\noindent To illustrate the performance of MCMLE relative to that of the bootstrapped MPLE we apply both approaches to the data on cosponsorship of bills in the U.S.
 House of Representatives for the 108th Congress (2003--2004), developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
-The cosponsorship network consists of 2635 nodes (i.e., senators) and contains information of over [CHRISTIAN, HOW MANY BILLS IN THE 108th?] pieces of legislation considered by the Senate during the 108th Congress.
+The cosponsorship network consists of 2,635 nodes, which we define as pieces of legislation (i.e., bills), considered by the Senate during the 108th Congress.
+Note that this formulation of the cosponsorship network differs from past research, which has defined the nodes as the individual legislators.
+Because there are more bills than legislators, studying bills as nodes provides a more disaggregated look at the network than is offered through studying the network of legislators.
 In this undirected network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
-Specifically, we include an edge between senators $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ sponsored each bill is greater than a random uniform draw.
+Specifically, we include an edge between bills $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ were sponsored each legislator s greater than a random uniform draw.
 This results in an undirected network with [CHRISTIAN, HOW MANY?] edges.
@@
 This is done through a term that accounts for the party of senators.
-The party homophily term is defined as $$ h(G,X) = \sum_{i < j} G_{ij}X_{ij},$$ where $G$ is the network adjacency matrix, and X is an indicator matrix that assumes the value 0 if $i$ and $j$ are from different political parties and 1 if they are from the same party.
+The party homophily term is defined as $$ h(G,X) = \sum_{i < j} G_{ij}X_{ij},$$ where $G$ is the network adjacency matrix, and X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
 $h(G,X)$ measures the number of intra-party ties in the network.
-A positive parameter value for this statistic indicates that ties tend to be formed between nodes with the same political party.
+A positive parameter value for this statistic indicates that ties tend to be formed between bills sponsored by the same political party.
 

### 2017-07-26 TeX/conference_paper.tex
revising intro

@@
 \section{The Exponential Random Graph Model}
-\noindent The \textit{exponential random graph model} (ERGM) is a probability model for directed or undirected binary networks.
-This means neither the weighting nor the temporal change of ties is considered in the model.
-In literature, ERGMs are sometimes also referred to as \textit{p-star} or $p^*$ models (see Wassermann and Pattison \cite{Wasserman.1996}, Robins et al.
-\cite{robins.pattison.2007}).
-In the following, we will introduce the ERGM for directed networks; however, undirected networks can be introduced very similar.\\[0.3cm]
+\noindent The \textit{exponential random graph model} (ERGM) is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
+They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
+The ERGM was first introduced by Holland and Leinhardt (1981) \cite{holland1981exponential}.
+However, due to the intractable normalizing constant in the likelihood function of the ERGM, it did not see widespread and complete use until the 2000s, following the development of algorithms and software for efficient simulation-based methods for working with ERGM \cite{snijders2002markov}.
+Training ERGM using simulation-based methods is computationally expensive, and can still be prohibitively burdensome with data on big networks.
+Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.
+
 The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$.

### 2017-07-27 TeX/conference_paper.tex
revising intro

@@
 \section{Introduction}
-\noindent The field of network science faces a double-edge sword when it comes to computational limitations to innovation.
-First, the networks under study are growing larger with each year.
-{\bf [CHRISTIAN, WOULD YOU CHECK OUT N NODES IN {\em SOCIAL NETWORKS}?]}.
-Second, analytical methods are growing more sophisticated, increasingly involving estimation and optimization, going beyond descriptive calculations {\bf [Could also classify articles based on whether a statistical model is used {\em SOCIAL NETWORKS}?]}.
-In order to avoid a limiting horizon in which state-of-the-art methods cannot be used with state-of-the-art ``big'' data, network methodologists need to consider all available options in easing the computational burden of tools for network analysis.
+\noindent The field of network science faces a double-edge sword when it comes to computationally intensive research.
+First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.
+Faraj et al.
+(2008, pp.
+19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques.
+This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.
+Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference.
+We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.
 
 \section{The Exponential Random Graph Model}
-\noindent The \textit{exponential random graph model} (ERGM) is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
+\noindent The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
 They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
@@
 
-The ERGM takes the adjacency matrix of an observed network $G^{obs}$ as the manifestation of a matrix-like random variable $Y$.
-This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{R}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$.
+The ERGM takes the adjacency matrix of an observed network $G^{obs}$, which is a matrix-valued random variable.
+This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{[0,1]}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$.
 $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected.
@@
 Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements.
-Therefore, for an already small number of actors the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
-For this reason calculating the MLE is either extremely time-consuming or with today's technology not achievable.
-As a consequence, literature usually makes use of Markov Chain Monte Carlo (MCMC) methods as we will intriduce in the next chapter.\\
+For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
+For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable.
+As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\
 With the definition of $\mathcal{G}(N)$ we define 

### 2017-07-27 TeX/conference_paper.tex
updating ERGM section

@@
 As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\
-With the definition of $\mathcal{G}(N)$ we define 
-$$Y: \Omega \to \mathcal{G}(N)~~,~~\omega \mapsto (Y_{ij}(\omega))_{i,j=1,\dots,N}$$
-as a matrix-like random variable. As the probability function from $Y$ to $\mathcal{G}(N)$ we define the ERGM as
+
+The probability function for the ERGM is defined as
 \begin{equation}
-\mathbb{P}_{\theta}(Y=G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
+\mathbb{P}_{\theta}(G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
 \label{ERGM}
@@
 %
-As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
-A key role when modeling an ERGM is played by the function $\Gamma(\cdot)$.
-The decision about which network statistics are incorporated into the model affects the model significantly (see Handcock \cite{Handcock.2003}).
+As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}) .
+The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al.
+(2006) \cite{SnijdersTomA.B..2006} for a detailed discussion).
+The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.
 
 \section{Estimation}
-\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology. Therefore, the question arises how one can estimate the parameter vector $\theta$?
-A first idea could be the following: One can assume that the \textit{dyads} are independent of each other, which means that the random variables $Y_{ij}$ inside the random matrix $Y$ are independent of each other.
-In this case, one can show that
+\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.
+Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible.
+The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}.
+Under maximum pseudolikelihood, the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
 %
 \begin{equation*}
-\text{logit}(\mathbb{P}_{\theta}(Y_{ij}=1))= \theta^T \cdot (\Delta G)_{ij}
+\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
 %\label{ILogit}
 \end{equation*}
+where $G_{-ij}$ is the adjacency matrix, excluding element $ij$.
+
 This corresponds with the \textit{logistic regression} approach, where the observations of the dependent variables are simply edge values of the observed adjacency matrix, and the observations of the covariate values are given as the scores of every single change statistic.
@@
 
+\section{Acknowledgements}
+
 % An example of a floating figure using the graphicx package.

### 2017-07-27 TeX/conference_paper.tex
revising estimation section

@@
 \end{equation*}
-where $G_{-ij}$ is the adjacency matrix, excluding element $ij$.
+where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
+\begin{equation*}
+\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
+%\label{ILogit}
+\end{equation*}
+where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}.
 

### 2017-07-27 TeX/conference_paper.tex
finished paper revisions

@@
 We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals.
-We find that bootstrapped MPLE can be run in 1/10th the time of MCMLE.
+We find that bootstrapped MPLE can be run in 1/5th the time of MCMLE.
 We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.
@@
 The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}.
-Under maximum pseudolikelihood, the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
+Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
 %
@@
 where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}.
+For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.
 
-This corresponds with the \textit{logistic regression} approach, where the observations of the dependent variables are simply edge values of the observed adjacency matrix, and the observations of the covariate values are given as the scores of every single change statistic.
-Therefore, the estimation of $\theta$ can then be obtained as usual using straight forward maximum-likelihood estimation.
-The resulting likelihood function is of the following form:
-\begin{equation}
-\text{lik}(\theta)= \mathbb{P}_{\theta}(Y=G)= \prod_{i,j} \dfrac{ \exp \left(\theta^T \Delta(G))_{ij} \right)}{1+\exp \left(\theta^T \Delta(G))_{ij} \right)}
-\label{NaiveLik}
-\end{equation}
-The problem with this simple estimation procedure is that the assumed hypothesis of the independence of the dyads turns out to be erroneous in many cases (see van Duijn et al.
+Despite the computational efficiency underlying the implementation of MPLE,  existing methods for assessing uncertainty with respect to the MPLE perform poorly (see van Duijn et al.
 \cite{vanDuijnetal2009}).
-This is a systematic problem: The presence of network data is inextricably connected with the presence of \textit{relational data}, which by definition should not be assumed to be independent of each other.
-If this dependency structure is deliberately ignored and equation (\ref{NaiveLik}) is used to estimate $\theta$, it results in a \textit{maximum pseudo-likelihood estimation} (MPLE).
-This technique tends to underestimate the standard error.
-However, Desmarais and Cranmer \cite{Desmarais.2012} argue that the pseudo-likelihood provides a consistent approximation of the maximum likelihood, meaning that the MPLE converges to the MLE as the size of the network increases.\\[0.3cm]
-There are several techniques to circumvent estimators, which underestimate the standard error of $\theta$.
-In the following, we will introduce a technique based on \textit{Markov Chain Monte Carlo (MCMC)} and maximum-likelihood methods.\\[0.3cm]
-The more rigorous technique is to estimate the parameters directly with the log-likelihood function derived from (\ref{ERGM}), which has the following form:
-%
-\begin{equation}
-\text{loglik}(\theta)=\theta^T \cdot \Gamma(G)-\log(c(\theta))
-\label{loglik}
-\end{equation}
-%
-where $G$ is the observed network. For the vector of network statistics, one can assume without loss of generality
-%
-\begin{equation}
-\Gamma(G)=0 \label{assumption}
-\end{equation}
-%
-This means that centering the vector of network statistics does not affect the probability function of the network variable $Y$.
-Therefore, in context of the likelihood function (\ref{loglik}) the vector of statistics can always be assumed to be centered on the observed network.\\
-Due to assumption (\ref{assumption}), one gets from (\ref{loglik}) the simplified log-likelihood function
-%
-\begin{equation}
-\text{loglik}(\theta)= -\log(c(\theta))
-\label{vloglik}
-\end{equation}
-%
-The problem resulting from estimating the parameters with (\ref{loglik}) is that the term
-%
-$$c(\theta):= \sum_{G^* \in \mathcal{G}(N_V)} \exp(\theta^T \cdot \Gamma(G^*))$$ 
-%
-which sums up the weighted network statistics of all possible networks of $N$ nodes, has to be evaluated.
-Even for networks with small numbers of nodes this presents an enormous computational obstacle, and the necessary calculations for larger networks cannot currently be completed in any reasonable timeframe.
-As a result, for sufficiently large networks it is not possible to estimate the parameters directly with the likelihood function.
-\\[0.3cm]
-An expedient for this limitation is based on the following consideration: Fix any vector of parameters $\theta_0 \in \Theta$ from the underlying parameter range $\Theta$ and compute for $\theta \in \Theta$ the expected value.
-Then, one can show that
-%
-\begin{equation*}
-\mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]=\frac{c(\theta)}{c(\theta_0)}
-\end{equation*}
-%
-This equation offers the following possibility: If one draws $L$ random networks $G_1, \dots ,G_L$ out of a distribution $\mathbb{P}_{\theta_0}$ appropriately, one gets with the \textit{law of big numbers} the following relation:
-%
-\begin{equation}\label{konver}
-\begin{split}
-\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right) &
-~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~\\ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]& = \frac{c(\theta)}{c(\theta_0)}
-\end{split}
-\end{equation}
-%
-For a big enough number, $L$, of random networks, the following approximation is reasonable:
-%
-\begin{equation}
-\frac{c(\theta)}{c(\theta_0)} \approx \frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
-\label{approx}
-\end{equation}
-%
-One can now use equation (\ref{approx}) to determine an approximation of the log-likelihood function (\ref{vloglik}):
+Estimating the uncertainty in parameter estimates (e.g., standard errors, confidence intervals), is a critical step in using the results from a statistical model.
+Estimates of uncertainty are used to test hypotheses about parameters, estimate variance (i.e., risk) in model predictions, and estimate effect sizes.
+Another limitation of MPLE regards the assessment of model fit.
+With ERGM, model fit is conventionally evaluated by comparing the structure of the observed network to networks simulated from the estimated model \cite{Hunter.2008}.
+This comparison can include diagnosing the model for a particular form of poor fit that arises with ERGM---model degeneracy \cite{schweinberger2011instability}.
+Degeneracy is a condition in which the model places nearly all of the probability mass on the completely empty or completely full network.
+Since MPLE does not require simulating networks, researchers can derive and report results without checking the fit of the model or checking for degeneracy.
+Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.
+
+The contemporary conventional approach to estimating $\theta$, introduced by Snijders (2002) \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\[0.3cm]
+This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}).
+The log-likelihood function is not evaluated directly, rather, the log ratio of the likelihood under a proposed value of the parameters $\theta$, and an initial value of the parameters $\theta_0$, is approximated using $L$ networks simulated from the ERGM with parameter values $\theta_0$.
+The approximation, detailed in Snijders (2002) \cite{snijders2002markov} is given by
 %
@@
                                              &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
-                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
+                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
 \end{eqnarray*}
-%
 By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
@@
 This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
-As a result, the approximate maximum-likelihood estimator for the parameters can be computed.\\[0.3cm]
-As pleasant as this may sound, the immediate question arises: How can a sufficient number of suitable drawings $G_1,...,G_L$ be taken from the distribution $\mathbb{P}_{\theta_0}$? \\
-For this purpose, the \textit{Markov Chain Monte Carlo (MCMC)} methods can be used.
-This approach does not deliberately ignore the dependency structures inside the network.
-Furthermore, the Markov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
-No matter which kind of MCMC algorithm is used, the basic idea is the following:
-\\One constructs a \textit{Markov chain} $(G_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{G}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$.
-One can show that every single realization (or \textit{trajectory})
-of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}).
-As a result, sub-sequences of $(G_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
-Snijders \cite{Snijders.2002} provides a Metropolis Hastings algorithm on how to simulate networks:
+
+MCMC methods are used to simulate the $L$ networks.
+As we demonstrate below, the MCMLE grows more accurate as $L$ increases.
+Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
+Snijders (2002) \cite{Snijders.2002} provides a Metropolis Hastings algorithm to simulate networks:
 Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
@@
 \vspace{0.3cm}
-\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$.
+\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
 Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
@@
 \section{Efficiency of MPLE and MCMLE}
-\noindent Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy.
-Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods.
-As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
-This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\[0.3cm]
+\noindent  As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
+This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time.
 At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework.
@@
 instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.
-\\[0.3cm]%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes.
-\\
-%A major advantage of the MCMLE method is, as shown by van Duijn et al.
-\cite{vanDuijnetal2009}, that the MCMLE is substantially more efficient than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE.
-Nevertheless, as shown by
-Next, we are going to compare the efficiency of the MCMLE compared to the MPLE.
-Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough.
-It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases.
-This means that for very large networks it becomes difficult to determine the number of simulated networks required in order for the MCMLE to outperform the MPLE.
-In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
-\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
+
+To illustrate the relative efficiency of MPLE and MCMLE Next, we run a simulation study.
+Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood in MCMLE is not large enough.
+It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network (i.e., the number of nodes in the network) increases.
+This means that, for very large networks, it becomes difficult to determine the number of simulated networks required for the MCMLE to outperform the MPLE.
+In other words, the larger the network of the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\[0.3cm]
+\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
 The data for both networks originates from Resnick et al.
 \cite{Resnicketal1997}.\\[0.3cm]
-For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
-Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.
-The number of edges in these unidrected networks are defined as
+For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
+Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
+The number of edges is defined as
 $$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
-In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$.
+In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$.
 The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
 The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on.
-Then, the statistics for a nodal covariate is defined as
+Then, the statistic for a nodal covariate is defined as
 $$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-A precise description of the gwesp statistic will be given in the chapter VI.\\[0.3cm]
+The GWESP statistic is given as
+$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
+where 
+$$
+\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
+$$
+$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. This statistic is used to model the tendency towards triangles and clustering in a network.\\
 We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
-For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10.000$ simulated networks used in the likelihood approximation.
-Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance.
+For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10,000$ simulated networks used in the likelihood approximation.
+Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both the bias and the variance.
 Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as
@@
 \noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
-While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE.
-These results propose especially for very large networks (e.g.
-social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method.
-The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extend the mere computation time.
+While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks, the larger Faux Magnolia High network requires a sample size of at least 1,500 networks for the MCMLE to surpass the MPLE.
+For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive, MCMLE method, may be prohibitive.
 
 \section{Bootstrapped MPLE}
+
 \noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
-Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough.
-The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al.
+Moreover, the MPLE is able to outperform the MCMLE if the sample size used in MCMLE is not large enough.
+The main reason why the MCMLE is still widely preferred is that, in contrast to the MPLE, it does not underestimate the standard errors (van Duijn et al.
 \cite{vanDuijnetal2009}).
-By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic.
-For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
+By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (G)$ is the sufficient statistic.
+For exponential family distributions, it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
 The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals.
 van Duijn et al.
-show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%.
+show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the nominal 95\%.
 In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix.
 \\[0.3cm]
-\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
-Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
+\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}.
+Bootstrap resampling refers to constructing a sampling distribution for the parameter estimate by resampling the data with replacement, and re-estimating the model on the resampled data.
+Under non-parametric bootstrap resampling, the data are resampled directly from the dataset.
+Under the parametric boostrap, the data is resampled from the estimated model.
+The idea of using boostrap resampling with MPLE for ERGM was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
+Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE.
 \\[0.3cm]
-\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
+\noindent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap.
+The non-parametric bootstrap cannot be applied when there is just a single network under study.
+
+For the case in which there is just a single network to be studied, which is indeed the most common case in the literature, we propose the use of a parametric bootstrap.
+Under the parametric bootstrap, the sampling distribution of the MPLE is derived by re-estimating the MPLE on a sample of networks simulated from the MPLE estimated on the observed network.
+We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
+
 For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values.
-We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$.
+We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $G$.
 For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
 In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values.
-For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals.
+For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain  95\% bootstrap confidence intervals.
 Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
@@
 Similar to the results of van Duijn et al.
-\cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
+\cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, confirming that the MPLE underestimates the variance of its estimates.
 Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates.
 We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.
-It is especially worthwile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.\\
+It is especially worthwhile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.\\
 \noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
-Furthermore, this method also adopts the advantage of the MPLE of being consistent.
-However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provides a computational benefit, especially if the size of the examined network is extremely large.
+In this simulation study we demonstrate that the parametric bootstrap can be used in combination with MPLE to provide a method of estimation that is both computationally efficient and provides valid estimates of model uncertainty.
 
@@
 Alternating k-star                              & 0.124                            & 0.064                             &     0.108                             & 0.006                                   & -0.011 & 0.2379  \\ \hline
-\end{tabular}
+\end{tabular}\vspace{.2cm}
 \caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
@@
 We test for this party-based clustering (i.e., homophily) in our ERGM.
-This is done through a term that accounts for the party of senators.
-The party homophily term is defined as $$ h(G,X) = \sum_{i < j} G_{ij}X_{ij},$$ where $G$ is the network adjacency matrix, and X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
+This is done through a term that accounts for the party of the senators who sponsored the two bills in the pair.
+The party homophily term is defined as $$ \Gamma(G,X) = \sum_{i < j} g_{ij}x_{ij},$$ where $G$ is X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
 $h(G,X)$ measures the number of intra-party ties in the network.
@@
 A positive parameter estimate associated with the alternating k-star statistic indicates that tie formation follows a form of preferential attachment \cite{SnijdersTomA.B..2006}.
+This could arise in a network of bill-to-bill ties if the majority party in power was particularly disciplined at rallying its partisans to pile on to the bills that its members proposes, thus producing a large set of very similar bills.
 The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic.
@@
 At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
-In a very similar way we can also define gwesp, as applied in chapter IV, as
-$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
-where 
-$$
-\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr)
-$$
-This means this statistic counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors.\\
-\noindent As nodal attribute we implement the partisanship of the bill sponsor.\\
-We estimate the coefficient of this matrix using both techniques, the MCMLE and the bootstrapped MPLE.
+
+We estimate the ERGM using MCMLE and the bootstrapped MPLE.
 The MCMLE requires a sample size of at least $1000$ networks to converge.
 The bootstrapped MPLE was estimated by using 500 simulated networks.
-As we described in the chapter {\it Estimation}, only one edge at a time is changed when simulating networks.
-For better comparison, we chose the same Burn-In ($300.000$ MH-steps) and the same number of iterations ($30.000$ MH-steps) for sampling networks.
+As we described in the section {\it Estimation}, only one edge at a time is changed when simulating networks.
+For better comparison, we chose the same Burn-In ($300,000$ MH-steps) and the same number of iterations ($30,000$ MH-steps) for sampling networks.
 The results can be found in table \ref{co_results}.\\
-It is interesting to see that the standard error calculated by the logistic regression approach is constantly smaller than the standard error of the MCMLE, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
-Furthermore, we can observe that the coefficients for the number of edges and the nodal attribute are very similar, but the estimates for the alternating k-star statistic do differ \textit{(REASON FOR altkstar?)}.
-The bootstrapped MPLE bounds, however, include each estimate for both methods.
-\\
+It is interesting to see that the standard error calculated by the logistic regression approach is much smaller than the standard error of the MCMLE for the alternating k-star statistic, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
+The MPLE estimate is equivalent to the logistic regression estimate, but the bootstrap confidence intervals, especially for the alternating k-star statistic, are much wider than would be calculated using the logistic regression standard errors.
+[CHRISTIAN, PLEASE CHECK THE SPONSOR PARTY LOGISTIC ESTIMATES.
+ARE THEY REALLY EXACTLY THE SAME AS THE MCMLE]  An estimate is generally considered statistically different from zero (i.e., statistically significant) if the confidence interval does not contain zero, or if the ratio of the estimate to the standard error exceeds 1.96 in magnitude.
+This cosponsorship network example perfectly illustrates the inferential problems that can arise with the conventional logistic regression standard errors when using MPLE.
+All of the parameter estimates are statistically significant according to the logistic regression estimates.
+However, the alternating k-star statistic is not significant according to either the MCMLE or the bootstrapped MPLE.
 
-\noindent The bootstrapped MPLE is not only simple and fast, it is also embarrassingly parallel, meaning that there is no problem separating the computation into several tasks.
-In other words, by using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
+\section{Parallel Computing with MPLE}
+
+\noindent The bootstrapped MPLE is not only simple and fast, it is highly parallel.
+Once the networks on which to estimate the bootstrap replicates are simulated, each re-estimate can be run in parallel.
+By using multiple cores, the computing time for estimating bootstrapped MPLE confidence intervals can be reduced substantially.
 Figure \ref{comptime} illustrates the relative computing time of the bootstrapped MPLE using 500 simulated networks and the MCMLE for the three networks Faux Mesa High (205 nodes), Faux Magnolia High (1461 nodes) and Cosponsorship (2635 nodes) for an increasing number of computing cores.
@@
 The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit.
+[CHRISTIAN, PLEASE QUOTE THE ACTUAL COMPUTE TIME, IN SECONDS, FOR EACH NETWORK USING MCMLE].
 We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:
@@
 Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
-If more computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at about $0.2$, meaning that the computing time can be quintupled using the bootstrapped MPLE.
+If more computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at about $0.2$ [CHRISTIAN, WOULD YOU LIST THE ACTUAL MINIMUM RATIO FOR EACH NETWORK], meaning that the computing time can be quintupled using the bootstrapped MPLE.
 This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\
 Of course the actual computing time for a network always depends on the statistics that are included in the network, but in general larger networks require longer computation times, since a larger MCMC sample size is required and more MC steps are necessary to simulate a new network that does not overly depend on the previous sample.
-This makes the bootstrapped MPLE a beneficial alternative especially for very large networks.
+This makes the bootstrapped MPLE a beneficial alternative, especially for very large networks.
 \\[0.3cm]
-One of the major disadvantages of MPLE over MCMLE is that it cannot assess {\it degeneracy}.
-When fitting an ERGM one usually has to deal with the problem of unreliable approximatively likelihood estimates for the model's parameters.
-The reason why degeneracy occurs is that the stochastic process generated by the MCMC-algorithm does not necessarily hold through the model's defined distribution of the random variable $Y$ as stationary distribution (see Handcock \cite{Handcock.2003} and Rinaldo et al.
-\cite{Rinaldo2009}).
-The bootstrapped MPLE, however, allows assessing degenerate models as well.
+One of the major disadvantages of MPLE over MCMLE is that degeneracy is not assessed while the model is being estimated.
+The bootstrapped MPLE, however, allows assessing degenerate models as well since the method requires simulating from the estimated parameters.
 In order to verify whether a model is degenerate or not, one can take a look at density and trace plots as visualized in figure \ref{diagnostics}.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-
+[CHRISTIAN, WOULD YOU GO THROUGH THE ENTIRE MANUSCRIPT AND MAKE SURE THAT THE SPACING BETWEEN PARAGRAPHS IS CONSISTENT. I THINK I MESSED UP SOME BY USING INDENTS].
 \begin{figure}[!t]
@@
 \section{Conclusion}
-\noindent In the past years the ERGM grew in popularity in many different fields and with this popularity also the size of the considered networks rose.
-As the size of the network increases, it becomes unclear how many simulated networks are necessary for the conventional MCMLE method in order to provide reliable results.
-Furthermore, as an increasing number of simulated networks is necessary for the MCMLE, the computing time rapidly goes sky high.
+\noindent In the past years the ERGM grew in popularity in many different fields as a flexible and powerful means of building probabilistic models for networks.
+With this popularity the size of the considered networks has grown.
+As the size of the network increases, it becomes unclear how many simulated networks are necessary for the conventional MCMLE method to perform better than the MPLE.
+Furthermore, as an increasing number of simulated networks is necessary for the MCMLE, the computing time rapidly grows.
 For this reason it is essential to develop different methods that provide faster estimation than the MCMLE, but still lead to reliable results.\\
-In this paper we introduced the bootstrapped MPLE as an alternative estimation method for ERGMs and compared the performance to the commonly applied MCMLE.
-Based on a simulation study we first pointed out that the larger the size of a network is the larger the MCMC sample size has to be chosen in order for the MCMLE to outperform the fast and simple MPLE.
-However, the big disadvantage of the MPLE is that even though it is an approximately exact estimator, it underestimates the standard error.
-For this reason, we propose a parametric bootstrap method to attune the standard error.
-On the basis of another simulation study on two different networks we demonstrate that the bootstrapped MPLE covers the true coefficients just as well as the MCMLE, while the simple MPLE performs clearly poorer.
-This means that the bootstrapped MPLE combines the advantages of both methods, the MPLE and the MCMLE, because it is still simple, fast and provides approximately exact results, but also does not underestimate the standard error.
-We conclude that the bootstrapped MPLE should be regarded as a propitious alternative to the MCMLE.
-It also has the advantage of being embarrassingly parallel, which leads to a rapid speed-up of the calculation if multiple computing cores are used.
+In this paper we introduced the bootstrapped MPLE as an alternative method of statistical inference for ERGMs and compared the performance to the commonly applied MCMLE.
+Based on a simulation study we demonstrated that the larger the size of a network is the larger the MCMC sample size has to be in order for the MCMLE to outperform the fast and simple MPLE.
+However, the big disadvantage of the MPLE is that, even though it is an approximately exact estimator, it underestimates the standard error.
+For this reason, we propose a parametric bootstrap method of evaluating model uncertainty.
+On the basis of another simulation study on two different networks, we demonstrate that the bootstrapped MPLE covers the true coefficients just as well as the MCMLE, while the simple MPLE performs clearly poorer.
+This means that the bootstrapped MPLE combines the advantages of both methods, the MPLE and the MCMLE, because it is still simple and fast, and provides approximately exact results, but also accurately estimates model uncertainty.
+We conclude that the bootstrapped MPLE should be regarded as an alternative to the MCMLE.
+It also has the advantage of being parallel, which leads to a rapid speed-up of the calculation if multiple computing cores are used.
 
 \section{Acknowledgements}
+
+This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219.
+Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsors.
 

### 2017-07-28 TeX/conference_paper.tex
finished revisions
finalized paper

@@
 \begin{abstract}
-\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data.
+With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data.
 ERGM parameter estimation requires the approximation of an intractable normalizing constant.
@@
 \section{Introduction}
-\noindent The field of network science faces a double-edge sword when it comes to computationally intensive research.
+The field of network science faces a double-edge sword when it comes to computationally intensive research.
 First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.
@@
 \section{The Exponential Random Graph Model}
-\noindent The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
+The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
 They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
@@
 Training ERGM using simulation-based methods is computationally expensive, and can still be prohibitively burdensome with data on big networks.
-Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.
-
-The ERGM takes the adjacency matrix of an observed network $G^{obs}$, which is a matrix-valued random variable.
-This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{[0,1]}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$.
+Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.\\
+\indent The ERGM takes the adjacency matrix of an observed network $G^{obs}$, which is a matrix-valued random variable.
+This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \{0,1\}^{(N \times N)}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$.
 $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected.
@@
 Furthermore, define
-$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
+$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
 as the set of all possible networks on $N$ nodes without loops.
@@
 As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\
-
 The probability function for the ERGM is defined as
@@
 %
-As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}) .
+As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
 The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al.
-(2006) \cite{SnijdersTomA.B..2006} for a detailed discussion).
+\cite{SnijdersTomA.B..2006} for a detailed discussion).
 The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.
@@
 \section{Estimation}
-\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.
+As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.
 Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible.
@@
 
-The contemporary conventional approach to estimating $\theta$, introduced by Snijders (2002) \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\[0.3cm]
+The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\
 This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}).
@@
 Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
-Snijders (2002) \cite{Snijders.2002} provides a Metropolis Hastings algorithm to simulate networks:
+Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:
 Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
@@
 \vspace{0.3cm}
-\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
+The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
 Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
@@
 \section{Efficiency of MPLE and MCMLE}
-\noindent  As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
+As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
 This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time.
@@
 
-To illustrate the relative efficiency of MPLE and MCMLE Next, we run a simulation study.
+To illustrate the relative efficiency of MPLE and MCMLE we run a simulation study.
 Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood in MCMLE is not large enough.
@@
 This means that, for very large networks, it becomes difficult to determine the number of simulated networks required for the MCMLE to outperform the MPLE.
-In other words, the larger the network of the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\[0.3cm]
-\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
+In other words, the larger the network, the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\
+\indent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students.
 The data for both networks originates from Resnick et al.
-\cite{Resnicketal1997}.\\[0.3cm]
-For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
+\cite{Resnicketal1997}.\\
+\indent For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
 Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
@@
 $\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. This statistic is used to model the tendency towards triangles and clustering in a network.\\
-We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
+\indent We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
 For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10,000$ simulated networks used in the likelihood approximation.
@@
 In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE.
-As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\[0.3cm]
-Figure \ref{rmse} visualizes the results of the simulation study.
+As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\
+\indent Figure \ref{rmse} visualizes the results of the simulation study.
 The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.
@@
 \end{figure}
-\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
+The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
 While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks, the larger Faux Magnolia High network requires a sample size of at least 1,500 networks for the MCMLE to surpass the MPLE.
-For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive, MCMLE method, may be prohibitive.
+For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive and potentially prohibitive MCMLE method.
 
 \section{Bootstrapped MPLE}
-
-\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
+As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases.
 Moreover, the MPLE is able to outperform the MCMLE if the sample size used in MCMLE is not large enough.
@@
 In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix.
-\\[0.3cm]
-\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}.
+\\
+\indent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}.
 Bootstrap resampling refers to constructing a sampling distribution for the parameter estimate by resampling the data with replacement, and re-estimating the model on the resampled data.
@@
 Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE.
-\\[0.3cm]
-\noindent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap.
+\\
+\indent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap.
 The non-parametric bootstrap cannot be applied when there is just a single network under study.
@@
 \end{figure}
-\noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
+Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
 The dashed line is set at $0.95$ and represents the optimal value.
@@
 It is especially worthwhile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.\\
-\noindent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
+\indent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
 In this simulation study we demonstrate that the parametric bootstrap can be used in combination with MPLE to provide a method of estimation that is both computationally efficient and provides valid estimates of model uncertainty.
@@
 \end{table*}
-\noindent To illustrate the performance of MCMLE relative to that of the bootstrapped MPLE we apply both approaches to the data on cosponsorship of bills in the U.S.
+To illustrate the performance of MCMLE relative to that of the bootstrapped MPLE we apply both approaches to the data on cosponsorship of bills in the U.S.
 House of Representatives for the 108th Congress (2003--2004), developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
@@
 In this undirected network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
-Specifically, we include an edge between bills $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ were sponsored each legislator s greater than a random uniform draw.
-This results in an undirected network with [CHRISTIAN, HOW MANY?] edges.
+Specifically, we include an edge between bills $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ were sponsored each legislator is greater than a random uniform draw.
+This results in an undirected network with $28060$ edges.
 
-We build an ERGM specification that extends the work of  Zhang et al.
-(2008) \cite{zhang2008community} in exploring the structure of cosponsorship ties.
+We build an ERGM specification that extends the work of  Zhang et al\cite{zhang2008community} in exploring the structure of cosponsorship ties.
 They find that congressional cosponsorship is primarily characterized by intra-party ties---among Republicans and among Democrats, but few cross-party ties.
@@
 This is done through a term that accounts for the party of the senators who sponsored the two bills in the pair.
-The party homophily term is defined as $$ \Gamma(G,X) = \sum_{i < j} g_{ij}x_{ij},$$ where $G$ is X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
-$h(G,X)$ measures the number of intra-party ties in the network.
+The party homophily term is defined as $$ \Gamma(G,X) = \sum_{i < j} g_{ij}x_{ij},$$ where X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
+$\Gamma(G,X)$ measures the number of intra-party ties in the network.
 A positive parameter value for this statistic indicates that ties tend to be formed between bills sponsored by the same political party.
@@
 The alternating k-star statistic was introduced by Snijders et al.
-\cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock (2006) \cite{Hunter.2006}.
+\cite{SnijdersTomA.B..2006} and modified by Hunter and Handcock  \cite{Hunter.2006}.
 A positive parameter estimate associated with the alternating k-star statistic indicates that tie formation follows a form of preferential attachment \cite{SnijdersTomA.B..2006}.
@@
 \end{equation}
-At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\[0.3cm]
-
-We estimate the ERGM using MCMLE and the bootstrapped MPLE.
+At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\
+\indent We estimate the ERGM using MCMLE and the bootstrapped MPLE.
 The MCMLE requires a sample size of at least $1000$ networks to converge.
@@
 The results can be found in table \ref{co_results}.\\
-It is interesting to see that the standard error calculated by the logistic regression approach is much smaller than the standard error of the MCMLE for the alternating k-star statistic, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
+\indent It is interesting to see that the standard error calculated by the logistic regression approach is much smaller than the standard error of the MCMLE for the alternating k-star statistic, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
 The MPLE estimate is equivalent to the logistic regression estimate, but the bootstrap confidence intervals, especially for the alternating k-star statistic, are much wider than would be calculated using the logistic regression standard errors.
-[CHRISTIAN, PLEASE CHECK THE SPONSOR PARTY LOGISTIC ESTIMATES.
-ARE THEY REALLY EXACTLY THE SAME AS THE MCMLE]  An estimate is generally considered statistically different from zero (i.e., statistically significant) if the confidence interval does not contain zero, or if the ratio of the estimate to the standard error exceeds 1.96 in magnitude.
+An estimate is generally considered statistically different from zero (i.e., statistically significant) if the confidence interval does not contain zero, or if the ratio of the estimate to the standard error exceeds 1.96 in magnitude.
 This cosponsorship network example perfectly illustrates the inferential problems that can arise with the conventional logistic regression standard errors when using MPLE.
@@
 \section{Parallel Computing with MPLE}
-
-\noindent The bootstrapped MPLE is not only simple and fast, it is highly parallel.
+The bootstrapped MPLE is not only simple and fast, it is highly parallel.
 Once the networks on which to estimate the bootstrap replicates are simulated, each re-estimate can be run in parallel.
@@
 The chosen sample sizes and MCMC steps are necessary to guarantee a good model fit.
-[CHRISTIAN, PLEASE QUOTE THE ACTUAL COMPUTE TIME, IN SECONDS, FOR EACH NETWORK USING MCMLE].
+The small network took $14$ seconds, the medium network took $123$ seconds and the large network took $986$ seconds to run.
 We define the simulation time of the bootstrapped MPLE as a function of the number of available computing cores x:
@@
 This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
-\\[0.3cm]
+\\
 \begin{figure}[!t]
@@
 Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
-If more computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at about $0.2$ [CHRISTIAN, WOULD YOU LIST THE ACTUAL MINIMUM RATIO FOR EACH NETWORK], meaning that the computing time can be quintupled using the bootstrapped MPLE.
+If exactly 500 computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at $0.20$ for the small and large network and $0.17$ for the medium network, meaning that the computing time can be quintupled using the bootstrapped MPLE.
 This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\
@@
 This makes the bootstrapped MPLE a beneficial alternative, especially for very large networks.
-\\[0.3cm]
-One of the major disadvantages of MPLE over MCMLE is that degeneracy is not assessed while the model is being estimated.
+\\
+\indent One of the major disadvantages of MPLE over MCMLE is that degeneracy is not assessed while the model is being estimated.
 The bootstrapped MPLE, however, allows assessing degenerate models as well since the method requires simulating from the estimated parameters.
@@
 The plots on the right side visualize the empirical density function of the respective statistic, based on the simulated networks (Hunter and Handcock \cite{Hunter.2006}).
-For a non-degenerated model the empirical density function should be approximately symmetrical around zero for every included centered statistic, since this corresponds with the expected value of a centered statistic (compare equation (\ref{assumption})).
+For a non-degenerated model the empirical density function should be approximately symmetrical around zero for every included centered statistic, since this corresponds with the expected value of a centered statistic.
 Otherwise, the values of the simulated networks systematically differ from the corresponding statistics in the observed network, making it unreasonable to assume that the simulated networks originate from the same distribution as the observed network.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-[CHRISTIAN, WOULD YOU GO THROUGH THE ENTIRE MANUSCRIPT AND MAKE SURE THAT THE SPACING BETWEEN PARAGRAPHS IS CONSISTENT. I THINK I MESSED UP SOME BY USING INDENTS].
 \begin{figure}[!t]
@@
 \section{Conclusion}
-\noindent In the past years the ERGM grew in popularity in many different fields as a flexible and powerful means of building probabilistic models for networks.
+In the past years the ERGM grew in popularity in many different fields as a flexible and powerful means of building probabilistic models for networks.
 With this popularity the size of the considered networks has grown.

### 2017-08-03 TeX/conference_paper.tex
added a reference

@@
 \section{Efficiency of MPLE and MCMLE}
-As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}).
+As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012,desmarais2010consistent}).
 This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time.

### 2017-10-05 TeX/complenet18.tex
Reduced CompleNet paper to 16 pages

--- 2017-08-03 TeX/conference_paper.tex
@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
+\documentclass{llncs}
 
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
+\usepackage{cite}
 
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
+\usepackage[pdftex]{graphicx}
 
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
-% Add the compsocconf option for Computer Society conferences.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-   \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
 \usepackage[cmex10]{amsmath}
 \usepackage{amssymb}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{dsfont}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
 
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
+\usepackage{array}
+\usepackage{algorithm2e}
 
-% *** ALIGNMENT PACKAGES ***
-%
-\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
 %
 \usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
 
@@
 \usepackage{multicol}
+%\setlength{\belowcaptionskip}{-15pt}
+%\usepackage[demo]{graphicx}% http://ctan.org/pkg/graphicx
+%\usepackage{setspace}% http://ctan.org/pkg/setspace
 
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+%
+\begin{document}
+%\setlength{\intextsep}{5pt}%
+%\setstretch{2}%
+%\setlength{\belowcaptionskip}{-\baselineskip}%\addtolength{\belowcaptionskip}{1.6ex}%
+%\setlength{\belowcaptionskip}{-20pt}
 
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
 \title{Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}
 
-% author names and affiliations
-% use a multiple column layout for up to two different
-% affiliations
+\author{Christian S. Schmid and Bruce A. Desmarais}
 
-\author{\IEEEauthorblockN{Christian S. Schmid}
-\IEEEauthorblockA{Department of Statistics\\
-The Pennsylvania State University\\
-State College, USA\\
-Email: schmid@psu.edu}
-\and
-\IEEEauthorblockN{Bruce A. Desmarais}
-\IEEEauthorblockA{Department of Political Science\\
-The Pennsylvania State University\\
-State College, USA\\
-Email:  bdesmarais@psu.edu}
-}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
+\institute{The Pennsylvania State University\\
+State College, USA}
 \maketitle
@@
 \end{abstract}
-
-\begin{IEEEkeywords}
-network, ERGM, parametric bootstrap, maximum pseudolikelihood
-
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
 
@@
 Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$.
-Furthermore, define
-$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
+Furthermore, define $\mathcal{G}(N)$
+%$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
 as the set of all possible networks on $N$ nodes without loops.
@@
 \cite{SnijdersTomA.B..2006} for a detailed discussion).
-The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.
+The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic}%,srivastava2011culture}
+, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011}%,lazer2010coevolution}
+, ecology \cite{dey2014individual}%,bouskila2015similarity}
+, and neuroscience \cite{simpson2011exponential}.%,simpson2012exponential}.
 
@@
 Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
-%
 \begin{equation*}
@@
 Since MPLE does not require simulating networks, researchers can derive and report results without checking the fit of the model or checking for degeneracy.
-Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.
-
-The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\
+Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.\\
+\indent The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.
 This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}).
@@
 %
-\begin{eqnarray*}
-&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
-                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
-                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
-                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
-\end{eqnarray*}
-By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
+\begin{eqnarray}
+\text{loglik}(\theta)-\text{loglik}(\theta_0)\approx - 
+\log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
+\end{eqnarray}
+
+%\begin{eqnarray*}
+%&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
+%                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
+%                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
+%                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
+%\end{eqnarray*}
+%By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
 %
-\begin{equation}
-s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
-\label{score}
-\end{equation}
+%\begin{equation}
+%s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
+%\label{score}
+%\end{equation}
 %
-This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
+%This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
 
-MCMC methods are used to simulate the $L$ networks.
+\noindent MCMC methods are used to simulate the $L$ networks.
 As we demonstrate below, the MCMLE grows more accurate as $L$ increases.
 Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
-Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:
-Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
-\begin{enumerate}
-\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
-\item Compute the value
-$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
-\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
-\begin{itemize}
-\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
-\item $Z=1$, define $G^{(k+1)}$ via
-$$g_{pq}^{(k+1)}=\begin{cases}
+Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:\\[0.2cm]
+
+\begin{algorithm}[H]
+ Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network)\\
+ \For{$k \in \{0,...,L-1\}$ }{
+  Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$\\
+  Compute
+$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$\\
+Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$.\\
+  \eIf{$Z=0$}{
+   let $G^{(k+1)} := G^{(k)}$
+   }{
+   define $G^{(k+1)}$ via
+$g_{pq}^{(k+1)}=\begin{cases}
 1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
 g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
-\end{cases}$$
-\end{itemize}
-\item Start at step 1 with $G^{(k+1)}$.
-\end{enumerate}
+\end{cases}$
+  }
+ }
+ \caption{Metropolis Hastings algorithm for Network Sampling}
+\end{algorithm}
+
+%Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
+%\begin{enumerate}
+%\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
+%\item Compute the value
+%$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
+%\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
+%\begin{itemize}
+%\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
+%\item $Z=1$, define $G^{(k+1)}$ via
+%$$g_{pq}^{(k+1)}=\begin{cases}
+%1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
+%g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
+%\end{cases}$$
+%\end{itemize}
+%\item Start at step 1 with $G^{(k+1)}$.
+%\end{enumerate}
 \vspace{0.3cm}
-The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
+\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
 Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
@@
 Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
-The number of edges is defined as
-$$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
-In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$.
+%The number of edges is defined as
+%$$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
+%In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$.
 The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
@@
 Then, the statistic for a nodal covariate is defined as
-$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-The GWESP statistic is given as
-$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
-where 
-$$
-\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
-$$
-$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. This statistic is used to model the tendency towards triangles and clustering in a network.\\
+%$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
+%The GWESP statistic is given as
+%$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
+%where 
+%$$
+%\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
+%$$
+%$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. 
+The gwesp statistic is used to model the tendency towards triangles and clustering in a network.\\
 \indent We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as
-$$RMSE = \sqrt{\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$$
+$RMSE = \sqrt{\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$
 implying that the smaller the RMSE, the more accurate is the estimator.
@@
 \centering
-\includegraphics[width=3.6in]{RMSE_10k}
+\includegraphics[width=3.5in]{RMSE_10k}
 \caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
@@
 This could arise in a network of bill-to-bill ties if the majority party in power was particularly disciplined at rallying its partisans to pile on to the bills that its members proposes, thus producing a large set of very similar bills.
-The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic.
+%The alternating k-star statistic adds one network statistic to the model equal to a weighted alternating sequence of k-star statistics with weight parameter $\lambda$ and is a way to include a networks entire degree distribution as a network statistic.
 In this model we fix the weight paramter $\lambda=0.4975$.
-Snijders et al.
+%Snijders et al.
 \cite{SnijdersTomA.B..2006} introduced an approach involving $k$-star statistics $S_1(G), \dots , S_{N-1}(G)$, where $S_k(G)$ denotes the number of $k$-stars in the network, $k \in \{1, \dots , N-1\}$.
 For simplicity, let us define
-$$S_k(G):=\Gamma_{star(k)}(G) $$
-where
-$$\Gamma_{star(k)}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i,j,k}^N g_{ij}g_{ik}$$
-Note that in every network $S_1(G)=\Gamma_{edges}(G)$, i.e., $S_1(G)$ is equal to the number of edges in the network.
-On this basis, Snijders introduces the \textit{alternating k-star statistics}
-\begin{equation*}
-\begin{split}
-\mathfrak{S}(G,\lambda):= & \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(G)\\= & ~   S_2(G)-\cfrac{S_3(G)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(G)}{\lambda^{N-3}}
-\end{split}
-\end{equation*} 
-Models with this statistic and a fixed decay parameter turn out to be standard ERGMs and Hunter and Handcock \cite{Hunter.2006} succeeded in proving that one can also rewrite alternating k-stars as a function of a network's degree distribution
-\begin{equation}
-\mathfrak{S}(G,\lambda)= \lambda\Biggr(\lambda\sum_{j=1}^{N-1}\Biggr( 1-\biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G) + 2S_1(G)\Biggl)
-\label{alto}
-\end{equation}
-where $D_j(G):= \Gamma_{deg(j)}(G)$ is the number of nodes with a degree of $j$. 
-In the next step, we define the \textit{geometrically weighted degree} (gwd) statistic as the first summand of (\ref{alto})
-\begin{equation}
-\Gamma_{gwd}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G)
-\end{equation}
-At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\
+%$$S_k(G):=\Gamma_{star(k)}(G) $$
+%where
+%$$\Gamma_{star(k)}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i,j,k}^N g_{ij}g_{ik}$$
+%Note that in every network $S_1(G)=\Gamma_{edges}(G)$, i.e., $S_1(G)$ is equal to the number of edges in the network.
+%On this basis, Snijders introduces the \textit{alternating k-star statistics}
+%\begin{equation*}
+%\begin{split}
+%\mathfrak{S}(G,\lambda):= & \sum_{k=2}^{N-1}(-\cfrac{1}{\lambda})^{k-2}S_k(G)\\= & ~   S_2(G)-\cfrac{S_3(G)}{\lambda}+ \dots + (-1)^{N-3}\cfrac{S_{N-1}(G)}{\lambda^{N-3}}
+%\end{split}
+%\end{equation*} 
+%Models with this statistic and a fixed decay parameter turn out to be standard ERGMs and Hunter and Handcock \cite{Hunter.2006} succeeded in proving that one can also rewrite alternating k-stars as a function of a network's degree distribution
+%\begin{equation}
+%\mathfrak{S}(G,\lambda)= \lambda\Biggr(\lambda\sum_{j=1}^{N-1}\Biggr( 1-\biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G) + 2S_1(G)\Biggl)
+%\label{alto}
+%\end{equation}
+%where $D_j(G):= \Gamma_{deg(j)}(G)$ is the number of nodes with a degree of $j$. 
+%In the next step, we define the \textit{geometrically weighted degree} (gwd) statistic as the first summand of (\ref{alto})
+%\begin{equation}
+%\Gamma_{gwd}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)D_j(G)
+%\end{equation}
+%At this point it also becomes obvious where the \textit{geometrically} comes from. It simply refers to the geometric sequence $(1-\frac{1}{\lambda})^j$ which appears in these statistics.\\
 \indent We estimate the ERGM using MCMLE and the bootstrapped MPLE.
@@
 \centering
-\includegraphics[width=3.5in, height=2.7in]{rel_time}
+\includegraphics[width=3.5in, height=2.5in]{rel_time}
 \caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
@@
 
-% that's all folks
 \end{document}

### 2017-10-05 TeX/complenet18.tex
Reduced CompleNet paper to 15 pages

@@
 First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.
-Faraj et al.
+% Faraj et al.
 (2008, pp.
 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques.
-This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.
+This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote}
+Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.
 Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference.
@@
 Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements.
-For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
+%For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
 For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable.
@@
 %
-where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
+where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}(N)$.
 %
-As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
+%As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). 
 The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al.
@@
 , and neuroscience \cite{simpson2011exponential}.%,simpson2012exponential}.
-
 \section{Estimation}
@@
 Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
-Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:\\[0.2cm]
+Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm \cite{chib1995understanding} to simulate networks:\\[0.2cm]
 
@@
 %\end{enumerate}
-\vspace{0.3cm}
-\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
+%\vspace{0.3cm}
+%\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
 Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
-the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
+%the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
 
@@
 %$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. 
+\begin{figure}[!t]
+\centering
+\includegraphics[width=3.5in]{RMSE_10k}
+\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
+\label{rmse}
+\end{figure}
 The gwesp statistic is used to model the tendency towards triangles and clustering in a network.\\
@@
 The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in]{RMSE_10k}
-\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
-\label{rmse}
-\end{figure}
 The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.

### 2017-10-05 TeX/llncs.tex
added in sideways captioning

--- 2017-08-03 TeX/conference_paper.tex
@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
+\documentclass{llncs}
 
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
+\usepackage{cite}
 
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
+\usepackage[pdftex]{graphicx}
 
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
-% Add the compsocconf option for Computer Society conferences.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-   \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
 \usepackage[cmex10]{amsmath}
 \usepackage{amssymb}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{dsfont}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
 
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
+\usepackage{array}
 
-% *** ALIGNMENT PACKAGES ***
-%
-\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
 %
 \usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
 
@@
 
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+%
+\begin{document}
 
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
-\begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
 \title{Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap}
 
-% author names and affiliations
-% use a multiple column layout for up to two different
-% affiliations
+\author{Christian S. Schmid and Bruce A. Desmarais}
 
-\author{\IEEEauthorblockN{Christian S. Schmid}
-\IEEEauthorblockA{Department of Statistics\\
-The Pennsylvania State University\\
-State College, USA\\
-Email: schmid@psu.edu}
-\and
-\IEEEauthorblockN{Bruce A. Desmarais}
-\IEEEauthorblockA{Department of Political Science\\
-The Pennsylvania State University\\
-State College, USA\\
-Email:  bdesmarais@psu.edu}
-}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
+\institute{The Pennsylvania State University\\
+State College, USA}
 \maketitle

### 2017-10-06 TeX/complenet18.tex
down to 12 pages

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{floatrow}
 
@@
 We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.
-
-\section{The Exponential Random Graph Model}
-The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
+\\
+\indent The ERGM is a probabilistic model for networks \cite{Wasserman.1996,robins.pattison.2007}.
 They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
@@
 As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\
-The probability function for the ERGM is defined as
+\indent The probability function for the ERGM is defined as \vspace{-.225cm}
 \begin{equation}
+\vspace{-.225cm}
 \mathbb{P}_{\theta}(G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
@@
 \end{equation}
-%
 where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}(N)$.
@@
 %As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). 
-The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al.
-\cite{SnijdersTomA.B..2006} for a detailed discussion).
-The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic}%,srivastava2011culture}
+The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated.
+The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including sociology \cite{smith2016ethnic}%,srivastava2011culture}
 , economics \cite{lomi2012networks}, political science \cite{Cranmer.2011}%,lazer2010coevolution}
@@
 \section{Estimation}
-As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.
-Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible.
 The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}.
 Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
-\begin{equation*}
-\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
-%\label{ILogit}
-\end{equation*}
-where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
+The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by
 \begin{equation*}
@@
 \end{equation*}
-where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}.
+where $G_{-ij}$ is the adjacency matrix, excluding element $ij$, $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}.
 For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.
@@
 Estimates of uncertainty are used to test hypotheses about parameters, estimate variance (i.e., risk) in model predictions, and estimate effect sizes.
-Another limitation of MPLE regards the assessment of model fit.
-With ERGM, model fit is conventionally evaluated by comparing the structure of the observed network to networks simulated from the estimated model \cite{Hunter.2008}.
-This comparison can include diagnosing the model for a particular form of poor fit that arises with ERGM---model degeneracy \cite{schweinberger2011instability}.
-Degeneracy is a condition in which the model places nearly all of the probability mass on the completely empty or completely full network.
-Since MPLE does not require simulating networks, researchers can derive and report results without checking the fit of the model or checking for degeneracy.
-Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.\\
-\indent The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.
+The current conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.
 This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}).
@@
 
-\noindent MCMC methods are used to simulate the $L$ networks.
-As we demonstrate below, the MCMLE grows more accurate as $L$ increases.
-Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
-Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm \cite{chib1995understanding} to simulate networks:\\[0.2cm]
-
-\begin{algorithm}[H]
- Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network)\\
- \For{$k \in \{0,...,L-1\}$ }{
-  Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$\\
-  Compute
-$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$\\
-Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$.\\
-  \eIf{$Z=0$}{
-   let $G^{(k+1)} := G^{(k)}$
-   }{
-   define $G^{(k+1)}$ via
-$g_{pq}^{(k+1)}=\begin{cases}
-1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
-g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
-\end{cases}$
-  }
- }
- \caption{Metropolis Hastings algorithm for Network Sampling}
-\end{algorithm}
-
-%Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
-%\begin{enumerate}
-%\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
-%\item Compute the value
-%$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
-%\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
-%\begin{itemize}
-%\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
-%\item $Z=1$, define $G^{(k+1)}$ via
-%$$g_{pq}^{(k+1)}=\begin{cases}
-%1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
-%g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
-%\end{cases}$$
-%\end{itemize}
-%\item Start at step 1 with $G^{(k+1)}$.
-%\end{enumerate}
-%\vspace{0.3cm}
-%\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
-Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
-%the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
-
+\noindent As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. 
 \section{Efficiency of MPLE and MCMLE}
@@
 This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time.
-At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework.
-They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.
-The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e.
-instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.
-
 To illustrate the relative efficiency of MPLE and MCMLE we run a simulation study.
@@
 The data for both networks originates from Resnick et al.
-\cite{Resnicketal1997}.\\
-\indent For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
+\cite{Resnicketal1997}.
+For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
 Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
+The gwesp statistic is used to model the tendency towards triangles and clustering in a network.
 %The number of edges is defined as
@@
 %$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in]{RMSE_10k}
-\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
-\label{rmse}
+
+\begin{figure}
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=2.5cm}}]{figure}[\FBwidth]
+{\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High} \label{rmse}}
+{\includegraphics[width=9.5cm]{RMSE_10k}} 
+\vspace{-.5cm}
 \end{figure}
-The gwesp statistic is used to model the tendency towards triangles and clustering in a network.\\
+
 \indent We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE.
-As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\
+As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.
 \indent Figure \ref{rmse} visualizes the results of the simulation study.
@@
 By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (G)$ is the sufficient statistic.
-For exponential family distributions, it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
+For exponential family distributions, a covariance matrix can be estimated by the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE.
 The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals.
@@
 The idea of using boostrap resampling with MPLE for ERGM was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals.
-Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE.
-\\
-\indent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap.
-The non-parametric bootstrap cannot be applied when there is just a single network under study.
-
-For the case in which there is just a single network to be studied, which is indeed the most common case in the literature, we propose the use of a parametric bootstrap.
+However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks).
+For the case in which there is just a single network to be studied, we propose the use of a parametric bootstrap.
 Under the parametric bootstrap, the sampling distribution of the MPLE is derived by re-estimating the MPLE on a sample of networks simulated from the MPLE estimated on the observed network.
 We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.
-
-For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values.
-We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $G$.
+\\
+\indent For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values.
+We use these parameter values to simulate a sample of $1000$ networks from the distribution of $G$.
 For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals.
@@
 For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain  95\% bootstrap confidence intervals.
-Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
-\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
-\label{coverage}
+We verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
+
+\begin{figure}
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=3cm}}]{figure}[\FBwidth]
+{\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression} \label{coverage}}
+{\includegraphics[width=8.5cm]{Coverage_Mesa_Magnolia_height05}} 
+\vspace{-.25cm}
 \end{figure}
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=1.75in]{bias_plots_05}
-\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right)}
-\label{bias}
+\begin{figure}
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=3cm}}]{figure}[\FBwidth]
+{\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right)} \label{bias}}
+{\includegraphics[width=8.5cm]{bias_plots_05}} 
+\vspace{-.25cm}
 \end{figure}
+
 Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
@@
 Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates.
-We can abstract from this figure that the median MPLE estimates concur with the networks 'true' coefficients.
-It is especially worthwhile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.\\
-\indent This simulation study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals.
-In this simulation study we demonstrate that the parametric bootstrap can be used in combination with MPLE to provide a method of estimation that is both computationally efficient and provides valid estimates of model uncertainty.
-
+The median MPLE estimates approximate the true parameters.
+It is especially worthwhile to mention that the bias of the larger Faux Magnolia High network is smaller than the bias of the Faux Mesa High network, supporting the fact that the MPLE converges to the MLE as the network size increases.
 \section{Cosponsorship Network Data}
@@
 To illustrate the performance of MCMLE relative to that of the bootstrapped MPLE we apply both approaches to the data on cosponsorship of bills in the U.S.
-House of Representatives for the 108th Congress (2003--2004), developed by Fowler (2006) \cite{Fowler2006a} \cite{Fowler2006b}.
+House of Representatives for the 108th Congress (2003--2004), developed by Fowler (2006) \cite{Fowler2006a}.
 The cosponsorship network consists of 2,635 nodes, which we define as pieces of legislation (i.e., bills), considered by the Senate during the 108th Congress.
-Note that this formulation of the cosponsorship network differs from past research, which has defined the nodes as the individual legislators.
-Because there are more bills than legislators, studying bills as nodes provides a more disaggregated look at the network than is offered through studying the network of legislators.
 In this undirected network bills are tied together based on the similarity of the sets of legislators who cosponsor them.
 Specifically, we include an edge between bills $i$ and $j$ if the correlation coefficient between the indicator vectors indicating whether $i$ and $j$ were sponsored each legislator is greater than a random uniform draw.
-This results in an undirected network with $28060$ edges.
-
-We build an ERGM specification that extends the work of  Zhang et al\cite{zhang2008community} in exploring the structure of cosponsorship ties.
+This results in an undirected network with $28060$ edges.\\
+\indent We build an ERGM specification that extends the work of  Zhang et al\cite{zhang2008community} in exploring the structure of cosponsorship ties.
 They find that congressional cosponsorship is primarily characterized by intra-party ties---among Republicans and among Democrats, but few cross-party ties.
@@
 This is done through a term that accounts for the party of the senators who sponsored the two bills in the pair.
-The party homophily term is defined as $$ \Gamma(G,X) = \sum_{i < j} g_{ij}x_{ij},$$ where X is an indicator matrix that assumes the value 0 if $i$ and $j$ were sponsored by legislators from different political parties and 1 if they were sponsored by legislators from the same party.
-$\Gamma(G,X)$ measures the number of intra-party ties in the network.
-A positive parameter value for this statistic indicates that ties tend to be formed between bills sponsored by the same political party.
-
-We extend the homophily-based model to account for a network dynamic that is commonly found in the study of networks---that of popularity or preferential attachment \cite{barabasi1999emergence}.
-Preferential attachment is the tendency for new ties to be formed with nodes who already have many ties (i.e., popular nodes).
+A positive parameter value for this statistic indicates that ties tend to be formed between bills sponsored by the same political party.\\
+\indent We extend the homophily-based model to account for a network dynamic that is commonly found in the study of networks---that of popularity or preferential attachment \cite{barabasi1999emergence}.
 The alternating k-star statistic was introduced by Snijders et al.
@@
 As we described in the section {\it Estimation}, only one edge at a time is changed when simulating networks.
-For better comparison, we chose the same Burn-In ($300,000$ MH-steps) and the same number of iterations ($30,000$ MH-steps) for sampling networks.
 The results can be found in table \ref{co_results}.\\
-\indent It is interesting to see that the standard error calculated by the logistic regression approach is much smaller than the standard error of the MCMLE for the alternating k-star statistic, which leads to inaccurate confidence intervals as shown in figure \ref{coverage}.
-The MPLE estimate is equivalent to the logistic regression estimate, but the bootstrap confidence intervals, especially for the alternating k-star statistic, are much wider than would be calculated using the logistic regression standard errors.
+\indent The MPLE estimate is equivalent to the logistic regression estimate, but the bootstrap confidence intervals, especially for the alternating k-star statistic, are much wider than would be calculated using the logistic regression standard errors.
 An estimate is generally considered statistically different from zero (i.e., statistically significant) if the confidence interval does not contain zero, or if the ratio of the estimate to the standard error exceeds 1.96 in magnitude.
@@
 \begin{split}
-\text{boostrapped MPLE time} = ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ \\ \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}
+\text{MPLE time} =  \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}
 \end{split}
@@
 Based on this, we define the relative computing time as
-$$\text{relative computing time}= \cfrac{\text{bootstrapped MPLE time}}{\text{MCMLE time}}$$
+$\frac{\text{MPLE time}}{\text{MCMLE time}}$.
 This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
 \\
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=2.5in]{rel_time}
-\caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
-\label{comptime}
+\begin{figure}
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=3.5cm}}]{figure}[\FBwidth]
+{\caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.} \label{comptime}}
+{\includegraphics[width=8cm]{rel_time}} 
+\vspace{-.5cm}
 \end{figure}
-Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
+\indent Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
 If exactly 500 computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at $0.20$ for the small and large network and $0.17$ for the medium network, meaning that the computing time can be quintupled using the bootstrapped MPLE.
 This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\
-Of course the actual computing time for a network always depends on the statistics that are included in the network, but in general larger networks require longer computation times, since a larger MCMC sample size is required and more MC steps are necessary to simulate a new network that does not overly depend on the previous sample.
-This makes the bootstrapped MPLE a beneficial alternative, especially for very large networks.
-\\
 \indent One of the major disadvantages of MPLE over MCMLE is that degeneracy is not assessed while the model is being estimated.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=2.7in]{MPLE_histogram_traceplot_cosponsorship}
-\caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.
+
+\begin{figure}
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=2.75cm}}]{figure}[\FBwidth]
+{\caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.
 The thick line in both, the traceplots and the histograms, represents the network statistics of the observed network.
-}
-\label{diagnostics}
+} \label{diagnostics}}
+{\includegraphics[width=9cm]{MPLE_histogram_traceplot_cosponsorship}} 
+\vspace{-.5cm}
 \end{figure}
@@
 \section{Conclusion}
-In the past years the ERGM grew in popularity in many different fields as a flexible and powerful means of building probabilistic models for networks.
-With this popularity the size of the considered networks has grown.
-As the size of the network increases, it becomes unclear how many simulated networks are necessary for the conventional MCMLE method to perform better than the MPLE.
-Furthermore, as an increasing number of simulated networks is necessary for the MCMLE, the computing time rapidly grows.
-For this reason it is essential to develop different methods that provide faster estimation than the MCMLE, but still lead to reliable results.\\
-In this paper we introduced the bootstrapped MPLE as an alternative method of statistical inference for ERGMs and compared the performance to the commonly applied MCMLE.
+
+\indent In this paper we introduced the bootstrapped MPLE as an alternative method of statistical inference for ERGMs and compared the performance to the commonly applied MCMLE.
 Based on a simulation study we demonstrated that the larger the size of a network is the larger the MCMC sample size has to be in order for the MCMLE to outperform the fast and simple MPLE.
@@
 \section{Acknowledgements}
-
 This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219.
-Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsors.
 

### 2017-10-16 TeX/IEEE_BigData.tex

--- 2017-08-03 TeX/conference_paper.tex
@@
 % to submit to, of course. )
+%\usepackage{floatrow}
+\usepackage{algorithm2e}
 
@@
 
-\begin{IEEEkeywords}
-network, ERGM, parametric bootstrap, maximum pseudolikelihood
-
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
-
 \section{Introduction}
@@
 First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.
-Faraj et al.
+% Faraj et al.
 (2008, pp.
 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques.
-This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.
+This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote}
+Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.
 Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference.
 We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.
-
-\section{The Exponential Random Graph Model}
-The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}.
+\\
+\indent The ERGM is a probabilistic model for networks \cite{Wasserman.1996,robins.pattison.2007}.
 They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
@@
 Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$.
-Furthermore, define
-$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
+Furthermore, define $\mathcal{G}(N)$
+%$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
 as the set of all possible networks on $N$ nodes without loops.
 Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements.
-For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
+%For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number.
 For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable.
 As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\
-The probability function for the ERGM is defined as
+\indent The probability function for the ERGM is defined as 
 \begin{equation}
@@
 \end{equation}
+where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}(N)$.
 %
-where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
-%
-As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}).
-The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al.
-\cite{SnijdersTomA.B..2006} for a detailed discussion).
-The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.
-
+%As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). 
+The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated.
+The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including sociology \cite{smith2016ethnic}%,srivastava2011culture}
+, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011}%,lazer2010coevolution}
+, ecology \cite{dey2014individual}%,bouskila2015similarity}
+, and neuroscience \cite{simpson2011exponential}.%,simpson2012exponential}.
 \section{Estimation}
-As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.
-Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible.
 The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}.
 Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.
-%
-\begin{equation*}
-\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
-%\label{ILogit}
-\end{equation*}
-where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
+The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by
 \begin{equation*}
@@
 Estimates of uncertainty are used to test hypotheses about parameters, estimate variance (i.e., risk) in model predictions, and estimate effect sizes.
-Another limitation of MPLE regards the assessment of model fit.
-With ERGM, model fit is conventionally evaluated by comparing the structure of the observed network to networks simulated from the estimated model \cite{Hunter.2008}.
-This comparison can include diagnosing the model for a particular form of poor fit that arises with ERGM---model degeneracy \cite{schweinberger2011instability}.
-Degeneracy is a condition in which the model places nearly all of the probability mass on the completely empty or completely full network.
-Since MPLE does not require simulating networks, researchers can derive and report results without checking the fit of the model or checking for degeneracy.
-Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.
-
-The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\
+The current conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.
 This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}).
@@
 \begin{eqnarray*}
-&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
-                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
-                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
-                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
+\small
+\text{loglik}(\theta)-\text{loglik}(\theta_0)\approx - 
+\log \left( \frac{1}{L}  \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T  \Gamma(G_i) \right) \right)
 \end{eqnarray*}
-By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
+
+%\begin{eqnarray*}
+%&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
+%                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
+%                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
+%                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
+%\end{eqnarray*}
+%By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
 %
-\begin{equation}
-s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
-\label{score}
-\end{equation}
+%\begin{equation}
+%s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
+%\label{score}
+%\end{equation}
 %
-This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
+%This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.
 
-MCMC methods are used to simulate the $L$ networks.
-As we demonstrate below, the MCMLE grows more accurate as $L$ increases.
-Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity.
-Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:
-Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
-\begin{enumerate}
-\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
-\item Compute the value
-$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
-\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
-\begin{itemize}
-\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
-\item $Z=1$, define $G^{(k+1)}$ via
-$$g_{pq}^{(k+1)}=\begin{cases}
-1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
-g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
-\end{cases}$$
-\end{itemize}
-\item Start at step 1 with $G^{(k+1)}$.
-\end{enumerate}
-\vspace{0.3cm}
-The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}.
-Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
-the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
-
+\noindent As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. 
 \section{Efficiency of MPLE and MCMLE}
@@
 The data for both networks originates from Resnick et al.
-\cite{Resnicketal1997}.\\
-\indent For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
+\cite{Resnicketal1997}.
+For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$.
 Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
-The number of edges is defined as
-$$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
-In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$.
+The gwesp statistic is used to model the tendency towards triangles and clustering in a network.
+%The number of edges is defined as
+%$$\Gamma_{edges}: \mathcal{G}(N) \to \mathbb{R}~~~, ~~~ G \to \sum_{i<j}^N g_{ij}$$
+%In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$.
 The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times.
@@
 Then, the statistic for a nodal covariate is defined as
-$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
-The GWESP statistic is given as
-$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
-where 
-$$
-\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
-$$
-$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. This statistic is used to model the tendency towards triangles and clustering in a network.\\
+%$$\Gamma_{nodal}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}g_{ij}c_{ij}$$
+%The GWESP statistic is given as
+%$$\Gamma_{gwesp}(G,\lambda):= \lambda\sum_{j=1}^{N-1}\Biggr(1- \biggr(1-\cfrac{1}{\lambda}\biggl)^j\Biggl)\Gamma_{esp(k)}(G)$$
+%where 
+%$$
+%\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
+%$$
+%$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors.
+
+\begin{figure}
+\centering
+\includegraphics[width=9cm]{RMSE_10k} 
+\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High} 
+\label{rmse}
+\vspace{-.5cm}
+\end{figure}
+
 \indent We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks.
@@
 In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE.
-As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\
+As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.
 \indent Figure \ref{rmse} visualizes the results of the simulation study.
 The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.6in]{RMSE_10k}
-\caption{The log of the ratio of the RMSE for the MCMLE to the MPLE for different sample sizes and two different networks, Faux Mesa High and Faux Magnolia High}
-\label{rmse}
-\end{figure}
 The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE.
@@
 For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain  95\% bootstrap confidence intervals.
-Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
+We verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.
+
+\begin{figure}
+\includegraphics[width=8.5cm]{Coverage_Mesa_Magnolia_height05}
 \caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
  \label{coverage}
+\vspace{-.25cm}
 \end{figure}
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=1.75in]{bias_plots_05}
+\begin{figure}
+\includegraphics[width=8.5cm]{bias_plots_05}
 \caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right)} 
 \label{bias}
+\vspace{-.25cm}
 \end{figure}
+
 Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks.
@@
 \begin{equation*}
+\small
 \begin{split}
-\text{boostrapped MPLE time} = ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ \\ \text{network simulation time} + \cfrac{500 \cdot \text{MPLE estimation time}}{x}
+\text{MPLE time} =  \text{network simulation time} + \cfrac{500 \cdot\text{MPLE estimation time}}{x}
 \end{split}
@@
 Based on this, we define the relative computing time as
-$$\text{relative computing time}= \cfrac{\text{bootstrapped MPLE time}}{\text{MCMLE time}}$$
+$\frac{\text{MPLE time}}{\text{MCMLE time}}$.
 This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results.
 \\
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=2.7in]{rel_time}
-\caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.}
-\label{comptime}
+\begin{figure}
+\includegraphics[width=8cm]{rel_time} 
+\caption{The y-axis gives the ratio of the bootstrapped MPLE time to that of the MCMLE time. Values below 1 indicate that the bootstrapped MPLE requires a shorter computing time.} \label{comptime}
+\vspace{-.5cm}
 \end{figure}
-Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
+\indent Figure \ref{comptime} demonstrates that all three networks only require three cores for the bootstrapped MPLE to outperform the computing time of the MCMLE and that the computing time can further be reduced if more computing cores are available.
 If exactly 500 computing cores are being used the ratio of the bootstrapped MPLE time to the MCMLE time levels off at $0.20$ for the small and large network and $0.17$ for the medium network, meaning that the computing time can be quintupled using the bootstrapped MPLE.
 This figure also depicts that larger network in general require a longer computation time and will benefit more if the bootstrapped MPLE is used.\\
-Of course the actual computing time for a network always depends on the statistics that are included in the network, but in general larger networks require longer computation times, since a larger MCMC sample size is required and more MC steps are necessary to simulate a new network that does not overly depend on the previous sample.
-This makes the bootstrapped MPLE a beneficial alternative, especially for very large networks.
-\\
 \indent One of the major disadvantages of MPLE over MCMLE is that degeneracy is not assessed while the model is being estimated.
@@
 This would be a signal that the constructed stochastic process violates the Markov properties.
-\begin{figure}[!t]
-\centering
-\includegraphics[width=3.5in, height=2.7in]{MPLE_histogram_traceplot_cosponsorship}
+
+\begin{figure}
+\includegraphics[width=9cm]{MPLE_histogram_traceplot_cosponsorship} 
 \caption{Network statistics of the 500 boootstrap samples for the cosponsorship network.
@@
 \label{diagnostics}
+\vspace{-.5cm}
 \end{figure}

### 2017-10-20 TeX/IEEE_BigData.tex
editied according to reviewers comments

@@
 Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as
-$RMSE = \sqrt{\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$
+$RMSE = \sqrt{\frac{1}{m}\sum_{i=1}^{m}(\theta - \hat{\theta}_i)^2}$
 implying that the smaller the RMSE, the more accurate is the estimator.
@@
 % \extrarowheight as needed to properly center the text within the cells
-\label{co_results}
 \centering
@@
 \caption{Estimation results for the Cosponsorship network using MCMLE, logistic regression and bootstrapped MPLE}
+\label{co_results}
 \end{table*}

### 2017-11-06 TeX/IEEE_BigData.tex
final edits

@@
 ERGM parameter estimation requires the approximation of an intractable normalizing constant.
-Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE).
+Simulation methods are used in the state-of-the-art approach to approximating the normalizing constant, which is the foundation of estimation by Monte Carlo maximum likelihood (MCMLE).
 MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant.
@@
 \indent The ERGM is a probabilistic model for networks \cite{Wasserman.1996,robins.pattison.2007}.
-They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
+They can be used for link prediction \cite{lu2010supervised, Desmarais:2011ieee}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}.
 The ERGM was first introduced by Holland and Leinhardt (1981) \cite{holland1981exponential}.
@@
 This results in an undirected network with $28060$ edges.\\
-\indent We build an ERGM specification that extends the work of  Zhang et al\cite{zhang2008community} in exploring the structure of cosponsorship ties.
-They find that congressional cosponsorship is primarily characterized by intra-party ties---among Republicans and among Democrats, but few cross-party ties.
+\indent We build an ERGM specification that extends past work \cite{zhang2008community, DesmaraisSoN} in exploring the structure of cosponsorship ties.
+Prior work has found that congressional cosponsorship is primarily characterized by intra-party ties---among Republicans and among Democrats, but few cross-party ties.
 We test for this party-based clustering (i.e., homophily) in our ERGM.
@@
 \begin{split}
-\text{MPLE time} =  \text{network simulation time} + \cfrac{500 \cdot\text{MPLE estimation time}}{x}
+\text{MPLE time} =  \text{simulation time} + \cfrac{500 \cdot\text{MPLE estimation time}}{x}
 \end{split}

### 2017-12-13 Presentation/beamer.tex
JSM slides added

\documentclass[xcolor=dvipsnames]{beamer}
%\usepackage[ngerman]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{xcolor}
\usepackage{dsfont}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{graphicx,tabularx,epsfig}
\usetheme{CambridgeUS} 
\usecolortheme{whale}
\useinnertheme{rounded}
\useoutertheme{infolines}
\setbeamercolor{frametitle}{fg=Blue,bg=Blue!20}
\setbeamercolor{block title}{use=structure,fg=white,bg=Blue}
\setbeamercolor{block body}{use=structure,fg=Black,bg=Blue!20}
\setbeamertemplate{footline}[frame number]
\beamertemplatenavigationsymbolsempty
\title{Exponential Random Graph Models with Big Networks:
Maximum Pseudolikelihood Estimation and the
Parametric Bootstrap }
\author{Christian Schmid\\
In collaboration with Bruce Desmarais}
\institute[The Pennsylvania State University]{The Pennsylvania State University}
\date{July 31st, 2017}

\begin{document} 
%
\frame{\titlepage

\scriptsize{This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219.}
} 
%

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item The Exponential Random Graph Model (ERGM) has an intractable normalizing constant.\\[0.4cm]
\item State-of-the-art approach: Approximization by Monte Carlo maximum likelihood (MCMLE).\\[0.4cm]
%\item MCMLE is accurate when a large sample of networks is used, but also becomes computationally expensive for large networks.\\[0.2cm]
\item For large networks the maximum pseudolikelihood (MPLE) is an option.\\[0.4cm]
\item We introduce a resampling method - the parametric bootstrap - that combines advantages of MCMLE and MPLE.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Exponential Random Graph Model}
\textbf{Idea:} Take the adjacency matrix of an observed network $A$ with $N$ nodes as a manifestation of a matrix-like random variable $Y$.\\[0.5cm]
%
\textbf{Definition:} 
$$\mathsf{P} (Y=A| \theta)=\dfrac{\exp(\theta^T \cdot \Gamma(A))}{c(\theta)}$$
%
where 
%
\begin{itemize}
\item $\theta \in \mathbb{R}^q$, is a vector of parameters
%\item $ \mathcal{A}(N) := \left\{ A \in \mathbb{R}^{(N\times N)}: a_{ij} \in \{0,1\},~a_{ii}=0\right\}$ 
\item $\Gamma:\mathcal{A}(N) \to \mathbb{R}^q~,~A \mapsto (\Gamma_1(A),\dots,\Gamma_q(A))^T$, is a vector of network statistics 
\item $c(\theta) :=\sum_{A^* \in \mathcal{A}(N)} \exp(\theta^T \cdot \Gamma(A^*))$, is a normalization constant
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Parameter Estimation Method 1: MCMLE}
\textbf{Idea:} Fix an auxiliary parameter vector $\theta_0 \in \mathbb{R}^q$. Then, we can show 
$$\frac{c(\theta)}{c(\theta_0)}= \mathbb{E}_{\theta_0}\bigl[exp((\theta-\theta_0)^T\Gamma(A))\bigr]$$
Approximate with MCMC sample of networks $A_1, \dots , A_L$.\\[0.4cm]
Then, by the law of big numbers, we get
$$\frac{c(\theta)}{c(\theta_0)}\approx \frac{1}{L}\sum_{i=1}^L\bigl[exp((\theta-\theta_0)^T\Gamma(A_i))\bigr]$$
%and therefore, 
%$$\ell(\theta)-\ell(\theta_0)\approx (\theta-\theta_0)^T\Gamma(A)-\log\biggl(\frac{1}{L}\sum_{i=1}^L\bigl[exp((\theta-\theta_0)^T\Gamma(A_i))\bigr]\biggr)$$
\end{frame}

\begin{frame}
\frametitle{Efficiency of MPLE/MCMLE}
\begin{itemize}
\item MCMLE is approximately exact.\\[0.8cm]
\item MPLE approaches the MLE as the size of the network increases.\\[0.8cm]
%$\Rightarrow$ For large networks MPLE is an alternative to the\\ ~~~~computationally expensive MCMLE.\\[0.4cm]
\item MCMLE requires a large sample size to perform well.\\[0.8cm]
\item The required sample size increases as network size increases.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Simulation Study}
\textbf{Data:}\\ 
We use two friendship networks, Faux Mesa High (203 students) and Faux Magnolia High (1451 students), and use the same parametrization (edges, sex and gwesp)\\[0.2cm]
\textbf{Treatment:} \\
\begin{itemize}
\item Take MCMLE of both networks as the 'true' coefficient $\theta$ and simulated 500 networks.\\[0.2cm]
\item Estimate the coefficients of these 500 networks using MPLE and MCMLE.\\[0.2cm]
\item For every single simulated
network the MCMLE calculation is being repeated several times for $25$ to
$10.000$ simulated networks used in the likelihood approximation.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Simulation Study Results}
This plot visualizes the log of the ratio of the root MSE of the MCMLE to the MPLE.
\begin{center}
\includegraphics[scale=.34]{RMSE_presentation} 
\end{center}
\end{frame}

\begin{frame}
\frametitle{Parameter Estimation Method 2: Bootstrapped MPLE}
In contrast to the MPLE, the MCMLE does not underestimate the standard error.\\[0.4cm]
We introduce the \textbf{bootstrapped MPLE} in order to obtain more reliable confidence intervals:\\
\setbeamertemplate{enumerate items}[default]
\begin{enumerate}
\item Estimate the MPLE\\[0.2cm]
\item Simulate a reasonable number of networks (e.g. 500)\\[0.2cm]
\item Estimate the MPLE for each simulated network\\[0.2cm]
\item Take the 2.5th and 97.5th percentile to obtain a 95\% bootstrap confidence interval\\[0.2cm]
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Bootstrapped MPLE vs. MCMLE: Coverage Probability}
\textbf{Treatment:} MCMLE as 'true' coefficients and simulate 1000 networks.\\[0.1cm] 
Calculate CI for each simulated network using bootstrapped MPLE, MCMLE and MPLE and report percentage the CI contains the 'true' value
\begin{center}
\includegraphics[scale=.4]{Coverage_presentation} 
\end{center}

\end{frame}

\begin{frame}
\frametitle{Bootstrapped MPLE vs. MCMLE: Computation Time}

$\bullet$ The bootstrapped MPLE is embarrassingly parallel.\\[0.2cm]
$\bullet$ By using multiple cores the computation time can be further reduced.\\[0.2cm]
This plot visualizes the time ratio of the bootstrapped MPLE to the MCMLE.

\begin{center}
\includegraphics[scale=.46]{relative_time_presentation} 
\end{center}

\end{frame}

\begin{frame}
\frametitle{Assessing Degeneracy}
Degeneracy occurs if the stochastic process generated by the MCMC- algorithm does not hold through the model’s defined distribution as stationary distribution.
\begin{center}
\includegraphics[scale=.4]{degeneracy_presentation} 
\end{center}
\end{frame}

\begin{frame}
\frametitle{Conclusion}
%MCMLE outperformed the two Bayesian methods in the simulation study. \\[0.2cm]
\begin{table}[]
\centering
\begin{tabular}{cll}
\hline
                      & \multicolumn{1}{c}{\textbf{Pros}}                                                                        & \multicolumn{1}{c}{\textbf{Cons}}                                                                         \\  \hline 
\multirow{2}{*}{\textbf{MCMLE}} & $\bullet$ approximately exact & $\bullet$ Computationally expensive            \\
                      &                                                                                                   $\bullet$ can assess degeneracy & \begin{tabular}[c]{@{}l@{}}$\bullet$ $\theta_0$ has to be picked  \\ close to $\theta$\medskip \end{tabular} \\ \hline
\multirow{2}{*}{\textbf{MPLE}} & $\bullet$ simple and fast & \multicolumn{1}{l}{$\bullet$ underestimates st. error}                                       \\ 
                      & \begin{tabular}[c]{@{}l@{}}$\bullet$ consistent estimator\medskip \end{tabular} &\\ \hline                                                                                               
\multirow{2}{*}{\textbf{b. MPLE}} & $\bullet$ simple and fast & \multicolumn{1}{l}{$\bullet$ slower than MPLE}                                       \\
                      & $\bullet$ can assess degeneracy & \\
                      & $\bullet$ consistent estimator& \\
                      & $\bullet$ reasonable CIs &
\end{tabular}
\end{table}
\end{frame}

\begin{frame}%[allowframebreaks]
  \frametitle{References}    
  \begin{thebibliography}{10}    
  \beamertemplatearticlebibitems
  \bibitem{Cranmer2011}
    B.A. Desmarais and S.J. Cranmer
    \newblock {\em Statistical mechanics of networks: Estimation and uncertainty}
    \newblock {\em Physica A}, 391(4):1865--1876, 2011.

  \beamertemplatearticlebibitems
  \bibitem{vanDuijn2009}
    M. van Duijn, K. Gile and M. Handcock
    \newblock {\em A framework for the comparison of maximum pseudo-likelihood and maximum likelihood estimation of exponential random graph models}
    \newblock {\em Social Networks}, 31(1):52--62, 2009.

\beamertemplatearticlebibitems
  \bibitem{Robins2007}
    G. Robins, P. Pattison, Y. Kalish and D. Lusher
    \newblock {\em An introduction to exponential random graph (p*) models for social networks}
    \newblock {\em Social Networks}, 29(2):173--191, 2007.

    \beamertemplatearticlebibitems
  \bibitem{StraussIkeda}
    D. Strauss and M. Ikeda
    \newblock {\em Pseudolikelihood estimation for social networks}
    \newblock {\em JASA}, 85:204--212, 1990.
  \end{thebibliography}
\end{frame}

\begin{frame}
\frametitle{Simulating Networks with MCMC}
Choose a matrix $A^{(0)} \in \mathcal{A}(N)$ to start with and proceed as follows:\\
\begin{enumerate}
\item Randomly choose a dyad $A_{ij}:=A[i,j]$ where $i \neq j$ from $A^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathsf{P}(Y_{ij} \neq A_{ij}^{(k)} | Y_{ij}^c=A_{ij}^c, \theta_0)}{\mathsf{P}(Y_{ij}= A_{ij}^{(k)} | Y_{ij}^c=A_{ij}^c, \theta_0)}$$%= exp(\theta^T(\Gamma(A_{ij}^+)-\Gamma(A_{ij}^-)))$$
where $Y_{ij}^c=A_{ij}^c$ is short for $Y_{pq}=A_{pq}$ for all $(p,q)\neq (i,j)$.\\[0.2cm]
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If \\[0.2cm]
\begin{itemize}
\item $Z=0$, let $A^{(k+1)} := A^{(k)}$ \\[0.2cm]
\item $Z=1$, set $A_{ij}^{(k+1)}=1- A_{ij}^{(k)}$
\end{itemize}
\item Start at step 1 with $A^{(k+1)}$.
\end{enumerate}

\end{frame}

\end{document}
